Peer Assessment 1
=============
# section 1: What is mean total number of steps taken per day?
```{r echo=TRUE ,cache=TRUE}
data<-read.table("d://download/repdata-data-activity/activity.csv",sep = ",",header = T,na.strings = "NA")
data$date<-strptime(data$date,"%Y-%m-%d")
summary(data)
str(data)
```


##the total number steps per day are as follows:

```{r echo=TRUE, cache=TRUE}
library(sqldf)

data$date<-as.Date(data$date)
bad<-is.na(data$steps)
data1<-data[!bad,]
data1$date<-as.Date(data1$date)
real<-sqldf("select date,sum(steps) from data1 group by date")
real
hist(real$'sum(steps)',main = "total number of steps taken each day",xlab = "total steps",ylab = "frequency",col = "red")
```

##the mean is:
```{r echo=TRUE,cache=TRUE}
mean(real$'sum(steps)')
```
##the median is:
```{r echo=TRUE,cache=TRUE}
median(real$'sum(steps)')

```

#section 2: What is the average daily activity pattern?
```{r echo=TRUE,cache=TRUE}
library(ggplot2)
summary(data1)
tb2<-sqldf("select interval,sum(steps)/count(distinct(date)) as stps from data group by interval")
g<-ggplot(tb2,aes(interval,stps))
g1<-g+geom_line()
g1
```

##the maximum of the average steps of the time interval is:
```{r echo=TRUE,cache=TRUE}
tb3<-tb2[order(tb2$stps,decreasing = T),]
tb3[1,1]
```

#section3: Imputing missing values

##total number of missing values is:
```{r echo=TRUE,cache=TRUE}

length(which(is.na(data$steps)==T))
```

##fill in the missing values and create a new dataset called "new":
```{r echo=TRUE,cache=TRUE}

val<-sqldf("select data.*,tb2.stps from data, tb2 where data.interval=tb2.interval")
  for(i in 1:length(val$steps)){
  if(is.na(val[i,1]))
    
    val[i,1]<-val[i,4]
    
  }
new<-val[,1:3]
```

##make a histogram of the total number of steps taken each day:
```{r echo=TRUE,cache=TRUE}
new1<-sqldf("select date,sum(steps) from new group by date")
hist(new1$'sum(steps)',main = "total number of steps taken each day",xlab = "total steps",ylab = "frequency",col = "red")
```



##compare new data and orginal data:
```{r echo=TRUE,cache=TRUE}
newdata<-c(mean(new1$'sum(steps)'),median(new1$'sum(steps)'))
originaldata<-c(mean(real$'sum(steps)'),median(real$'sum(steps)'))
data.frame(new=c(mean(new1$'sum(steps)'),median(new1$'sum(steps)')),original=c(mean(real$'sum(steps)'),median(real$'sum(steps)')),row.names = c("mean","median"))
```

#section 4: Are there differences in activity patterns between weekdays and weekends?

```{r echo=TRUE,cache=TRUE}
new$week<-weekdays(new$date)

for(i in 1:length(new$date)){
  if(new[i,4]=="ÐÇÆÚÁù"||new$week=="ÐÇÆÚÈÕ"){
    new[i,4]="weekend"}
  else {
    new[i,4]="weekday"}
}
tb4<-sqldf("select interval,week,sum(steps)/count(distinct(date)) as stps from new group by interval,week")
```
```{r echo=TRUE,cache=TRUE}
ggplot(tb4,aes(interval,stps))+geom_line(col="blue")+facet_grid(week ~ .)
```

