---
title: "Activity Monitoring"
author: "George Lwevoola"
date: "May 13, 2016"
output: html_document
---

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals throughout the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

Loading and preprocessing the data

```{r}
## Loading and preprocessing the data
## check whether the file has been downloaded previously
## if not proceed to download data file
if(!file.exists("activity.zip")){
  fileUrl <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileUrl,destfile="activity.zip")
  dateDownloaded <-date()
}

## Data file exists. Set up connection to read file
con1 <- unz(description="activity.zip", filename="activity.csv")
activity_dt<- read.csv(con1,sep=",",header=TRUE)
```
What is the mean total number of steps taken per day?

```{r}
## Summarize dataset to calculate the mean steps per day
## Ignoring missing data
library(dplyr)
mydt <- group_by(activity_dt,date)
mydtmeans <-summarize(mydt,meansteps=mean(steps))
print(mydtmeans)
```
Make a histogram of the total number of steps taken each day

```{r}
## Calculate the total number of steps taken per day
## Make a histogram of the total number of steps taken each day
mydtsums <-summarize(mydt,totalsteps=sum(steps))
Total_Steps<-mydtsums$totalsteps
hist(Total_Steps)
```
Calculate and report the mean and median of the total number of steps taken per day
```{r}
##Calculate and report the mean and median of the total number of steps taken per day
mydtmm <-summarize(mydt, mean=mean(steps),median=median(steps))
print(mydtmm)
```
What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
mydt2 <- group_by(activity_dt,interval)
myplot<-summarize(na.omit(mydt2),steps=mean(as.numeric(steps)))
plot(myplot$interval,myplot$steps,type="l",xlab="intervals",ylab="average steps")

```
5-minute interval, on average across all the days in the dataset, with the maximum number of steps?
```{r}
subset(myplot, steps == max(steps))

```
Imputing missing values
Calculate and report the total number of missing values 
in the dataset (i.e. the total number of rows with NAs)
```{r}
print(missing <-sum(is.na(activity_dt$steps)))
```
The strategy devised for filling out the missing values is the replacement of NA's with the mean
```{r}
mydata_NA<-arrange(activity_dt,interval)
library(plyr)
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
mydata_NA2 <- ddply(mydata_NA, ~ interval, transform, steps = impute.mean(steps))
mynew_dt <-mydata_NA2[order(mydata_NA2$date), ]       
## detach plyr pacakage and reattach dplyr
detach(package:plyr) 
library(dplyr)
```
New dataset that is equal to the original dataset but with the missing data filled in has been created.

Histogram of the total number of steps taken each day and the mean and median total number of steps taken per day
```{r}
## Make a histogram of the total number of steps taken each day and 
myhdata <-group_by(mynew_dt,date)
myhdata2 <-summarize(myhdata,steps=sum(steps))
hist(myhdata2$steps)
```
The mean and median from the new data set

```{r}
print(myhdata3<-summarize(myhdata,mean=mean(steps),median=median(steps)))

```

Determine differences in activity patterns between weekdays and weekends.
Calculate and report the mean and median total number of steps taken per day
```{r}
## Create a new factor variable in the dataset with two levels - 
## "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
mynew_dt$day <-weekdays(as.Date(mynew_dt$date))
mynew_dt$dday= as.factor(ifelse(as.character(mynew_dt$day) == "Saturday" | as.character(mynew_dt$day) == "Sunday", "weekend","weekday"))

## Make a panel plot containing a time series plot (i.e. type = "l") of the 
## 5-minute interval (x-axis) and the average number of steps taken, 
## averaged across all weekday days or weekend days (y-axis)
dt <- group_by(mynew_dt,interval,dday)
dt2<- summarize(dt,steps=mean(steps))

library(lattice)
xyplot(steps~interval|dday,type="l", data=dt2)

```

