---
title: "FitBit Data Analysis"
author: "Elizabeth Betts"
date: "January 15, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is Assignment 1 for Reproducible Research  in Coursera's Data Science specialization track. The purpose of the project was to create a R Markdown document to describe some data collected by FitBit.

##Purpose of project  

The purpose of this project was to practice

1. loading and preprocessing data  
2. imputing missing values  
3. interpreting data to answer research questions  

##Data for the assigment was downloaded from the course website  

Link to required Dataset: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

The variables included are:

1. steps: Number of steps taking in a 5-minute interval (missing values are coded as \color{red}{\verb|NA|}NA)
2. date: The date on which the measurement was taken in YYYY-MM-DD format
3. interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

##Loading and preprocessing the data

```{r}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
        temp <- tempfile()
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
        unzip(temp)
        unlink(temp)
}

data <- read.csv("activity.csv")
```

##What is the average (mean) total number of steps taken per day?
```{r}
steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Daily Steps"), col="red", xlab="Number of Steps")
rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)



```
The mean is <rmean> and the median is <rmedian>.

##What is the average daily activity pattern?

Calculate average steps for each interval for all days.
Plot the Average Number Steps per Day by Interval.
Which interval has most average steps?

```{r}
steps_interval <- aggregate(steps ~ interval, data, mean)

plot(steps_interval$interval, steps_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",
     main = "Average daily steps by time interval")

max_interval <- steps_interval[which.max(steps_interval$steps),1]

```

The highest number of steps during a single 5 minute interval is <max_interval>.

##Impute missing values and compare the results from imputed to non imputed data
```{r}
incomplete <- sum(!complete.cases(data))
imputed_data <- transform(data, steps = ifelse(is.na(data$steps), steps_interval$steps[match(data$interval, steps_interval$interval)], data$steps))

steps_by_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")

#Create Histogram to show difference. 
hist(steps_by_day$steps, main = paste("Total Daily Steps"), col="red", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)

```
Calculate new mean and median for the imputed data.
```{r}
rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)
```
Calculate difference between imputed and non imputed data
```{r}
mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian

```
Calculate the total difference
```{r}
total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)
```

The imputed mean is <rmean.i>
The imputed data median is <rmedian.i>
The difference between the non imputed mean and the imputed mean is <mean_diff>
The difference between the non imputed median and the imputed median is <med_diff>

##Are there any differences between weekday and weekend step counts?
Step volumes during weekdays vs. weekend days were analysed to observe any differences.

```{r}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
imputed_data$wkday = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))

steps_by_interval_i <- aggregate(steps ~ interval + wkday, imputed_data, mean)

library(lattice)

xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$wkday, main="Average Daily Steps by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
```


