---
title: "Activity Monitoring Data Analysis"
author: "Viji Vennelakanti"
date: "December 16, 2015"
output: html_document
---

This is the R Markdown File to submit an assignment for the Reproducible Research Course from Coursera.

Load necessary packages
```{r load_packages, echo=FALSE, warning=FALSE}
library(plyr)
library(ggplot2)
library(gridExtra)
library(Rmisc)
```


## Loading and preprocessing the data
### Read Data and see how it looks
Also made a copy of the dataframe for later use
```{r Read_data}
actdf <- read.csv("activity.csv",header = TRUE,sep = ",")
actdfcopy <- actdf
head(actdf)
summary(actdf)
str(actdf)
```


## What is mean total number of steps taken per day? 
First get Mean and Median per day  
```{r Mean_perDay}
totperday <- ddply(actdf, .(date), summarize, 
                    dailymean = round(mean(steps, na.rm=TRUE),2),
                    dailymedian = median(steps),
                    dailytotal = sum(steps, na.rm=TRUE)
                    )
head(totperday)        
```
###Date Transformations
```{r DateTransforms}
totperday$mydate <- strptime(totperday$date, "%Y-%m-%d")
totperday$mymonth <- months(totperday$mydate, FALSE)
head(totperday)
str(totperday)
```

Total Mean & Median Calculation
```{r MeanMedian}
totmean = round(mean(totperday$dailytotal, na.rm=TRUE),0)
totmedian = round(median(totperday$dailytotal, na.rm = TRUE),0)
totmean
totmedian
```

### Plot Steps Taken Per Day  
Also draw a Horizontal Line for Mean and Median  
```{r plot_totals}
g1 <- ggplot(data = totperday, aes(x=date, y=dailytotal, fill = mymonth))+
        geom_bar(stat="identity") +
        geom_hline(aes(yintercept = totmean), size=1, color="blue", 
                   linetype = "dashed") +
        geom_hline(aes(yintercept = totmedian), size=1, color = "black",
                   linetype = "dashed") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        labs(x = "Date", y="Total Number of Steps") +
        ggtitle("Steps taken per day") +
        scale_fill_discrete(name="Months",
                      breaks= totperday$mymonth,
                      labels= totperday$mymonth) +
         annotate("text", x = 4, y = (totmean-500), label = "Mean") + 
         annotate("text", x = 4, y = (totmedian+500), label = "Median")  
g1  
```    

## What is the average daily activity pattern?  

###Calculate interval averages and time at max. steps  
```{r time_series}
timeintavg <- aggregate(steps~interval, actdf, mean, na.action = na.omit)
timeintavg$steps <- round(timeintavg$steps,2)
maxsteps = max(timeintavg$steps, na.rm = TRUE)
maxsteps
timeatmaxsteps <- timeintavg[timeintavg$steps==maxsteps,]
```    


###Now Plot this  
```{r plot_ts}
ggplot(data = timeintavg, aes(x=interval, y=steps)) +
        geom_line() +
        ggtitle("Time Series Plot") +
        labs(x = "Interval", y = "Average Steps taken") +
        geom_vline(aes(xintercept = timeatmaxsteps$interval), size=1, 
                   color = "red", linetype = "dashed") +
         annotate("text", x = (timeatmaxsteps$interval-200), 
                  y = (timeatmaxsteps$steps), label = "Max Steps")
         
```  

## Imputing missing values  

We can count missing NAs.    
Then we can replace NA values in steps with corresponding time interval average
for steps and if that is also NA, by the daily mean and then daily median. There may be cases where all the values are NA.   

```{r missing_data}
summary(actdf)
sum(is.na(actdf["steps"]))

for (i in 1:nrow(actdfcopy))
{
        if (is.na(actdfcopy[i,"steps"]) == TRUE)
        {actdfcopy[i,"steps"] <- timeintavg[i,"steps"]}
        if (is.na(actdfcopy[i,"steps"]) == TRUE)
        {actdfcopy[i,"steps"] <- totperday[i,"dailymean"]}
        if (is.na(actdfcopy[i,"steps"]) == TRUE)
        {actdfcopy[i,"steps"] <- totperday[i,"dailymedian"]}
        
}
```
Total number of missing values in the original dataset is 
`r sum(is.na(actdf["steps"]))`  
Total number of missing values after replacing with timeinterval means is `r sum(is.na(actdfcopy["steps"]))`  

###Calculcate totals for this new dataframe   
```{r total_copy}
totperdaycopy <- ddply(actdfcopy, .(date), summarize, 
                    dailymean = round(mean(steps, na.rm=TRUE),2),
                    dailymedian = median(steps),
                    dailytotal = sum(steps, na.rm=TRUE)
                    )
head(totperdaycopy)
```
###Date transformations for this new dataframe
```{r date_transform_copy}
totperdaycopy$mydate <- strptime(totperdaycopy$date, "%Y-%m-%d")
totperdaycopy$mymonth <- months(totperdaycopy$mydate, FALSE)
head(totperdaycopy)
str(totperdaycopy)
```
###Calculate the Total Mean and Medians for the new Dataframe
```{r mean_median_copy}
totmeancopy = round(mean(totperdaycopy$dailytotal, na.rm=TRUE),0)
totmediancopy = round(median(totperdaycopy$dailytotal, na.rm = TRUE),0)
totmeancopy
totmediancopy
```

###Finally plot this info and compare with previous similar plot
```{r plot_copy}
g2 <- ggplot(data = totperdaycopy, aes(x=date, y=dailytotal, fill = mymonth))+
        geom_bar(stat="identity") +
        geom_hline(aes(yintercept = totmeancopy), size=1, color="blue", 
                   linetype = "dashed") +
        geom_hline(aes(yintercept = totmediancopy), size=1, color = "black",
                   linetype = "dashed") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        labs(x = "Date", y="Total Number of Steps") +
        ggtitle("Steps taken per day  with NAs replaced with Median") +
        scale_fill_discrete(name="Months",
                      breaks= totperdaycopy$mymonth,
                      labels= totperdaycopy$mymonth) +
         annotate("text", x = 4, y = (totmeancopy-500), label = "Mean") + 
         annotate("text", x = 4, y = (totmediancopy+500), label = "Median") 

g2

grid.arrange(g1, g2, ncol=1, nrow =2) 
```

> We can see from the plots that there is no perceivable difference of replacing
NA values with means for time value intervals. The total mean and median values also change very little:  
1. From `r totmean` to `r totmeancopy` for Mean   
2. From `r totmedian` to `r totmediancopy` for Median  

## Are there differences in activity patterns between weekdays and weekends?      
###Create a new column of type factor for weekday and weekend

```{r weekday_weekend}
actdfcopy$mydate <- strptime(actdfcopy$date, "%Y-%m-%d")
actdfcopy$Day <- weekdays(actdfcopy$mydate,abbreviate = TRUE)
actdfcopy$DayType = "X"
        
for (i in 1:nrow(actdfcopy))
{
        if ((actdfcopy[i, "Day"] == 'Sat')  | 
                (actdfcopy[i, "Day"] == 'Sun'))
                {actdfcopy[i, "DayType"] = "Weekend"} 
        else {actdfcopy[i, "DayType"] = "Weekday"}      
}
str(actdfcopy)
actdfcopy$DayType <- as.factor(actdfcopy$DayType)
str(actdfcopy)
```
###Averaging per timeinterval seperately for weekend and weekday
```{r time_series_weekday_weekend}
timeintavgw <- aggregate(steps ~ DayType + interval, actdfcopy, mean, 
                          na.action = na.omit)

timeintavgw$steps <- round(timeintavgw$steps,0)

```

###Now Plot this for weekday and weekend data separately and then together in a grid
```{r plot_ts_w}
gwd <- ggplot(data = subset(timeintavgw, DayType == "Weekday"), aes(x=interval, 
                                                             y=steps)) +
        geom_line() +
        ggtitle("weekday")

gwd

gwe <- ggplot(data = subset(timeintavgw, DayType == "Weekend"), aes(x=interval, 
                                                             y=steps)) +
        geom_line() +
        ggtitle("weekend")

gwe

grid.arrange(gwd,gwe, ncol=1, nrow=2)

```


        


