---
title: "Course_Assignment_1"
author: "Felix Barangan"
date: "April 24, 2016"
output: html_document

---
                                         
## Introduction

It is now possible to collect a large amount of data about personal
movement using activity monitoring devices such as a
[Fitbit](http://www.fitbit.com), [Nike
Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or
[Jawbone Up](https://jawbone.com/up). These type of devices are part of
the "quantified self" movement -- a group of enthusiasts who take
measurements about themselves regularly to improve their health, to
find patterns in their behavior, or because they are tech geeks. But
these data remain under-utilized both because the raw data are hard to
obtain and there is a lack of statistical methods and software for
processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring
device. This device collects data at 5 minute intervals through out the
day. The data consists of two months of data from an anonymous
individual collected during the months of October and November, 2012
and include the number of steps taken in 5 minute intervals each day.

## Data

The data for this assignment can be downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken




The dataset is stored in a comma-separated-value (CSV) file and there
are a total of 17,568 observations in this
dataset.

### Loading and preprocessing the data

The following are libraries and packages used:
```{r, echo = FALSE}
library(dplyr)
library(ggplot2)
library(mice)
library(knitr)
```             

default global options:          
```{r setopions, echo=FALSE}
opts_chunk$set(eval=TRUE)
```

Zip file downloaded from the Rdpeng Repo and eventually unzip using OS specific program. Check and set specific directory using:
```{r, echo=FALSE}
getwd()
```

In my case, I named the .csv file as "activity.csv"
Load fil:
```{r load_file, echo=FALSE}
#load csv file
activity <- read.csv( "activity.csv")
```
       
This may not be necessary by I do like to have a general rule to check the data file
```{r checkData, echo=FALSE}
str(activity)
names(activity)
summary(activity)
```

Based on the above summary, It trned out that there are   NA's   :2304   on one of the variable (step).

Run a code to determine if there are any other NA value that will affect analysis. Coerce Date variable from character to date to extract days of the week value

```{r sumNull, echo=TRUE}
stepsNull <- sum(is.na(activity$steps))
dateNull <- sum(is.na(activity$date))

activity$Date <- as.POSIXct(activity$date)
```                

Converted the file to tbl.df using dplyr package for ease of table manipulation
Please check [Hadley Wickham's tutorial on dpylr][1]
        
```{r, echo=FALSE}
#convert csv to dplyr (tbl.df) removing NA's
activity <- tbl_df(activity)
activity_No_NA <- activity %>%
        filter(!is.na(steps)) %>%
        select(steps, Date, interval)
```
```{r actvitytbl, echo=TRUE}
activity
```
## Assingment rubrics
###What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day
2. Calculate and report the mean and median of the total number of steps taken per day
3. Make a histogram of the total number of steps taken each day
```{r question1,echo=FALSE}
stepsPerDay <- activity_No_NA %>%
                group_by(Date) %>%
                summarise(Steps = sum(steps)) %>%
                select(Date, Steps)
```

```{r}
statsPerDay <- activity_No_NA %>%
        group_by(Date) %>%
        summarise(Steps = sum(steps), meanSteps = mean(steps)) %>%
        select(Date, Steps, meanSteps)
```

Histogram
```{r, echo=TRUE}
hist(stepsPerDay$Steps , breaks = 10,
        xlab = "Number of steps per day",
        freq = TRUE,
        main = "Bell Shape distribution (Steps per Day)",
        axes = TRUE,
        col = "red",
        labels = TRUE,
        type = "count")
```




        
[1]: https://www.youtube.com/watch?v=8SGif63VW6E
