---
title: "Activity monitoring data analysis"
author: "Kidpea LAU"
date: "25/09/2018"
output:
  html_document: default
  pdf_document: default
---
#About the data:
The data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

#load the data

```{r simulatedata,echo=TRUE}
fileUrl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile="./data1/Dataset.zip")
unzip(zipfile="./data1/Dataset.zip",exdir="./data1")
path_rf <- file.path("./data1" , "data1.csv")
files<-list.files(path_rf, recursive=TRUE)
act<-read.csv("./data1/activity.csv")
```

#Histogram of the total number of steps taken each day
```{r echo=TRUE}
stepperday<-tapply(act$steps,act$date,sum,na.rm=T)
hist(stepperday, xlab = "number of steps",
     main = "the total number of steps taken each day")
```

#Mean and median number of steps taken each day
```{r echo=TRUE}
summary(stepperday)
```

#Time series plot of the average number of steps taken
```{r echo=TRUE}
mean_interval <- tapply(act$steps, act$interval, mean, na.rm = T)

plot(mean_interval, type = "l", main = "time series plot", xlab = "the 5-minute interval", ylab = "the average number of steps")
```

#The 5-minute interval that, on average, contains the maximum number of steps
```{r echo=TRUE}
which.max(mean_interval)
```

#Code to describe and show a strategy for imputing missing data
```{r echo=TRUE}
stepsNA <- sum(is.na(act$steps))
dateNA <- sum(is.na(act$date))
intervalNA <- sum(is.na(act$interval))
```

#Histogram of the total number of steps taken each day after missing values are imputed
```{r echo=TRUE}
numMissingValues <- length(which(is.na(act$steps)))
numMissingValues
```
###set a function to impute the NA
```{r echo=TRUE}
impute <- function(x, x.impute){ifelse(is.na(x),x.impute,x)}
act2 <- act
act2$steps <- impute(act$steps, mean(act$steps))
```
###Make a histogram of the total number of steps taken each day
```{r echo=TRUE}
act2step <- tapply(act2$steps, act2$date, sum,na.rm=TRUE)
hist(act2step, xlab='Total steps per day (Imputed)')
```

#Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
##make a new vector to describe weekday and weekend
```{r echo=TRUE}
act2$dateType <-  ifelse(as.POSIXlt(act2$date)$wday %in% c(0,6), 'weekend', 'weekday')
```
##make the plot
```{r scatterplot, echo=TRUE, fig.height=4, fig.show=T}
library(ggplot2)
avgAct2 <- aggregate(steps ~ interval + dateType, data=act2, mean)
ggplot(avgAct2, aes(interval, steps)) + 
  geom_line() + 
  facet_grid(dateType ~ .) +
  xlab("5-minute interval") + 
  ylab("avarage number of steps")
```
