---
title: "Reproducable Research"
date: "2/8/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Download Data and Make it Available in R

The first step is to acquire teh data and put it into a usable format.

```{r load data}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", temp)
data <- read.csv(unz(temp, "activity.csv"))
unlink(temp)
```

## Transform Data

You can also embed plots, for example:

```{r transform}
library(dplyr)
daily_steps <- data %>% group_by(date) %>% summarise(daily_steps = sum(steps, na.rm=TRUE))
```

## Create Histogram & Report Median




```{r histogram}
library(ggplot2)
ggplot(daily_steps, aes(daily_steps)) + 
  geom_histogram(binwidth=2500, fill ='blue', color='black') + 
  ggtitle("Distribution of Daily steps")

median(daily_steps$daily_steps)
mean(daily_steps$daily_steps)
```

## Create Time Series & Identify Period with Highest Number of Steps

```{r time series}
time_series <- data %>% group_by(interval) %>% summarise(median_steps = median(steps, na.rm=TRUE))
with(time_series, plot(x=interval, y=median_steps, type='l'), title='Number of Steps Over Time')

max(time_series$median_steps)
```

## Impute Missing Values

Here are the total missing data points
```{r impute missing values}
sum(is.na(data$steps))
sum(data$steps, na.rm=TRUE)
data2 <- data
for(i in 1:nrow(data2)){
  if(is.na(data2$steps[i])==TRUE){ data2$steps[i] <-time_series$median_steps[time_series$interval==data2$interval[i]]}
  
}
distribution <- data2 %>% 
  group_by(date) %>% 
  summarise(daily_step = sum(steps))

ggplot(distribution, aes(daily_step)) + geom_histogram(binwidth=2500, color='black', fill='blue') 

median(distribution$daily_step)
mean(distribution$daily_step)
```

## Weekdays versus Weekends

```{r Day of Week}
library(lubridate)
data2$day_of_week <- ifelse(weekdays(ymd(data2$date))=='Saturday' | weekdays(ymd(data2$date))=='Sunday', 'Weekend', 'Weekday')
time_series2 <- data2 %>%
  group_by(interval, day_of_week) %>%
  summarise(daily_steps = mean(steps))
library(lattice)
xyplot(daily_steps ~ interval|day_of_week, 
       data=time_series2, 
       type='l', 
       layout=c(1,2),
       main = 'Average Steps Taken Over Time')
