Assignment
==========

#Set the directory and load the csv file
```{r,echo=TRUE}
setwd("D:/Download/IDM Download/Compressed/New folder (5)")
Data<-read.csv("activity.csv", header = TRUE)
```
#Total steps taken perday and histogram of the data

```{r,echo=TRUE}
c<-aggregate(steps~date,Data,sum)
hist(c$steps,
     main = "Total Steps per Day",
     xlab = "Number of Steps")
```

#Mean and median of total steps taken per day

```{r,echo=TRUE}
meanSteps <- mean(c$steps, na.rm = TRUE)
medSteps <- median(c$steps, na.rm = TRUE)
```
mean of total steps taken per day is `r meanSteps`
median of total steps taken per day is `r medSteps`

#Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days
```{r,echo=TRUE}
library(ggplot2)
Fu<-aggregate(steps~interval,Data,mean)
ggplot(Fu,aes(interval,steps))+geom_line()+
ggtitle("Average Daily Activity Pattern") +
        xlab("5-minute Interval") +
        ylab("Average Number of Steps") +
        theme(plot.title = element_text(hjust = 0.5))
```

#Which 5-minute interval across all days contain the maximum number of steps
```{r,echo=TRUE}
maxIn<-max(Fu$interval[which.max(Fu$steps)])
```
`r maxIn` interval contain the maximum steps.


#Calculate and report the total number of missing values in the dataset
```{r,echo=TRUE}
Total<-sum(is.na(Data$steps))
```
There are `r Total` missing values. I will replace these missing values with the 5-day average of that respective interval.


#The missing data filled in
```{r,echo=TRUE}
import<-transform(Data,steps=ifelse(is.na(Data$steps),Fu$steps[match(Data$interval,Fu$interval)],Data$steps))
```


#Histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r,echo=TRUE}
d<-aggregate(steps~date,import,sum)
hist(d$steps,main = "Imputed Number of Steps Per Day",
     xlab = "Number of Steps")
meanSteps2 <- mean(d$steps, na.rm = TRUE)
medSteps2 <- median(d$steps, na.rm = TRUE)
diffMean =meanSteps2 - meanSteps
diffMed = medSteps2 - medSteps
diffTotal = sum(d$steps) - sum(c$steps)
```

There is a difference of `r diffMean` in the mean steps of the two dataset. There is a difference of `r diffMed` in the median steps of the two dataset. There is a difference of `r diffTotal` in the total steps of the two dataset.


#Create a new factor variable in the dataset with two levels - "weekend" and "weekday"

```{r,echo=TRUE}
datatype<-function(date){
        day<-weekdays(date)
        if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
        else if (day%in% c("Saturday", "Sunday"))
                return("weekend")
        else
                stop("Invalid")
}
import$date=as.Date(import$date)
import$day<-sapply(import$date,FUN = datatype)
```


#A panel plot containnig a time-series plot of the 5-minute interval and the average number of steps taken across all weekdays or weekends.


```{r,echo=TRUE}
New<-aggregate(steps~interval+day,import,mean)
ggplot(New,aes(interval,steps,col=day))+geom_line()+facet_grid(day~.)+ggtitle("Average Daily Activity Pattern") +
        xlab("5-minute Interval") +
        ylab("Average Number of Steps") +
        theme(plot.title = element_text(hjust = 0.5))
```
