---
title: "Reproducible Research: Peer Assessment 1 - Isabel Méndez"
output: 
  html_document:
    keep_md: true
---

library(ggplot2)
## Loading and preprocessing the data

Load the data  (i.e. read.csv())
```{r}
if (!file.exists('activity.csv')) {
  unzip('activity.zip',overwrite=TRUE)
}
```
Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
activity <- read.csv("activity.csv", header = T, sep = ",")
```

## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day
```{r}
Totalsteps <- aggregate(steps - date, activityData, FUN=sum)
```

Histogram of the total number of steps taken each day
```{r}
hist(Totalsteps$steps, xlab = "Steps sum", main = "Steps taken per day")
```

Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean_Totalsteps <- round(mean(Totalsteps$steps, na.rm = TRUE))
median_Totalsteps <- round(median(Totalsteps$steps, na.rm =TRUE))
print(c("Mean",mean_Totalsteps))
print(c("Median",median_Totalsteps))
```


## What is the average daily activity pattern?
### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}

meanStepsInterval <- aggregate(steps ~ interval, activityData, mean)
ggplot(data = meanStepsInterval, aes(x = interval, y = steps)) +
  geom_line() +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))
```
# Which 5-minute interval across all days contain the maximum number of steps
```{r}

meanStepsInterval[which.max(meanStepsInterval)]
maxInt <- meanStepsByInt[which.max(meanStepsByInt$steps),]
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).
Total number of missing values in the dataset
```{r}
missingValues <- is.na(activityData$steps)
```
Strategy: replace  NA with the mean of the interval
Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r}
imp_activityData <- transform(activityData, steps = ifelse(is.na(activityData$steps), meanStepsInterval$steps[match(activityData$interval, 
                                                                        meanStepsInterval$interval)], activityData$steps))
```

Histogram of the total number of steps taken each day
```{r}
imp_StepsInterval <- aggregate(steps ~ date, imp_activityData, FUN=sum)
hist(imp_StepsInterval$steps,
     main = "Imputed Number of Steps Per Day",
     xlab = "Number of Steps")
```

New mean and median total number of steps taken per day
```{r}
imp_MeanSteps <- mean(imp_StepsInterval$steps, na.rm = TRUE)
imp_MedSteps <- median(imp_StepsInterval$steps, na.rm = TRUE)
diffMean = imp_MeanSteps - meanSteps
diffMed = imp_MedSteps - medSteps
diffTotal = sum(imp_StepsInterval$steps) - sum(totalSteps$steps)
print(c("imp_MeanStepsean=",imp_MeanSteps))
print(c("imp_MedSteps=",imp_MedSteps))
print(c("Diff Mean=",diffMean))
print(c("Diff Med=",diffMed))
```
The Difference between the two datasets is: Mean = 0.18, Median = 1.18
What is the impact of imputing missing data on the estimates of the total daily number of steps?
The% new values were set with the same value of the mean. Thus, the result will give mean and median values closer than the previous mean.
The mean and the median are the same: Mean = 10766, Median = 10766


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day
```{r}
DayType <- function(date) {
  day <- weekdays(date)
  if (day %in% c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'))
      return ("weekeday")
  else if (day %in% c('Saturday', 'Sunday'))
      return ("weekend")
  else
      stop ("Invalid Date Format! Try again.")
}
imp_activityData$date <- as.Date(imp_activityData$date)
imp_activityData$day <- sapply(imp_activityData$date, FUN = DayType)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r}
mean_StepsByDay <- aggregate(steps ~ interval + day, imp_activityData, mean)
ggplot(data = mean_StepsByDay, aes(x = interval, y = steps)) + 
  geom_line() +
  facet_grid(day ~ .) +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))

```
