---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
Here is where I load the data.
```{r loadData, echo=TRUE, results='asis'}
library(knitr)
library(ggplot2)
library(lubridate)
library(lattice)
dados <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
Plotting Histogram  
```{r separateMonth, echo=TRUE}
passosOctNov <- aggregate(steps ~ date, dados, sum)
hist(passosOctNov$steps, breaks = length(passosOctNov$date), main = "Histogrma de Passos por Dia", xlab = "Passos", ylab = "Frequencia")
```

Calculating mean and median
```{r meanMedian, echo=TRUE}
meanSteps <- format(mean(passosOctNov$steps, na.rm = TRUE), digits = 7, nsmall = 2)
medianSteps <- median(passosOctNov$steps, na.rm = TRUE)
```
The mean of steps is `r meanSteps` and the median of steps is `r medianSteps` 


## What is the average daily activity pattern?
Plotting time serie
```{r timeSeries, echo=TRUE}
passosInt <- aggregate(dados$steps ~ dados$interval, dados, mean)

graf <- ggplot(data = passosInt, aes(x=passosInt$`dados$interval`, y=passosInt$`dados$steps`))

graf <- graf + geom_line() + geom_point(color = "red")

graf <- graf + labs(title = "Time Serie 5-Minute Interval - Average Steps", x = "Interval" , y = "Average Steps")

print(graf)
```

```{r calcMax, echo=TRUE}
maximo <- passosInt[which.max(passosInt$`dados$steps`), 1]
```

The 5-Minute interval with the maximum number of steps is `r maximo `.


## Imputing missing values
Calculating NA values
```{r calcNA, echo=TRUE}
dadosNA <- dados[is.na(dados), ]
tamanhoDadosNA <- dim(dadosNA)
```
Total NA values in dataset is `r tamanhoDadosNA[1]`.

Replacing NA values with random values between range of passosOctNov$steps and creating a new dataset that is equal to the original but with the repalced values:
```{r replNA, echo=TRUE}
limites <- range(passosOctNov$steps)
numeros <- sample(0:200, tamanhoDadosNA[1], replace = TRUE)
dados$steps[which(is.na(dados$steps))] <- numeros
```

New histogram:
```{r}
passosOctNov2 <- aggregate(steps ~ date, dados, sum)
hist(passosOctNov2$steps, breaks = length(passosOctNov2$date), main = "Histogrma de Passos por Dia", xlab = "Passos", ylab = "Frequencia")
```

Calculating mean and median
```{r meanMedian2, echo=TRUE}
meanSteps2 <- format(mean(passosOctNov2$steps, na.rm = TRUE), digits = 7, nsmall = 2)
medianSteps2 <- median(passosOctNov2$steps, na.rm = TRUE)
```
The mean of steps is `r meanSteps2` and the median of steps is `r medianSteps2`

We can observe an increase in media and median values by adding sample values replacing NA's.

## Are there differences in activity patterns between weekdays and weekends?
First, we declare a function **tipoDia** which categorize weekdays in "weekend" or "weekday":
```{r tipoDia}
tipoDia <- function(dia){
        if (wday(dia) %in% c(1,7)){
                resultado <- 0
        } else {
                resultado <- 1
        }
        
        resultado
}
```

After that, we insert a new field called dayType in dataset:
```{r includeDayType, echo=TRUE}
passosInt <- aggregate(dados$steps ~ dados$interval, dados, mean)
dados$dayType <- sapply(dados$date, tipoDia)
```

Then, we make a new dataset which will be used for plot:
```{r newDataSet, echo=TRUE}
passosTipoDia <- aggregate(dados$steps ~ dados$interval + dados$dayType, dados, mean)
```

Factor variable called **fatores** 
```{r factorVariable, echo=TRUE}
fatores <- factor(passosTipoDia$`dados$dayType`, labels = c("weekend", "weekday"))

```

At least, we plot the time series:
```{r plottingChart, echo=TRUE}
xyplot(passosTipoDia$`dados$steps`~ passosTipoDia$`dados$interval` | fatores, type = "l", layout = c(1,2), main = "Average Steps by 5-Minute Interval", xlab = "Inteval", ylab = "mean(Steps)")
```

