---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Loading and preprocessing the data

- We will read in the data and get the total number, mean and median steps per day
- We consolidate results in a dataframe **Summarized_data** as follows:
```{r Preprocessing, echo = TRUE}

Activities<-read.csv("activity.csv",header = T)

No_steps_per_Day<-tapply(Activities$steps,Activities$date,sum)
Mean_steps_per_Day<-tapply(Activities$steps,Activities$date,mean)
Median_steps_per_Day<-tapply(Activities$steps,Activities$date,median)

Activities$Days_as_chr<-as.character(Activities$date)
Activities$Days_as_dates<-as.Date(Activities$Days_as_chr)
Avg_Days<-unique(Activities$Days_as_dates)

Summarized_data <- data.frame(
    day = Avg_Days,
    No_Steps = No_steps_per_Day,
    Mean = Mean_steps_per_Day,
    Median = Median_steps_per_Day
)
```

## What is mean total number of steps taken per day?

- We just computed total, mean and median steps per day, using functions **sum()**, **mean()** and **median()**
- A histogram and time series plot of total/mean steps per day is depicted, where we observe must frequent total steps are from 10k-15k steps, and highest total steps is around 20k.
- Similarly from  day 30 to day 40 we got the most frequent Average steps per Day,and highest mean steps per day is about 70. Overall **Mean steps is 37.4 steps per Day**
```{r Mean Steps, echo = TRUE}
par(mfrow = c(1,2))
hist(No_steps_per_Day,col = "red",main = "Number of steps per day",xlim = c(0,25000),
     xlab = "No of Steps range",ylab = "FREQUENCY")
with(Summarized_data,plot(day,No_Steps,type = 'l',main = "Number of steps per day",
                          xlab = "Days",ylab = "No. steps/day",col = "red",lwd = 3)) #makes a time series plot

par(mfrow = c(1,2))
hist(Mean_steps_per_Day,col = "blue",main = "Mean steps per day",xlim = c(0,100),
     xlab = "Mean No of Steps range",ylab = "FREQUENCY")
with(Summarized_data,plot(day,Mean,type = 'l',main = "Mean steps per day",
                          xlab = "Days",ylab = "Mean steps/day",col = "blue",lwd = 3)) #makes a time series plot

summary(Summarized_data)
```

## What is the average daily activity pattern?

- Using **ggplot package** we make a time series plot. Because there are some missing data, the plot will look discontinuous. 
- We can observe from the plot that *the day with highest number of steps was around Nov-21 of that year*.
```{r Timeseries plot, echo = TRUE}

library(ggplot2)
# A time series plot
ggplot(Summarized_data, aes(x=day, y=No_Steps)) +
    geom_line() + 
    geom_line( color="steelblue",size = 2) +
    xlab("") +
    labs(title = "Time series plot Average Daily No. of steps") 

```

## Imputing missing values

- We simply take the average from the previous averaged data, and substitute NA's values with this calculated average. **This average is 37.38 steps/day**. 
- We can always compare the Total steps before and after NA has been imputed. In general,average No. of steps remain the same as seen in the histograms.
- A new data set called **Summarized_data_NA** is created, to gather computations after NA imputation, including per interval statistics.
- A new *continuous* time series plot can be built, with no NA's in it.


```{r Imputing NA, echo = TRUE}

Value_To_Replace<-mean(Mean_steps_per_Day,na.rm=T)
Activities_no_NA<-Activities
Activities_no_NA[is.na(Activities$steps),1]<-Value_To_Replace

No_steps_per_Day_no_NA<-tapply(Activities_no_NA$steps,Activities_no_NA$date,sum)
No_steps_per_interval<-tapply(Activities_no_NA$steps,Activities_no_NA$interval,sum)
Mean_steps_per_interval<-tapply(Activities_no_NA$steps,Activities_no_NA$interval,mean)
Median_steps_per_interval<-tapply(Activities_no_NA$steps,Activities_no_NA$interval,median)
Avg_5min_int<-unique(Activities_no_NA$interval)

Summarized_data_no_NA <- data.frame(
    interval = Avg_5min_int,
    No_Steps = No_steps_per_interval,
    Mean = Mean_steps_per_interval,
    Median = Median_steps_per_interval
)

Summarized_data$No_Steps_no_NA<-No_steps_per_Day_no_NA

par(mfrow = c(1,2), oma = c(2,2,2,2))
hist(No_steps_per_Day,col = "red",main = "No. of steps per day with NA",xlim = c(0,25000),
     xlab = "No of Steps range",ylab = "FREQUENCY")
hist(No_steps_per_Day_no_NA,col = "blue",main = "No. of Steps per day without NA",xlim = c(0,25000),
     xlab = "No of Steps range",ylab = "FREQUENCY")

par(mfrow = c(1,1))
with(Summarized_data,plot(day,No_Steps_no_NA,type = 'l',main = "No. Steps per day",
                          xlab = "Days",ylab = "No steps/day",col = "cyan",lwd = 3))
```

## What is the 5-min interval with the max number of steps?

- We learned from the time series plot, that *the maximum Number of steps was around Nov-21st with a value of 21194 steps*. 
- Using the stats from the imputed dataset, We compute the 5-min interval that contains the maximum number of steps. *This turns out to be the 835th interval* 
- We can observe this in the following time series plot.

```{r 5 min interval with max no. of steps, echo = TRUE}

Maximum_Step<-max(No_steps_per_interval)
Activities_5min_int<-subset(Summarized_data_no_NA$interval,
                            Summarized_data_no_NA$No_Steps==Maximum_Step)
par(mfrow = c(1,1))
with(Summarized_data_no_NA ,plot(interval,Mean,type = 'l',main = "Avg Steps per interval",
                          xlab = "5-min interval",ylab = "Avg steps/interval",col = "green",lwd = 3))
abline(v=835,col = "black",lty=2,lwd = 2)

print(paste("The 5 min interval with max avg of steps is:", Activities_5min_int))
```

## Are there differences in activity patterns between weekdays and weekends?

- We use the function **weekdays()** to discriminate weekdays and weekend days.
- We also use **dplyr package** to average the number of steps per 5 min interval by weekdays and weekends categories.
- We  create a plot using **lattice package**, to compare activity pattern between weekdays and weekends.
- From the plots *it appears on weekends people is active throught the day, whereas on the weekdays, they are more active in morning/noon hours*. 
```{r activity patterns, echo = TRUE, results="hide"}

Activities_no_NA$Days_of_week<-weekdays(Activities_no_NA$Days_as_dates)
Activities_no_NA$DayType<-ifelse(Activities_no_NA$Days_of_week=="Saturday"|Activities_no_NA$Days_of_week=="Sunday",
                                    "Weekend","Weekday")
library(dplyr)
tbl_df(Activities_no_NA)
Activities_Summarized <-
    Activities_no_NA %>%
    group_by(interval,DayType) %>%
    summarize(m = mean(steps),v = var(steps)) %>%
    print
```

```{r activity patternsfig, echo = TRUE}
library(lattice)
xyplot(m ~ interval | DayType, data = Activities_Summarized, layout = c(1, 2))
```
