---

title: "Reproducible Research: Peer Assessment 1"
output: html_document
keep_md: true

---
# Reproducible Research: Peer Assessment 1

## 0. Loading the libraries required to run the script

```{r, echo = TRUE, results = 'hide'}
library(data.table)
library(ggplot2)
library(lubridate)
library(dplyr)
```

## 1. Code for reading in the dataset and/or processing the data

```{r loadandprep, echo = TRUE}
## Downloading the file

url <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
dest <- paste0(getwd(), "/" , "Activity_monitoring_data.zip")
download.file(url, dest)
zipF<- "./Activity_monitoring_data.zip"
outDir<-"./Activity_monitoring_data"
unzip(zipF, exdir = outDir)
setwd("./Activity_monitoring_data")

## Reading data set

Activity_monitoring <- read.table("activity.csv", sep = ",", header = TRUE)
setwd("..")

## Date format

Activity_monitoring$date <- as.Date(as.character(Activity_monitoring$date))
ndate <- as.POSIXlt(Activity_monitoring$date)
ndate$hour <- ndate$hour + Activity_monitoring$interval%/%100
ndate$min <- ndate$min + Activity_monitoring$interval%%100
Activity_monitoring$int_label <- paste(paste(Activity_monitoring$interval%/%100, 
                                 "h", sep = " "), 
                                 paste(Activity_monitoring$interval%%100, "min", 
                                 sep = " "), sep = ":")
head(Activity_monitoring, 10)
```

## 2. Histogram of the total number of steps taken each day

### Here is the summary of the steps taken per day

```{r totali, echo = TRUE}
## Steps taken each day

steps_sum <- aggregate(Activity_monitoring$steps, by = list(Category = 
                       Activity_monitoring$date), FUN=sum, na.rm = TRUE)
names(steps_sum) <- c("date", "steps_sum")
head(steps_sum, 10)
```

### Histrogram

```{r histogrami, echo = TRUE}
g <- ggplot(data = steps_sum, aes(steps_sum))
        g + 
        geom_histogram(bins = 61, aes(fill = ..count..)) + 
        ggtitle("Histogram of the total number of steps taken each day") +
        theme(plot.title = element_text(hjust = 0.5))
```

## 3. Mean and median of the number of steps taken each day

```{r meanmediani, echo = TRUE}
summary_steps <- group_by(Activity_monitoring, date)
summary_steps <- summarize(summary_steps, 
                           steps_sum = sum(steps, na.rm = TRUE), 
                           steps_mean = mean(steps, na.rm = TRUE), 
                           steps_median = median(steps, na.rm = TRUE))
head(summary_steps)
```

## 4. Time series plot of the average number of steps taken

```{r timesseries1i, echo = TRUE}
g <- ggplot(data = summary_steps, aes(date, steps_mean))
        g + 
        scale_x_date(date_labels = "%b %d", 
                     date_breaks = "5 days", 
                     date_minor_breaks = "1 day") +       
        geom_line(na.rm = TRUE, color = "blue") +
        geom_point(na.rm = TRUE, color = "blue", size = 1, alpha = 0.5) + 
        xlab("Date") + 
        ylab("Average number of steps taken") +
        ggtitle("Average number of steps taken per day") +
        theme(plot.title = element_text(hjust = 0.5)) 
```

## 5. The 5-minute interval that, on average, contains the maximum number of steps

### First the data frame is grouped by interval

```{r intervali, echo = TRUE}
summary_interval <- group_by(Activity_monitoring, interval)
summary_interval <- summarize(summary_interval, 
                              steps_sum = sum(steps, na.rm = TRUE), 
                              steps_mean = mean(steps, na.rm = TRUE), 
                              steps_median = median(steps, na.rm = TRUE))
head(summary_interval, 10)
```

### Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r timesseries2i, echo = TRUE}
g <- ggplot(data = summary_interval, aes(interval, steps_mean))
        g + 
        geom_line(na.rm = TRUE, color = "blue") +
        geom_point(na.rm = TRUE, color = "blue", size = 1, alpha = 0.5) + 
        scale_x_continuous(breaks = seq(0, 2360, 100)) +
        xlab("Interval") + 
        ylab("Average number of steps taken") +
        ggtitle("Average number of steps taken in every 5-minute interval") +
        theme(plot.title = element_text(hjust = 0.5)) 
```


```{r maxi, echo = TRUE}
max_int <- summary_interval[which.max(summary_interval$steps_mean), 1]
max_int <- paste(paste(max_int%/%100, "h", sep = " "), 
                 paste(max_int%%100, "min", sep = " "), sep = ":")
```

### The 5-minute interval that, on average, contains the maximum number of steps is at `r max_int`.

## 6. Code describing and showing the strategy for imputing missing data

### Total number of missing values in the dataset

```{r missing, echo = TRUE}
data.na <- sum(!complete.cases(Activity_monitoring))
```

### The total number of missing values in the dataset is `r data.na`. 

### As a strategy for filling in all of the missing values in the datase I used the mean for each that 5-minute interval

```{r replacem, echo = TRUE}
## Function to replace the na values with the mean

fill_mean <- function(x) {
        replace(x, is.na(x), mean(x, na.rm = TRUE))
}

## Create a new dataset that is equal to the original dataset but with the 
## missing data filled in.

Activity_monitoring_c <- Activity_monitoring %>%
        group_by(interval) %>%
        mutate(steps = fill_mean(steps))
Activity_monitoring_c <- ungroup(Activity_monitoring_c)
head(Activity_monitoring_c, 10)
```

### Steps taken each day after missing values are imputed

```{r totalf, echo = TRUE}
## Steps taken each day
steps_sum_c <- aggregate(Activity_monitoring_c$steps, 
                by = list(Category = Activity_monitoring_c$date), 
                FUN = sum, na.rm = TRUE)
names(steps_sum_c) <- c("date","steps_sum")
head(steps_sum_c)
```

### Mean and median of the number of steps taken each day after missing values are imputed

```{r meanmedianf, echo = TRUE}
## Calculate and report the mean and median total number of steps taken per day
summary_steps_c <- group_by(Activity_monitoring_c, date)
summary_steps_c <- summarize(summary_steps_c, 
                             steps_sum = sum(steps, na.rm = TRUE), 
                             steps_mean = mean(steps, na.rm = TRUE), 
                             steps_median = median(steps, na.rm = TRUE))
head(summary_steps_c)
```

### Comparison with the values from the estimates from the first part of the assignment

```{r comparison, echo = TRUE}
comparison <- data.frame(summary_steps$date, 
                         summary_steps$steps_sum, 
                         summary_steps_c$steps_sum,
                         summary_steps$steps_mean, 
                         summary_steps_c$steps_mean, 
                         summary_steps$steps_median, 
                         summary_steps_c$steps_median)
names(comparison) <- c("date", "sum_steps_i", "sum_steps_f", 
                       "mean_steps_i", "mean_steps_f",
                       "median_steps_i", "median_steps_f")
head(comparison, 61)
```

## 7. Histogram of the total number of steps taken each day after missing values are imputed

```{r histogramf, echo = TRUE}
g <- ggplot(data = steps_sum_c, aes(steps_sum))
        g + 
        geom_histogram(bins = 61, aes(fill = ..count..)) + 
        ggtitle("Histogram of the total number of steps taken each day") +
        theme(plot.title = element_text(hjust = 0.5))
```

## 8.  Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

### New factor variable in the dataset (after missing values are imputed) with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r day, echo = TRUE}
Weekday <- c("Monday", "Tuesday", "Wednesday", "Tuesday", "Friday")
Activity_monitoring_c$day <- ifelse(weekdays(Activity_monitoring_c$date) %in% 
                                            Weekday, 'weekday', 'weekend')
```

### Panel plot

```{r panel, echo = TRUE}
g <- ggplot(data = Activity_monitoring_c, aes(interval, steps))
        g + 
        facet_wrap(~ day, nrow = 2, ncol = 1) +    
        stat_summary(fun.y = "sum", geom = "line", col = "blue") + 
        scale_x_continuous(breaks = seq(0, 2360, 100)) +
        xlab("Interval") + 
        ylab("Total number of steps taken") +
        ggtitle("Total number of steps taken in every 5-minute interval") +
        theme(plot.title = element_text(hjust = 0.5)) 
```

### As shown there are differences in activity patterns between weekdays and weekends

```{r final, echo = TRUE}
summary_all <- Activity_monitoring_c %>% group_by(interval, day) %>% summarise(sum = sum(steps))
head(summary_all)
```
