---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
This report is the analysis of an invidual's walking activity collected during the months of October 2012 and November 2012, by a monitoring device.The data comes from Coursera's Reproducible Research Class.   

## 1. The first step in the analysis consists of downloading the data from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

## create file if it does not exist
if (!file.exists("Rsteps")) {
  dir.create("Rsteps")
}
##Download url
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = "./Rsteps/steps.zip", method = "curl")
dateDownloaded <- date()

##unzip and read file plus change date from factor to POSIXct 
ss <-unzip("./Rsteps/steps.zip", exdir = "./Rsteps")
dat <- read.csv("./Rsteps/activity.csv")
dat$date <-as.POSIXct(dat$date, format="%Y-%m-%d")
date <- as.Date(dat$date)
date2 <- unique(date)
```

## 2. Calculate the number of steps per day and produce a histogram.

```{r }
knitr::opts_chunk$set(echo = TRUE)
sum1 <- aggregate(.~date, data=dat, sum, na.action=NULL)
hist(sum1$steps, main = "Frequency of the Number of Steps per Day", xlab = "Steps per Day", ylab = "Frequency")



```

## 3. Calculate the median and mean number of steps per day.

```{r pander}
mean1 <- aggregate(.~date, data=dat, mean, na.action=NULL)
median1 <- aggregate(.~date, data=dat, median, na.action=NULL)
median_mean <- cbind(mean1[1:2], median1[2])
colnames(median_mean)<- c("date","mean-steps","median-steps")
mean2 <- aggregate(.~interval, data=dat, mean)


library(pander)
panderOptions("digits", 4)
pander(median_mean)

#library(xtable)
#print(median_mean, type="html")
#library(knitr)
#kable(median_mean)
```

## 4. Time series plot for the average number of steps taken

```{r }

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), oma = c(0, 0, 2, 0))
plot(x = median_mean$date, y=median_mean$mean, type = "l", ylab = "average steps taken", xlab = "Individual Day mean")
plot(x = mean2$interval, y=mean2$steps, type = "l", ylab = "average steps taken", xlab = " consecutive 5 minute intervals \n across the day")




```

## 5. Determine the max interval

```{r}

max(median_mean$`mean-steps`, na.rm=TRUE)

which(median_mean$`mean-steps` == max(median_mean$`mean-steps`, na.rm=TRUE))


```

## 6. Imputing missing data by taking the average number of steps across all days and times and replacing the "NA".

```{r}

dat3 <- dat
dat3$steps[is.na(dat3$steps)] <- as.numeric(37)
dat3$steps <- as.numeric(as.character(dat3$steps))

#mean_steps_v2 <-tapply(dat3$steps, dat3$interval, FUN=mean)
#median_steps_v2 <-tapply(dat3$steps, dat3$interval, FUN=median)

mean_steps_v2 <- aggregate(.~date, data=dat3, mean, na.action=NULL)
median_steps_v2 <- aggregate(.~date, data=dat3, median, na.action=NULL)
median_mean2 <- cbind(mean_steps_v2[1:2], median_steps_v2[2])
colnames(median_mean2)<- c("date","mean-steps","median-steps")
mean3 <- aggregate(.~interval, data=dat3, mean)


library(pander)
panderOptions("digits", 4)
pander(median_mean2)

#library(xtable)
#print(median_mean, type="html")
#library(knitr)
#kable(median_mean)








```

## 7. Plotting the incorparted imputed data with original data

```{r}
maintitlev2 = "Time Series of 5 minute interval averaged across all days \n with missing data imputed"
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), oma = c(0, 0, 2, 0))
plot(x = median_mean2$date, y=median_mean2$mean, type = "l", ylab = "average steps taken", xlab = "Individual Day mean")
plot(x = mean3$interval, y=mean3$steps, type = "l", ylab = "average steps taken", xlab = " consecutive 5 minute intervals \n across the day")


```

## 8. Determining if individual is more active on weekends or weekdays

```{r}
dat4 <- dat3
dat4$dayOFweek <- weekdays(as.Date(dat4$date))
dat7 <- subset(dat4, dayOFweek =="Sunday" | dayOFweek =="Saturday")
dat8 <- subset(dat4, dayOFweek !="Sunday")
dat8 <- subset(dat8, dayOFweek !="Saturday")
wkend <-tapply(dat7$steps, dat7$interval, FUN=mean)
wkday <-tapply(dat8$steps, dat8$interval, FUN=mean)
mainmain <- "Comparison of steps for Weekend vs Weekday"
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), oma = c(0, 0, 2, 0))
plot(wkend, type = "l", main = "Weekend Steps", ylab = "steps", xlab = "interval", ylim= c(0,200))
plot(wkday, type = "l", main = "Weekday Steps", ylab = "steps", xlab = "interval", ylim= c(0,200))
title(mainmain, outer = TRUE, cex = 0.5)

```



