## Loading and preprocessing the data
```{r, echo=TRUE}

if(!file.exists("activity.csv")){
    unzip("activity.zip") 
}

stepsTakenActivity <- read.csv("activity.csv", header=TRUE)

# Let's look at the dimensions at our data frame
dim(stepsTakenActivity)

# Take steps variable  as a dependant variable while date is an independane variable
stepsPerDay <- aggregate(steps ~ date, stepsTakenActivity, sum)

# Translate date variable from a  string type to a date type of a given format
stepsTakenActivity <- as.Date(stepsTakenActivity$date, format = '%Y-%m-%d')

# Process our Data Frame by finding the total rows 
TotalRows <- nrow(stepsTakenActivity)
omitNA    <-  na.omit(stepsTakenActivity)

```

## What is mean total number of steps taken per day?
We first draw the histogram of the total number of steps taken each day and then,
calculate and report

*  The Mean

*  The Median

of the total of steps taken per day.

```{r, echo=TRUE}
hist(stepsPerDay$steps, main = "Total Steps Per Day", col="blue", xlab="Number of Steps")
```

####  Mean for the Total number of steps taken per day:
```{r, echo=TRUE}
mean(stepsPerDay$steps)

```

####  Median for the Total number of steps taken per day:
```{r, echo=TRUE}
median(stepsPerDay$steps)

```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?


# http://www.cookbook-r.com/Manipulating_data/Summarizing_data/

