# Reproducible Research: Peer Assessment 1
=============================================================

Data of steps given by a person, measured at 5 minutes interval across the day, was registered for October and November of 2012, with a Nike Fuelband device and a samsung galaxy S4 with android 4.1 phone.

The data was downloaded from <a href="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"> this url </a>, consulted in 15-07-2014 in teh working directory of R-Studio 0.98.507.  


## Loading and preprocessing the data

The data was loaded and processed, keeping only the complete cases (rows without NA) as valid data, the rows with al least one NA value was deleted. 
```{r Parte 1, results='hide'}
rm(list=ls()) #cleaning vars
cat("\014")  #clear screen

## LOADING DATA
crudos<-read.table("activity.csv",sep=",",header=T) #names steps date interval

## SELECTING VALID DATA (non na data, days with non data was excluded)
datos<-crudos[complete.cases(crudos),]

```

## What is mean total number of steps taken per day?

The total, mean and median of the steps per each day, was calculated.
```{r,results='hide'}
# Grouping data per day
A<-1;AA<-1;AAA<-1;fechas<-datos$date[datos$interval==0]
for (i in 1:max(as.numeric(datos$date))) {
  A[i]<-mean(datos$steps[as.numeric(datos$date)==i])
  AA[i]<-sum(datos$steps[as.numeric(datos$date)==i])
  AAA[i]<-median(datos$steps[as.numeric(datos$date)==i])  
}

# Total, mean and median of steps per day
DD<-data.frame(steps=AA[AA!=0],mean=A[AA!=0],median=AAA[AA!=0],date=as.Date(fechas))
```

In brief, we have the following scenario for each day analyzed:
```{r}
DD
```

A histogram of the steps in an average day, is shown below:
```{r fig.width=7, fig.height=6}
# histogram of total
hist(DD$steps,xlab="interval")
```


## What is the average daily activity pattern?

The data was grouped per interval, across all days, and their mean, median and total steps were calculated.
```{r,results='hide'}
## INTERVAL ANALYSIS
#Grouping dat per interval
B<-1;BB<-1;BBB<-1;intervalos<-datos$interval[as.numeric(datos$date)==2]
for (i in 1:length(intervalos)){
  B[i]<-sum(datos$steps[datos$interval==intervalos[i]])
  BB[i]<-mean(datos$steps[datos$interval==intervalos[i]])
  BBB[i]<-median(datos$steps[datos$interval==intervalos[i]])
}

# Total, mean and median of steps per interval
ID<-data.frame(steps=B,mean=BB,median=BBB,interval=intervalos)
```

The table below shows the steps per interval:
```{r}
ID
```

A plot of the mean of steps per interval is as follow:
```{r fig.width=7, fig.height=6}
#Plot steps per interval
plot(ID$interval,ID$mean,xlab="interval",ylab="mean")
```

The interval with the maximum steps in average is:
```{r}
#Interval with max average steps
ID[ID$mean==max(ID$mean),]
```


## Imputing missing values

The total number of Na values in raw data is:

```{r,results='asis'}
NANumber<-length(crudos$steps[is.na(crudos$steps)])
NANumber
```

Considering the total the raw data, the NA values represent a percentage of (expressed as decimal):
```{r,results='asis'}
NANumber/length(crudos$steps)
```

Each interval was inspected, and all NA values was replaced for the respective interval mean across all days (considering only valid data). 
```{r,results='hide'}
##MISSING DATA
#Number of NA
NANumber<-length(crudos$steps[is.na(crudos$steps)])

#Replacing the NA values (for the mean of the interval)
RD<-crudos;temp<-0
for (i in 1:length(RD$steps)){
  if(is.na(RD$steps[i])) {RD$steps[i]<-ID$mean[ID$interval==RD$interval[i]]}
}
```

A sample of the original and replaced data is shown below:
```{r}
data.frame(crudos[280:300,],RD[280:300,],Int.mean=c(ID$mean[280:288],ID$mean[1:12]))
```
From left to right, the column 1, 2 and 3 represent the raw data. Columns 4, 5 and 6 represent the replaced data, and the last one show the mean of the respective interval.

A histogram of the steps in an average day, of the replaced data is shown below:
```{r fig.width=7, fig.height=6}
# histogram of total
hist(RD$steps,xlab="interval")
```

Replaced data were grouped by day, as was done with the previous set.
```{r,results='hide'}
#Grouping NEW data per day
AR<-1;AAR<-1;AAAR<-1; fechasR<-as.Date(RD$date[RD$interval==0])

for (i in 1:length(fechasR)) {
  AR[i]<-sum(RD$steps[as.numeric(RD$date)==i])
  AAR[i]<-mean(RD$steps[as.numeric(RD$date)==i])
  AAAR[i]<-median(RD$steps[as.numeric(RD$date)==i])
}

# Total, mean and median of steps per day, with replaced data
RDD<-data.frame(steps=AR[AR!=0],mean=AAR[AR!=0],median=AAAR[AR!=0], date=as.Date(fechasR))

#Histogram
#hist(RDD$steps)
```

The scenario for each day, with replaced data is:
```{r}
# Total, mean and median of steps per day, with replaced data
RDD
```

## Are there differences in activity patterns between weekdays and weekends?

A final analysis was made, looking for differences in the mean of the intervals, in days of week and weekend days. The data was divided, according if came from a day of the week or of a day of weekend.
```{r}
##WEEK WEEKEND PATTERN (in revised data)
# Creating two levels variable with week and weekend days
days<-as.factor(c("weekday","weekend"))
for (i in 1:length(RD$date)){
  if(weekdays(as.Date(RD$date[i])) == "sabado" || weekdays(as.Date(RD$date[i])) == "domingo")
  {days[i]="weekend"}
  if(weekdays(as.Date(RD$date[i])) != "sabado" & weekdays(as.Date(RD$date[i])) != "domingo")
  {days[i]="weekday"} 
}
RDW<-data.frame(RD,days)

#Grouping data per interval of weekday end weekend
C<-1;D<-1;intervalos<-datos$interval[as.numeric(datos$date)==2]
for (i in 1:length(intervalos)){ 
    C[i]<-mean(RDW$steps[RDW$interval==intervalos[i] & RDW$days=="weekday"])
    D[i]<-mean(RDW$steps[RDW$interval==intervalos[i] & RDW$days=="weekend"])  
}




```


Figures 3 and 4 show the respective pattern for a week day and weekend respectively. 
```{r}
par(mfrow=c(2,1))
plot(intervalos,C, main=" Figure 3: weekdays")
plot(intervalos,D, main=" Figure 4: weekend")
```
