---
title: 'Reproducible Research: Peer Assessment 1'
author: "Mostafa Amin"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
```

## Loading and preprocessing the data
```{r}
activity <- read.csv("activity.csv", sep = ",", header = TRUE)
```

## What is mean total number of steps taken per day?

###Number of steps per day
```{r}
steps <- activity %>%
  group_by(date) %>%
  summarize(steps = sum(steps))
```

###Histogram of the total number of steps taken each day
```{r}
hist(steps$steps, xlab = "Steps per Day", main = "Total number of steps taken each day", col = "turquoise")
```

###The mean and median total number of steps taken per day
```{r}
mean <- mean(steps$steps, na.rm=TRUE)
median <- median(steps$steps, na.rm=TRUE)
```

```{r}
mean

median
```

##What is the average daily activity pattern?

###Average number of steps per interval
```{r}
interval <- activity %>%
  group_by(interval) %>%
  summarize(steps = mean(steps, na.rm = TRUE))
```

###time series plot
```{r}
plot.ts(interval$steps, xlab = "Interval", ylab = "Average steps", col = "Maroon")
```

###The interval with maximum number of steps
```{r}
interval[which.max(interval$steps),]
```

## Imputing missing values

###Calculate and report the total number of missing values in the dataset
```{r}
sum(is.na(activity))
```

###Imputing strategy and new dataset
```{r}
impact <- activity 
for (i in 1:nrow(impact)) {
    if (is.na(impact$steps[i])) {
        impact$steps[i] <- interval[which(impact$interval[i] == interval$interval), ]$steps
    }
}
```

```{r}
sum(is.na(impact))
```

###Number of steps per day of the imputed data
```{r}
imp_steps <- impact %>%
  group_by(date) %>%
  summarize(steps = sum(steps))
```

###Histogram of the total number of steps taken each day of the imputed data
```{r}
hist(imp_steps$steps, xlab = "Steps per Day", main = "Total number of steps taken each day of the imputed data", col = "orange")
```

###The mean and median total number of steps taken per day of the imputed data
```{r}
imp_mean <- mean(imp_steps$steps, na.rm=TRUE)
imp_median <- median(imp_steps$steps, na.rm=TRUE)
```

```{r}
imp_mean

imp_median
```

Imputing the data resulted in mean and median became the same value

## Are there differences in activity patterns between weekdays and weekends?

### Creating new variable (daytype)
```{r}
library(chron)
impact$daytype = chron::is.weekend(impact$date)
impact$daytype <- ifelse(impact$daytype=="TRUE", "weekend", "weekday")
```

### Average steps
```{r}
interval_wk <- impact %>%
  group_by(interval, daytype) %>%
  summarise(steps = mean(steps))
```

###The plot
```{r}
library(ggplot2)
plot <- ggplot(interval_wk, aes(x=interval, y=steps, color = daytype)) +
  geom_line() +
  facet_wrap(~daytype, ncol = 1, nrow=2)
print(plot)
```
