Reproducible Research\Peer Assessment 1
========================================================

```{r echo=FALSE, results='hide'} 
setwd("C:\\Mac Move\\eBooks\\Coursera\\Specialization\\Data Science\\05 Reproducible Research\\repdata-data-activity")
library(plyr) 
library(lattice) 
```

## Loading and preprocessing the data
```{r} 
#Load the data (i.e. read.csv())
activity=read.csv("activity.csv")
#Processing the data into a format suitable for analysis
data1=ddply(activity,'date',function(x) c(TotalSteps=sum(x$steps)))
data2=ddply(activity,'interval',function(x) c(AvgSteps=mean(x$steps,na.rm=TRUE)))
```

## What is mean total number of steps taken per day?
```{r fig.width=7, fig.height=6}  
#Histogram of the total number of steps taken each day
  hist(data1$TotalSteps,main="Histogram of the total number of steps taken each day",xlab="Total Steps taken per day",xlim = c(0, 25000), ylim = c(0, 35))
```
**For the total number of steps taken:**

- *the Mean is `r mean(data1$TotalSteps, na.rm=TRUE)` &*
- *the Median is `r median(data1$TotalSteps, na.rm=TRUE)`*

## What is the average daily activity pattern?
```{r fig.width=7, fig.height=6}
#Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
plot(data2$interval,data2$AvgSteps,type="l",xlim=c(0, 2500), ylim=c(0, 200),main="Average daily activity pattern",xlab="Interval", ylab="Number of Steps")  
```


```{r results='hide',echo=FALSE}
##Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
  interval=data2[which.max(data2$AvgSteps),]
```
On Average across all the days in the data set, **Interval `r interval$interval`** has the maximum number of steps

## Imputing missing values
The total number of missing values in the dataset (i.e. the total number of rows with NAs) are: **`r nrow(activity[!complete.cases(activity),])`**

**STRATEGY**: The NAs are filled with the mean from the 5-minute interval
```{r} 
#The NA are filled with the mean from the 5-minute interval and a new dataset is created
activity_NA_filled_w_Mean = activity
activity_NA_filled_w_Mean[is.na(activity_NA_filled_w_Mean)] <- mean(data2$AvgSteps,na.rm=TRUE)
data3=ddply(activity_NA_filled_w_Mean,'date',function(x) c(TotalSteps=sum(x$steps)))

```

```{r fig.width=7, fig.height=6}
#Histogram of the total number of steps taken each day
hist(data3$TotalSteps,main="Histogram of the total number of steps taken each day",sub="After filling the missing values (NA), with mean from the 5-minute interval",xlab="Total Steps taken per day",xlim = c(0, 25000), ylim = c(0, 35))
```

**For the total number of steps taken:**

- *the Mean is `r mean(data3$TotalSteps, na.rm=TRUE)` &*
- *the Median is `r median(data3$TotalSteps, na.rm=TRUE)`*

**Difference between estimates**
Frequency between the range of steps 10000 and 15000 has increased from a frequency of 25 to 35. Comparing the mean from the 2 estimates, there is no significant change.

Mean of Estimate1: **`r mean(data1$TotalSteps, na.rm=TRUE)`**  vs Mean of Estimate2: **`r mean(data3$TotalSteps, na.rm=TRUE)`**

```{r fig.width=7, fig.height=6}
#Differ between the estimates from the first part of the assignment! 
  par(mfrow=c(2,1))
  hist(data1$TotalSteps,main="Histogram of the total number of steps taken each day",xlab="Total Steps taken per day",xlim = c(0, 25000), ylim = c(0, 35))
  hist(data3$TotalSteps,main="Histogram of the total number of steps taken each day",sub="After filling the missing values (NA), with mean from the 5-minute interval",xlab="Total Steps taken per day",xlim = c(0, 25000), ylim = c(0, 35))
```
```{r echo=FALSE}
  par(mfrow=c(1,1))
```


## Are there differences in activity patterns between weekdays and weekends?
```{r}
#Creating new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
activity_NA_filled_w_Mean$day=weekdays(as.Date(activity_NA_filled_w_Mean$date))
activity_NA_filled_w_Mean$weekday = NA

#Creating factors Weekend and Weekday
activity_NA_filled_w_Mean[row.names(subset(activity_NA_filled_w_Mean,day %in% c("Saturday","Sunday"))),"weekday"]="weekend"
activity_NA_filled_w_Mean[row.names(subset(activity_NA_filled_w_Mean,!(day %in% c("Saturday","Sunday")))),"weekday"]="weekday"

#Creating Dataset for Weekday and Weekend analysis
data_weekday=subset(activity_NA_filled_w_Mean,weekday=='weekday')
data_weekend=subset(activity_NA_filled_w_Mean,weekday=='weekend')

#Dataset to show the difference in activity patterns
data7=ddply(activity_NA_filled_w_Mean,c('interval','weekday'),function(x) c(AvgSteps=mean(x$steps,na.rm=TRUE)))
```  
**Comparison between Weekday and Weekend on the Average number of steps taken per day**
```{r fig.width=7, fig.height=6}  
xyplot(data7$AvgSteps~data7$interval|data7$weekday,type="l",layout=c(1,2),xlab="Interval",ylab="Number of steps")  
```
