---
title: "RR week 1"
author: "MV"
date: "March 7, 2016"
output: html_document
---

Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.


Dataset: Activity monitoring data [52K]
The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

date: The date on which the measurement was taken in YYYY-MM-DD format

interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

The data for this assignment was downloaded from the course web site and read with the read.csv R syntax:

```{r}
act <- read.csv("activity.csv",
            colClasses = c("numeric", "character","integer"))
summary(act)
```

In the next steps, the data is aggregated by date in order to:

1. Calculate the total number of steps taken per day
2. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
actAgg <- tapply(act$steps, act$date, FUN = sum, na.rm = TRUE)
print(actAgg)

library(lubridate)
act$date <- ymd(act$date)
summary(actAgg)
library(data.table)
actAgg_dt=data.table(actAgg)
summary(actAgg_dt)

mean(actAgg_dt$V1)
median(actAgg_dt$V1)

library(plyr)
library(dplyr)

steps1 <- act %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print


barplot(steps1$steps, names.arg=steps1$date, 
        main= 'Total Number of Steps per day',
        xlab='Dates', ylab='Steps')
```

What is the average daily activity pattern?


```{r]}
intervl <- act %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps)) %>%
  print    
```

Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
plot(intervl, type = 'l')
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
intervl[which.max(intervl$steps), ]$interval
```
Imputing missing values

There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).
```{r}
#values missing in orIginal dataset
missing1 <- sum(is.na(act))
missing1
```

Create a new dataset with the missing data filled in:
```{r}
act_miss_rep  <- transform(act, steps = ifelse(is.na(steps), round(mean(steps, na.rm=TRUE)), steps))

```

missing values replaced verification
```{r}
missing2 <- sum(is.na(act_miss_rep))
missing2
```

Summary of new dataset with missing values replaced:
```{r}
#values missing in orginal dataset
summary(act_miss_rep)
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
```{r}
steps2 <- act_miss_rep %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print      

barplot(steps2$steps, names.arg=steps2$date, 
        main= 'Total numbers of Steps per day',
        xlab='Dates', ylab='Steps - *missing values replaced with the mean of steps')
```
Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
#aggregate corrected data by date to calculate daily stats 

actAgg2 <- tapply(act_miss_rep$steps, act_miss_rep$date, FUN = sum, na.rm = TRUE)
act$date <- ymd(act_miss_rep$date)

#mean and median total number of steps taken per day
mean(actAgg2)
median(actAgg2)

#DO VALUES DIFFER?
diff_in_means=mean(actAgg)-mean(actAgg2)
diff_in_means

# Do the data estimates differ from the first part of the assignment?

summary(actAgg)

summary(actAgg2)

#What is the impact of imputing missing data on the estimates 
#of the total daily number of steps?

summary(actAgg2)-summary(actAgg)


```

#third part

Are there differences in activity patterns between weekdays and weekends?
 
A new factor variable (d_type) in a third dataset (act_weekday_ind) has been created.  It has   two levels - "weekday" and "weekend"-- indicating whether a given date is a weekday or weekend day.

```{r}
act_weekday_ind=act_miss_rep
act_weekday_ind$d_type[as.POSIXlt(act_weekday_ind$date)$wday %in% c(0,6)] <- "weekday"
act_weekday_ind$d_type[as.POSIXlt(act_weekday_ind$date)$wday %in% c(1:5)] <- "weekend"
table(act_weekday_ind$d_type,as.POSIXlt(act_weekday_ind$date)$wday)
```
Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).  
```{r}
actAgg3 <- act_weekday_ind  %>% 
  group_by(d_type, interval) %>%
  summarize(steps3 = mean(steps))
```

plot

```{r}
library(ggplot2)
ggplot(actAgg3, aes(interval,steps3))+geom_line(color="blue")+
    facet_wrap(~d_type, ncol=1)

````
