---
title: "PA1ArturTuros"
author: "Artur Turos"
date: "11 stycznia 2016"
output: html_document
---

Loading and preprocessing the data:

```{r}
#Read a file and save to the variable
activity<-read.csv("C:/temp/activity.csv", header=TRUE)
#Make copy of the data frame
activityNA<-activity
#Show a data
print(activityNA)
#Remove the NAs and save to another data frame
activity<-na.omit(activity)
```

What is mean total number of steps taken per day (ignoring the missing values)?

1. Calculate the total number of steps taken per day

```{r}
#Total number of steps taken per day
TotNoStepsPerDay <- aggregate(steps ~ date, activity, sum)
#Attach the data
attach(TotNoStepsPerDay)
#Create a histogram
color<-c('yellow','blue','red','purple','black','green')
hist(steps, breaks = 7, main = "Histogram of total number of steps per day", xlab = "Number of steps per day", ylab = "Frequency", col = color)
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
#Mean
mean(steps)
#Median
median(steps)
```

What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
#Average steps per interval for each day
StepsInterval <- aggregate(steps ~ interval, activity, mean)
#Attach a data
attach(StepsInterval)
#Plot
plot(interval, steps, type='l', main="Average steps per interval for each day", xlab="Intervals", ylab="Avg no of steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
#Finding a row id
rowID <- which.max(steps)
#Which interval
MaxAvgStepsInterval<-data.frame(StepsInterval[rowID,])
print(MaxAvgStepsInterval)
```

Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
#Number of the NAs
sum(is.na(activityNA))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

The NAs values in the quantitative attributes will be replaced by the mean value and in the qualitative attributes will be replaced by the mode value.

3. The NAs values in the quantitative attributes will be replaced by the mean value and in the qualitative attributes will be replaced by the mode value.

```{r}
#Missing insert function
missing.insert = function(x,na.string = NA){
  
if(missing(x))
  stop("You must specify a variable that stores data set!")

znak<-as.character(na.string)
x<-data.frame(x)
r<-nrow(x)
c<-ncol(x)

if(!is.na(znak)){
  for(i in 1:r){
    for(j in 1:c){
      if(x[i,j] == znak){
        x[i,j] <- NA
      }
    }
  }
}

x2<-na.omit(x)

for(k in 1:c){
  if(is.numeric(x2[,k])){
    meanX <- mean(x2[,k])
    for(l in 1:r){
      if(is.na(x[l,k])){
        x[l,k] <- meanX
      }
    }
  }else{
    modeX<-names(sort(-table(x2[,k])))[1]
    for(l in 1:r){
      if(is.na(x[l,k])){
        x[l,k] <- modeX
      }
    }
  }
}
return(x)
}

activityNAIns<-missing.insert(activityNA)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
#Total number of steps taken per day
TotNoStepsPerDayNAIns <- aggregate(steps ~ date, activityNAIns, sum)
#Create a histogram
color<-c('yellow','blue','red','purple','black','green')
hist(TotNoStepsPerDayNAIns$steps, breaks = 7, main = "Histogram of total number of steps per day", xlab = "Number of steps per day", ylab = "Frequency", col = color)
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
#Mean without NAs
mean(TotNoStepsPerDayNAIns$steps)
#Mean with NAs
mean(TotNoStepsPerDay$steps)
#Comparing
mean(TotNoStepsPerDayNAIns$steps)==mean(TotNoStepsPerDay$steps)
#Median without NAs
median(TotNoStepsPerDayNAIns$steps)
#Median with NAs
median(TotNoStepsPerDay$steps)
#Comparing
median(TotNoStepsPerDayNAIns$steps)==median(TotNoStepsPerDay$steps)
```

Conclusion:

The mean value didn't change. The data distribution is quite similar. As we fill the NAs by the mean value we observe the median value change. 


Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
#Day type insert
tmpLT <- as.POSIXlt(activityNAIns$date, format = "%Y-%m-%d")
tmpWD <- tmpLT$wday
tmpWD[tmpWD == 0] = 'weekend'
tmpWD[tmpWD == 6] = 'weekend'
tmpWD[tmpWD != 'weekend'] = 'weekday'
tmpWD <- factor(tmpWD, levels = c('weekend', 'weekday'))

activityNAIns$weekDay <- tmpWD
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
par(mfrow = c(2, 1))

for(type in c('weekend', 'weekday')){
  StepsInterval <- aggregate(steps ~ interval, activityNAIns, mean, subset=activityNAIns$weekDay==type )
  plot(StepsInterval, type = "l", main = type)
}
```