---
title: "Reproducible Research: Peer Assessment 1"
date: "4/9/2021"
output: html_document
---


## Introduction

This .Rmd file is written to fulfill the "Peer-graded Assignment: Reproducible Research Course Project 1". 

The file makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

This .Rmd file addresses five parts of the assignment:

1. Loading and preprocessing the data
2. What is mean total number of steps taken per day?
3. What is the average daily activity pattern?
4. Imputing missing values
5. Are there differences in activity patterns between weekdays and weekends?


## 1. Loading and preprocessing the data

####  Load the data

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations and three variables in this dataset.

The variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as \color{red}{\verb|NA|}NA)
* date: The date on which the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken


```{r readcsv, echo=TRUE} 
col_class = c("numeric", "Date", "numeric")
activity <- read.csv("activity.csv", head=TRUE, colClasses=col_class, na.strings="NA")

```

Check if data is correctly loaded. 

```{r checkdata, echo=TRUE} 
head(activity)
dim(activity)
summary(activity)
```

####  Preprocess the Data 

Create a New Variable, date_type, Indicating "weekday" or "weekend" for Each Day. 

```{r date_type, echo=TRUE} 
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
weekdays <- weekdays(activity$date)
table(weekdays)
date_type <-ifelse(weekdays=='Saturday' | weekdays=='Sunday', 'weekend','weekday')
activity <- cbind(activity,date_type)
table(activity$date_type)
str(activity)
```


## 2. What is mean total number of steps taken per day?

#### For this part of the assignment, we ignore the missing values in the dataset.


#### (1). Calculate the total number of steps taken per day

```{r Steps_Day, echo=TRUE}
Steps_Day <- aggregate(steps ~ date, activity, FUN=sum)
colnames(Steps_Day) <- c("Date", "Total_Steps_Per_Day")
```
Take a look at the aggregated data

```{r, echo=TRUE}
head(Steps_Day)
summary(Steps_Day)
```


#### (2). Make a histogram of the total number of steps taken each day


```{r histogram, echo=TRUE}

hist(x      = Steps_Day$Total_Steps_Per_Day, 
     xlab   = "Total Steps Per Day",
     ylab   = "Frequency",
     main   = "Histogram of the Total Number of Steps Taken Each Day",  
     col    = "lightblue", 
     ylim   = c(0,20), 
     breaks = 18)
```



#### (3). Calculate and report the mean and median of the total number of steps taken per day


```{r mean, echo=TRUE}
Steps_mean<-mean(Steps_Day$Total_Steps_Per_Day)
Steps_mean
```

#### Mean of Total Steps Per Day: `r Steps_mean`

```{r median, echo=TRUE}
Steps_median<-median(Steps_Day$Total_Steps_Per_Day)
Steps_median
```

#### Median of Total Steps Per Day: `r Steps_median`



## 3. What is the average daily activity pattern?

#### (1). Make a time series plot (i.e.type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

create a variable, Average_Steps_Per_Interval,  averaging number of steps taken across all days by interval

```{r Steps_interval, echo=TRUE}
Steps_interval <- aggregate(steps ~ interval, activity, FUN=mean)
colnames(Steps_interval) <- c("Interval", "Average_Steps_Per_Interval")
```
Take a look at the averaged data

```{r, echo=TRUE}
head(Steps_interval)
summary(Steps_interval)
```

#### Time Series Plot

```{r, echo=TRUE}
plot(   x  = Steps_interval$Interval, 
     xlab  = "5-minute Interval in Which Measurement Was Taken",
        y  = Steps_interval$Average_Steps_Per_Interval, 
     ylab  = "Average Steps across All Days", 
     main  = "Average Daily Activity Pattern", 
     type  = "l")
```


#### (2). Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}
Interval_maxstep <- Steps_interval[which.max(Steps_interval$Average_Steps_Per_Interval),1]
Interval_maxstep
maxstep<-Steps_interval[which.max(Steps_interval$Average_Steps_Per_Interval),2]
maxstep
```

#### The Interval, on average across all the days in the dataset, contains the maximum number of steps (i.e., maximum avarage steps = `r maxstep`), is Interval: `r Interval_maxstep`. 


## 4. Imputing missing values

#### Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

#### (1). Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA)

```{r, echo=TRUE}
na<-sum(is.na(activity$steps))
na
```

#### The total number of rows with NA is `r na`. 


#### (2). Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

My Strategy is to replace the 'NA's with the mean steps across all days for that 5-minute interval. 

#### (3). Create a new dataset that is equal to the original dataset but with the missing data filled in, where 'NA' being replaced with the mean steps across all days for that 5-minute interval.

```{r, echo=TRUE}
activity <- cbind(activity,Steps_interval)
activity$imputed_Steps <- ifelse(is.na(activity$steps), activity$Average_Steps_Per_Interval, activity$steps)
activity_new <- data.frame(steps=activity$imputed_Steps, interval=activity$interval, date=activity$date)
head(activity_new)
summary(activity_new)
```


#### (4). Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

calculate the total number of imputed steps taken each day

```{r, echo=TRUE}
Imputedstep_Day <- aggregate(steps ~ date, activity_new, FUN=sum)
colnames(Imputedstep_Day) <- c("Date", "Imputed_Total_Steps_Per_Day")
```
Take a look at the aggregated data

```{r, echo=TRUE}
head(Imputedstep_Day)
summary(Imputedstep_Day)
```


#### histogram of the total number of imputed steps taken each day


```{r , echo=TRUE}

hist(x      = Imputedstep_Day$Imputed_Total_Steps_Per_Day, 
     xlab   = "Total Imputed Steps Per Day",
     ylab   = "Frequency",
     main   = "Histogram of the Total Number of Imputed Steps Taken Each Day",  
     col    = "pink", 
     ylim   = c(0,20), 
     breaks = 18)
```



####  Calculate and report the mean and median of the total number of imputed steps taken per day


```{r , echo=TRUE}
imputedSteps_mean<-mean(Imputedstep_Day$Imputed_Total_Steps_Per_Day)
imputedSteps_mean
```

#### Mean of Total Imputed Steps Per Day: `r imputedSteps_mean`

```{r , echo=TRUE}
imputedSteps_median<-median(Imputedstep_Day$Imputed_Total_Steps_Per_Day)
imputedSteps_median
```

#### Median of Total Imputed Steps Per Day: `r imputedSteps_median`




#### Summary: After missing data imputation, the mean estimates stay the same as the mean in the first part of the assignment, while the median estimats are slight different. The median of the total imputed steps per day is  `r imputedSteps_median`, and the median of the total steps per day in the first part of the assignment is  `r Steps_median`. The impact of imputing missing data on the estimates of the total daily number of steps is minor. The missing data imputing strategy has a slight influence on the median, but not mean. 




## 5. Are there differences in activity patterns between weekdays and weekends?

#### (1). Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

#### The new factor variable has already been created in the data preprocess step. The variable name is date_type. 
```{r , echo=TRUE}
table(activity$date_type)
```


#### (2)  Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r, echo=TRUE}
Steps_interval_1 <- aggregate(steps ~ interval+ date_type, activity, FUN=mean)
colnames(Steps_interval_1) <- c("Interval", "date_type", "Average_Steps_Per_Interval")
```
####  Check aggregated data 
```{r, echo=TRUE}
summary(Steps_interval_1)
table(Steps_interval_1$date_type)
library(psych)    
describeBy(Steps_interval_1$Average_Steps_Per_Interval, Steps_interval_1$date_type)
```

####  Make a Panel Plot
```{r, echo=TRUE}
library(lattice)
lattice.options(default.theme = standard.theme(color = FALSE))


xyplot( Average_Steps_Per_Interval ~ Interval | date_type,
        data  = Steps_interval_1,
        xlab  = "5-minute Interval in Which Measurement Was Taken",
        ylab  = "Average Steps across All Days", 
        type  = "l",
        auto.key = TRUE,
        layout = c(1, 2)
        )

```



#### Summary: 
#### It seems people tend to excise around the morning time during the weekday, while tend to excise late during the weekend.  