# Reproducible Research: Peer Assessment 1

Step count data from a monitoring device collected at 5 minute intervals 
throughout the day are examined. The data are from one individual 
during the months of October and November 2012.

## Loading and preprocessing the data

First some setup is necessary. The ggplot2 library is loaded. The working
directory is set and the workspace is cleared.

```{r setoptions,echo=TRUE}
opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
sessionInfo()
setwd("~/studyJH-DataSci/5-repData/RepData_PeerAssessment1")
rm(list = ls())
```

Next the data file is loaded and some date-time vectors are created 
for graphing proposes. Also a factor vector is created to identify 
entries as following on weekdays or weekends.

```{r}
stp <- read.csv("activity.csv", header = TRUE, na.strings = "NA" )
t9HM <- as.character(90000 + as.numeric(stp$interval))
datetime <- strptime(paste(stp$date, t9HM), "%Y-%m-%d 9%H%M")
steptime <- strptime(t9HM[1:288], "9%H%M")
wd <- weekdays(datetime)
wd <- ifelse((wd == "Saturday" | wd == "Sunday"), "weekend", "weekday")
stp <- cbind(stp, datetime, wd)

```

## What is mean total number of steps taken per day?

Executing the following code.

```{r}
dsum <- aggregate(stp$steps, by=list(stp$date), FUN=sum, na.rm=TRUE)
names(dsum) <- c("Day", "cntStep")
summary(dsum$cntStep)
```

We see the step count mean is `r mean(dsum$cntStep)` and 
the median is `r median(dsum$cntStep)`. 
The histogram of the step count by days is here.


```{r plot1, fig.width=7, fig.height=6}
hist(dsum$cntStep, breaks = seq(0,22000,1000), col = "red", 
     main = "Number of Steps Taken Each Day", 
     xlab = "Steps per Day")
```

## What is the average daily activity pattern?

```{r}
dstep <- aggregate(stp$steps, by=list(factor(stp$interval)), 
                   FUN=mean, na.rm=TRUE)
names(dstep) <- c("int5min", "meanStep")
summary(dstep)
```

Here is a time series plot of the 5-minute interval (x-axis) and the average 
number of steps taken, averaged across all days (y-axis).

```{r plot2, fig.width=7, fig.height=6}
plot(steptime, dstep$meanStep, type = "l",
     main = "Average Number of Steps Taken by 5-minute interval", 
     xlab = "5-minute intervals", 
     ylab = "Step Count")
```

The maximum step count averaged over time of day intervals is 
`r max(dstep$meanStep)` and occures in the 
`r dstep$int5min[dstep$meanStep == max(dstep$meanStep)]` interval.

## Imputing missing values

The total number of missing values in the dataset (i.e. the total number of 
rows with NAs) is `r sum(is.na(stp$steps))`. This correspones to to eight day 
on which the recording device was probabily not on and recording.

```{r}
sum(is.na(stp$steps))
```

A new dataset that is equal to the original dataset but with the missing 
data filled in by using the mean for that 5-minute interval is created.

```{r}
steps2 <- stp$steps
steps_mean <- rep(dstep$meanStep, 61)
steps2[is.na(steps2)] <- steps_mean[is.na(steps2)]
```

The summary statistics for this imputed data set are as follows:

```{r}
dsum2 <- aggregate(steps2, by=list(stp$date), FUN=sum)
dstep2 <- aggregate(steps2, by=list(factor(stp$interval)), FUN=mean)
names(dsum2) <- c("Day", "cntStep")
summary(dsum2$cntStep)
```

We see the step count mean is `r mean(dsum2$cntStep)` and 
the median is `r median(dsum2$cntStep)`. 
The histogram of the step count by days for the imputed data is here. 

```{r plot3, fig.width=7, fig.height=6}
hist(dsum2$cntStep, breaks = seq(0,22000,1000), col = "blue", 
     main = "Number of Steps Taken Each Day (imputed data)", 
     xlab = "Steps per Day")
```

The time series plot for the imputed data of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis) 
is here. 

```{r plot4, fig.width=7, fig.height=6}
plot(steptime, dstep2$x, type = "l",
     main = "Average Number of Steps Taken (imputed data)", 
     xlab = "5-minute intervals", 
     ylab = "Step Count")
```

## Are there differences in activity patterns between weekdays and weekends?

The imputed data is used to look at activity patterns between weekdays and 
weekends.

```{r}
dstep3 <- aggregate(steps2, by=list(factor(stp$interval), stp$wd), FUN=mean)
names(dstep3) <- c("Interval", "wd", "Number_of_steps")
```

```{r plot5, fig.width=7, fig.height=6}
p <- ggplot(dstep3, aes(x=Interval, y=Number_of_steps, group=wd, main="Title")) + geom_line() 
p <- p + ggtitle("Comparision of weekday and weekend Activity")
p <- p + facet_grid(wd ~ .)
p <- p + scale_x_discrete(breaks = dstep3$Interval[seq(1,288, by = 36)])
print(p)

```

While the peak activity is not as high on weekends as weekdays, activity 
is more substained thoughout the day on weekends.

