---
title: "Project 1"
author: "Husam"
date: "July 7, 2018"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
```

# Assignment 1 Markdown

Loading data: 

```{r}
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
                     destfile = "AMD.zip")
unzip("AMD.zip")
activity <- read_csv("activity.csv", col_types = cols(date = col_date(format = "%Y - %m - %d"), interval = col_double(), steps = col_double()))
```

## What is mean total number of steps taken per day?
### 1. Calculate the total number of steps taken per day

```{r}
activity.date <- activity %>%
  group_by(date) %>%
  summarise(total.steps = sum(steps, na.rm = TRUE))
```

###2.Make histogram of total number of steps taken each day
```{r}
hist(activity.date$total.steps, xlab = "Total Steps", main = "Frequency of Total Steps by Day", col = "purple")
```

###3. Calculate and report mean and median of total number of steps taken per day 
```{r}
activity.date <- activity %>%
  group_by(date) %>%
  summarise (total.steps = sum(steps, na.rm = TRUE), mean.steps = mean(steps, na.rm = TRUE), median.steps = median(steps, na.rm = TRUE))
View(activity.date)
```

## What is the average daily activity pattern 
### 1. Make time series plot of 5 minute interval and average number of steps taken averaged across all days
```{r}
activity.interval <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)
activity.interval <- setNames(activity.interval, c("interval", "mean.steps"))
plot(activity.interval$interval, activity.interval$mean.steps, type = "l", xlab = "Interval", ylab = "Mean Steps")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
activity.interval[which.max(activity.interval$mean.steps),]$interval
```

##Imputing Missing Values 
### 1. Calculate and report the total number of missing values in the dataset 
```{r}
sum(is.na(activity$steps))
```

### 2. Devise a strategy for filling in all of the missing values in the dataset.
```{r}
interval.mean.steps <- activity.interval$mean[match(activity$interval, activity.interval$interval)]
```

### 3. Create a new dataset that is equal to the original dataset but with missing values filled 
```{r}
activity.filled <- transform(activity, steps = ifelse(is.na(activity$steps), yes = interval.mean.steps, no = activity$steps))
```

### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day
```{r}
activity.filled.date <- activity.filled %>%
  group_by(date) %>%
  summarise(total.steps = sum(steps, na.rm = TRUE))

hist(activity.filled.date$total.steps, xlab = "Total Steps", main = "Frequency of Total Steps by Day", col = "blue")

activity.filled.date <- activity.filled %>%
  group_by(date) %>%
  summarise (total.steps = sum(steps, na.rm = TRUE), mean.steps = mean(steps, na.rm = TRUE), median.steps = median(steps, na.rm = TRUE))
View(activity.filled.date)
```

They are different.

## Are there differences in activity patterns between weekdays and weekends? 
```{r}
activity.filled <- mutate(activity.filled, "weekday" = weekdays(activity.filled$date))

activity.filled$datetype <- sapply(activity.filled$date, function(x) {
  if (weekdays(x) == "Saturday"|weekdays(x) == "Sunday") 
    y <- "Weekend"
    else 
    y <- "Weekday"
  y
})

activity.interval.filled <- aggregate(steps~interval + datetype, activity.filled, FUN=mean, na.rm=TRUE)

library(ggplot2)
g <- ggplot(activity.interval.filled, aes(x = interval, y = steps))
g + geom_line() + facet_wrap(~datetype)
```


