---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
For this example, it is assumed that  ***activity.zip*** or  ***activity.csv***  is located in the working directory. In case that the CSV file wasn´t unzipped, the code will load it from the zip file. 

One additional improvement for this code would be to include a logic statement to check if  the **activity.zip** file is in the directory, if not the code could download it from the repository.

```{r}

if(file.exists("activity.csv")){
    data <- read.csv("activity.csv")
} else {
    data <- read.csv(unz("activity.zip", "activity.csv"))
}

# Remove the NA values
dataClean <- data[!(is.na(data$steps)),]

```

## What is mean total number of steps taken per day?
Load the libraries needed for the project.

``` {r , warning=FALSE ,results='hide'}
library(dplyr)

```

Now there are two datasets, one with the original data and the other one skipping missing values. For this case the dplyr library is used to group the data by day. 

``` {r}
a <- dataClean %>% group_by(date) %>% summarise(TotalSum = sum(steps), Mean = mean(steps), Median = median(steps))
barplot(a$TotalSum , names.arg = a$date , xlab = "Day" , ylab = "Total number of steps per day")

MeanStepsPerDay <- mean(a$TotalSum , na.rm = TRUE)
MedianStepsPerDay <- median(a$TotalSum , na.rm = TRUE)

```

The mean number of steps per day are `r format(MeanStepsPerDay , digits=10)` and the Median `r format(MedianStepsPerDay, digits=10)`.

## What is the average daily activity pattern?
As first step, the clean data will be summarize by interval calculating the mean, total sum and the median for each interval for all the days. 

Afterwards the time series for the average step for all days in each interval is plotted. There is a clear trend with initial interval with low number of steps then around interval 500 minutes, the average number steps increases until the maximum values.

The code that generates the graph is : 

``` {r}

b <- dataClean %>% group_by(interval) %>% summarise(TotalSum = sum(steps), Mean = mean(steps), Median = median(steps))

plot(b$interval, b$Mean , type = "l", xlab = "Interval" , ylab = "Average Steps per day")
Interval <- b$interval[grep(max(b$Mean),b$Mean)]
ValueMax <- max(b$Mean)

```

The interval with higher average steps in the day is `r Interval` with an average value of `r ValueMax` steps in that interval.

## Imputing missing values

- The calculation of the total number of missing value can be calculated in one line of code using the ***is.na()*** function to check if the value ***NA***
``` {r}

 nrow(data[(is.na(data$steps)),]) 

```

- Replace the missing values with the average number of steps for all days for the giving interval. Lets look the existing data from the first 5 row 

``` {r}
head(data ,5 ) # See the errors in the first 5 line.
head(b , 5) # mean values for the first 5 intervals
```

- From the previous question, there is a dataset with the mean steps values per interval, therefore it is possible to look for the NA values in steps and replace those values with the *appropiate mean value of steps for all days in the define interval that appear in the data entry.* For this purpose, the functions ***mutate()*** , ***ifelse*** and ***rownumber()*** are used : 


``` {r}
DataReplace <- mutate(data, steps = ifelse(is.na(steps), b$Mean[which(b[,1] == data$interval[row_number()])], steps)) # mutate goes row by row, replace by the appropiate value in case of step value is NA

nrow(data[(is.na(DataReplace$steps)),])  ## Verify there is no missing values 

head(DataReplace, 5 )

```

Now that the missing values has been filled, it is possible to compute the mean and median number of steps per day and compare them with the values calculated in the other section. 

``` {r}

a2 <- DataReplace %>% group_by(date) %>% summarise(TotalSum = sum(steps), Mean = mean(steps), Median = median(steps))
barplot(a2$TotalSum , names.arg = a2$date , xlab = "Day" , ylab = "Total number of steps per day")

MeanStepsPerDay2 <- mean(a2$TotalSum )
MedianStepsPerDay2 <- median(a2$TotalSum )


```

 -  Mean `r format(MeanStepsPerDay , digits=10)` and Median `r format(MedianStepsPerDay, digits=10)` *** without*** the missing values
 - Mean `r format(MeanStepsPerDay2, digits=10)` and Median `r format(MedianStepsPerDay2,digits=10)` *** replacing*** the missing values



## Are there differences in activity patterns between weekdays and weekends?
Create a new function to determine if a date is a weekday or a weekend and then vectorize it.

``` {r}
isweekend <- function(date) {
    # use a date that are Saturday and Sunday. Weekday() give different the day name in different languages depending on the computer settings in my case is Spanish
        if (weekdays(date) == weekdays(as.Date("2016-01-09")) | weekdays(date) == weekdays(as.Date("2016-01-10")))
        return("weekend")
    else
        return("weekday")
}

visweekend <- Vectorize(isweekend, "date") 
```
Now mutate the new data set and add a new factor variable to it

``` {r}
newdata <- mutate(dataClean, DayType = visweekend(as.Date(date)))
newdata$DayType <- as.factor(newdata$DayType)
```
Aggregate the data by interval for weekdays and weekends.
``` {r}
newdailyav <- aggregate(steps ~ interval + DayType, newdata, mean)
```

Use lattice to plot the data:
``` {r}
library(lattice)
xyplot(steps ~ interval | DayType, newdailyav, type = "l", layout = c(1,2),
    main = "Average number of steps taken by interval for weekdays and Weekends",
    xlab = "Interval (time of day)", ylab = "Steps" )
```

