---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Reproducible Research: 
## Assessment 1
### ========================================================

###  1.	Code for reading in the dataset and/or processing the data

```{r p1}

library(knitr);
library(ggplot2)
library(dplyr)
library(ggthemes)

setwd("C:/Users/KRAUSE/Documents/GitHub/RepData_PeerAssessment1")
activity <- read.csv("activity.csv")

#Exploring the basics of this data

dim(activity)
names(activity)
head(activity)
str(activity)
#total number of missing data
sum(is.na(activity$steps))/dim(activity)[[1]]
#transforming the date column into date format
activity = activity %>% 
  mutate(date = as.Date(date))
```
### 2.	Histogram of the total number of steps taken each day

```{r p2, warning=TRUE, echo = TRUE}


# total number of steps taken per day
StepsTotal <- aggregate(steps ~ date, data=activity, sum, na.rm=TRUE)

summary(StepsTotal$steps)

ggplot(data=StepsTotal, aes(steps)) + 
  geom_histogram(breaks=seq(0, 20000, by=500),col="darkgreen", 
                 fill="green", 
                 alpha = .2) + 
  labs(title="Daily Number of Steps", x="Steps Per Day", y="Number of Days") + 
  xlim(c(0,22000)) + 
  ylim(c(0,8))
```

### 3.	Mean and median number of steps taken each day


```{r p3, warning=TRUE, echo = TRUE}


# mean
stmean <- round(mean(StepsTotal$steps),0)

# median
stmedian <- median(StepsTotal$steps)

stmean <- sprintf("%6.0f", stmean)

```
Mean  number of steps taken each day is `r stmean`

Median number of steps taken each day is `r stmedian`


### 4.	Time series plot of the average number of steps taken
```{r p4, warning=TRUE, echo = TRUE}

avg_steps_per_day <- aggregate(steps ~ date, activity, mean)

ggplot(avg_steps_per_day, aes(x = date, y = steps))+ geom_line( color="blue", size=0.1) +
  theme_hc () +
  scale_colour_manual("",values="blue")+
  scale_fill_manual("",values="grey12")+  
  labs(title="Average number of steps", y="Steps Per Day", x="Date") + 
  theme(axis.text.x=element_text(angle=0, hjust=1), panel.background = element_rect(fill = NA)) 



```
### 5.	The 5-minute interval that, on average, contains the maximum number of steps
```{r p5, warning=TRUE, echo = TRUE}

avg_steps_per_interval <- aggregate(steps ~ interval, activity, mean)

interval_idx <- which.max(avg_steps_per_interval$steps)



```
### 6.	Code to describe and show a strategy for imputing missing data
```{r p6, warning=TRUE, echo = TRUE}



# Loop thru all the rows of activity, find the one with NA for steps.
# For each identify the interval for that row
# Then identify the avg steps for that interval in avg_steps_per_interval
# Substitute the NA value with that value

avg_steps_per_interval <- aggregate(steps ~ interval, activity, mean)
activity_im <- activity
for (i in 1:nrow(activity_im)) {
  if(is.na(activity_im$steps[i])) {
    val <- lapply(avg_steps_per_interval$steps[which(avg_steps_per_interval$interval == activity_im$interval[i])], round,0)
    activity_im$steps[i] <- as.numeric(val)
  }
}


```
### 7.	Histogram of the total number of steps taken each day after missing values are imputed
```{r p7, warning=TRUE, echo = TRUE}



# Aggregate the steps per day with the imputed values
steps_per_day_im <- aggregate(steps ~ date, activity_im, sum)

# Draw a histogram of the value 
ggplot(data=steps_per_day_im, aes(steps)) + 
  geom_histogram(bins  = 50, col="darkgreen", 
                 fill="green", 
                 alpha = .2) + 
  labs(title="Daily Number of Steps", x="Steps Per Day", y="Number of Days") 



```
### 8.	Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r p8, warning=TRUE, echo = TRUE}
# 

#Cretae a function to determine if the date is a weekday
week_day <- function(date_val) {
  wd <- weekdays(as.Date(date_val, '%Y-%m-%d'))
  if  (!(wd == 'sábado' || wd == 'domingo')) {
    x <- 'Weekday'
  } else {
    x <- 'Weekend'
  }
  x
}

# Apply the week_day function and add a new column to activity dataset
activity$day_type <- as.factor(sapply(activity$date, week_day))



# Create the aggregated data frame by intervals and day_type
steps_per_day_impute <- aggregate(steps ~ interval+day_type, activity, mean)

# Create the plot
ggplot(steps_per_day_impute, aes(interval, steps)) +
  geom_line(stat = "identity", aes(colour = day_type)) +
  theme_gray() +
  facet_grid(day_type ~ ., scales="fixed", space="fixed") +
  labs(x="Interval", y=expression("No of Steps")) +
  ggtitle("Steps Per Interval by day type")


```
