# Reproducible Research (Peer Assessment 1)  
**Data Science -> Reproducible Research -> Peer Assessment 1**   
   
   
-------------------  

### Welcome  

Hi,  
Welcome to my first Course Project / Peer Assessment Assignment.  
The purpose of this work is to produce a R Markdown document with all the code needed to answer the required questions.  
  
This code was prepared in the following environment:
 - OS: Linux x86_64 (3.12.15-pclos1)
 - R/RStudio: 3.0.2 / 99.9.9, with the following packages:
   - tools (3.0.2)
   - knitr (1.6)
  
I'll start by preparing the environment to load required packages and to always display the code chunks in the html output file.  

```{r Setup}
library(tools)
library(knitr)
opts_chunk$set(echo=TRUE, results="markup")

```
   
-------------------
### Loading and preprocessing the data

I will use two separate code chunks.  
  
The first code chunk is responsible to make sure the right data file is in the directory and then it loads it the `dataset` variable.  
If something goes wrong during execution an error should be rised. In case no handled error ocurrs a message is returned to let user know everything is fine.

```{r Loading}

kRemote <- "https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip"
kFile   <- "activity.zip"
kFileDt <- "activity.csv"
kFileMd <- "activity.md5"
    
if (!file.exists(kFile)){
    # File is not present at the working directory. Let's download it!
    method <-"auto"  # Default method: shall be fine for MS Windows (untested!)
    if (.Platform$OS.type=="unix") method <- "curl"  # Use for unix-like systems
    download.file(kRemote, kFile, method, FALSE, "wb")
    if (!file.exists(kFile))
        stop("Download failed!")
} else {
    # File is now present (it was before or it was sucessfully downloaded).
    # Let's check if it is the expected file. I'll compare it's MD5 hash with
    # a precomputed one.
    if (readLines(kFileMd)!= as.vector(md5sum(kFile))){
        stop("File is not correct!")
    } else {
        # The correct file shall be present. I'll read the data directly from 
        # inside the Zip file.
        myConn <- unz(kFile, kFileDt)
        dataset <- read.csv(myConn, na.strings="NA", stringsAsFactors=FALSE)
        print("Data sucessfully loaded!")
    }
}
```

``` {r Preprocessing}
if (!exists("dataset")) {
    stop("It looks something went wrong in the 'Loading' chunk!")
} else {
    
}

```
-------------------
## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
