---
title: "Course Project 1"
author: "Antón Blanco"
date: "5 de abril de 2016"
output: html_document
---

About
---------------------
This was the first project for the *Reproducible Research* course in Coursera's Data Science specialization track. The purpose of the project was to answer a series of questions using data collected from a FitBit.

Synopsis
---------------------
This project has 3 aims:

1.loading and preprocessing data.

2.imputing missing values.

3.interpreting data to answer research questions.


Data
---------

The variables included in this dataset are:

Steps: Number of steps taking in a 5-minute interval (missing values are coded as NA).

Date: The date on which the measurement was taken in YYYY-MM-DD format.

Interval: Identifier for the 5-minute interval in which measurement was taken.

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

The process of this project is what it follows:

Loading and preprocessing the data
------------
Download, unzip and load data into data frame
```{r}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
        temp <- tempfile()
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
        unzip(temp)
        unlink(temp)
}

data <- read.csv("activity.csv")
```



```{r}
stepsday <- aggregate(steps ~ date, data, sum)
hist(stepsday$steps, main = paste("Total Steps every Day"), col="green", xlab="Steps")
```


```{r}
dmean <- mean(stepsday$steps)
dmedian <- median(stepsday$steps)
```
The **mean** is `r dmean` and the **median** is `r dmedian`.

What is mean total number of steps taken per day?
----------
+Calculate average steps for each interval for all days.

+Plot the Average Number Steps per Day by Interval.

+Find interval with most average steps.

```{r}
stepsinterval <- aggregate(steps ~ interval, data, mean)
plot(stepsinterval$interval,stepsinterval$steps, type="l", xlab="Interval", ylab="Steps",main="Average Number of Steps per Day by Interval")

```

```{r}
max_interval <- stepsinterval[which.max(stepsinterval$steps),1]
```

The 5-minute interval, on average across all the days in the data set, containing the maximum number of steps is `r max_interval`.

Impute missing values. Compare imputed to non-imputed data
----------------


```{r}
incompletecases<-sum(!complete.cases(data$steps))
```
The number of cases with *NA* is `r incompletecases`.
```{r}
imputed_data <- transform(data, steps = ifelse(is.na(data$steps), stepsinterval$steps[match(data$interval, stepsinterval$interval)], data$steps))

imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0
```
Recount total steps by day and create Histogram.
```{r}
stepsday_i <- aggregate(steps ~ date, imputed_data, sum)
hist(stepsday_i$steps, main = paste("Total Steps Every Day"), col="blue", xlab="Steps")

#Create Histogram to show difference. 
hist(stepsday$steps, main = paste("Total Steps Every Day"), col="red", xlab="Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)
```
Calculate new mean and median for imputed data.
```{r}
dmean.i <- mean(stepsday_i$steps)
dmedian.i <- median(stepsday_i$steps)


```
Calculate difference between imputed and non-imputed data.
```{r}
mean_diff <- dmean.i - dmean
med_diff <- dmedian.i - dmedian
```
Calculate total difference.
```{r}
total_diff <- sum(stepsday_i$steps) - sum(stepsday$steps)
```

+The imputed data mean is `r dmean.i`.

+The imputed data median is `r dmedian.i`.

+The difference between the non-imputed mean and imputed mean is `r mean_diff`.

+The difference between the non-imputed mean and imputed mean is `r med_diff`.

+The difference between total number of steps between imputed and non-imputed data is `r total_diff`. According to this, there were `r total_diff` more steps in the imputed data.

Are there differences in activity patterns between weekdays and weekends?
--------
Created a plot to compare and contrast number of steps between the week and weekend. There is a higher peak earlier on weekdays, and more overall activity on weekends.

```{r}
library(lattice)
library(reshape2)
data[,2]<-as.Date(data[,2])
data[,"days"]<-weekdays(data[,2])
for (i in 1:nrow(data)) {
  if (data$days[i] == "sábado" | data$days[i]  == "domingo") {
    data$dayOfWeek[i] = "weekend"
  } else {
    data$dayOfWeek[i] = "weekday"
  }
}
```
```{r}
w <- tapply(data$steps,list(data$interval,data$dayOfWeek),mean,na.rm=T)

w <- melt(w)

colnames(w) <- c("interval","day","steps")

library(lattice)
library(reshape2)
xyplot(w$steps ~ w$interval | w$day, layout=c(1,2),type="l",main="Time Series Plot of the Average of Total Steps (weekday vs. weekend)",xlab="Time intervals (in minutes)",ylab="Average of Total Steps")
```