---
title: "Reproducible Research: Peer Assessment 1"
author: Alejandro
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load the necessary R packages.
```{r loadPackages}
library(dplyr)
library(knitr)
library(ggplot2)
```

### 1. Load data from zipfile.
```{r loadData}
zip_file = unzip("activity.zip")
dat = read.csv(zip_file)
```

### 2. Transfrom data for analysis.
```{r transformData}
dat = transform(dat, date = as.Date(date))
```

## What is mean total number of steps taken per day?
Compute total and average numbers of steps per day using *group_by()* and *summarize()* functions from *dplyr* package.
```{r meanStepsByDayTable, results='asis'}
dat2 = group_by(dat, date)
steps.by.day = summarise(dat2, 
                              total.steps = sum(steps) / 1000,
                                mean.steps = mean(steps, na.rm = TRUE),
                                median.steps = median(steps, na.rm = TRUE))
```
### 1. Make a histogram of the total number of steps by day
```{r totalStepsHistogram}
with(steps.by.day, {
        hist(total.steps, breaks = 30, col = "light blue",
                        border = "white", 
                        main = "Figure 1: Histogram of Total Steps per Day",
                        xlab = "Number of Steps (in thousands)")
        abline(v = mean(total.steps, na.rm = T), col = "blue", lwd = 2,
               lty = 4)
        text(16, 8, 
             labels = paste("Mean Total Steps =",
                                   mean(total.steps, na.rm = T) %>%
                                           round(1),
                                   "(in thousands)"),
             cex = 0.8)
})
```

### 2. Create a table to report the mean, median, and total number of steps taken per day
```{r stepsTable, results='asis'}
steps.by.day %>% 
        kable(digits = 2, caption = "Table 1. Steps taken per day") %>%
        print
```

## What is the average daily activity pattern?
### 1. Time series plot of daily average per 5-min intervals
```{r averageStepsPerInterval}
dat3 = group_by(dat, interval)
steps.by.interval = summarise(dat3, mean.steps = mean(steps, na.rm = TRUE))
with(steps.by.interval, {
        plot(y = mean.steps, x = dat3$interval %>% unique,
             type = "l", col = "light blue", lwd = 2,
             main = "Figure 2: Average Number of Steps per Interval",
             xlab = "Interval",
             ylab = "Average steps taken")
        abline(v = dat3$interval[which.max(mean.steps)], 
               col = "blue", lwd = 2, lty = 4)
        text(1400, 170, 
             labels = paste("Interval =",
                                   dat3$interval[which.max(mean.steps)],
                            ", Max Mean Steps =", max(mean.steps) %>%
                                    round(1)),
             cex = 0.8)
})
```

### 2. Maximum 5-min mean steps interval
```{r maxMeanStepsInterval}
cat("Interval =",
        dat3$interval[which.max(steps.by.interval$mean.steps)])
cat("Max mean steps =",
        max(steps.by.interval$mean.steps))
```

## Imputing missing values

### 1. Total missing values in data set
```{r missingTotal}
apply(dat, 2, function(x) is.na(x) %>% sum) 
```

### 2. Filling in missing values using 5-min daily average
```{r fillingInMissing}
dat4 = merge(dat, steps.by.interval, by = "interval")
head(dat4)
```

### 3. Create new dataset equal to original dataset with missing data filled in
```{r newDataNonMissing}
dat4 = transmute(dat4, steps = ifelse(is.na(steps), mean.steps, steps),
                 date = date, interval = interval)
dat4 = dat4[order(dat4$date),]
head(dat4)
apply(dat4, 2, function(x) is.na(x) %>% sum)
```

### 4. Histogram of total number of steps
```{r totalStepsHistogramNonMissing}
dat5 = group_by(dat4, date)
steps.by.day = summarise(dat5, 
                              total.steps = sum(steps) / 1000,
                                mean.steps = mean(steps, na.rm = TRUE),
                                median.steps = median(steps, na.rm = TRUE))
with(steps.by.day, {
        hist(total.steps, breaks = 30, col = "light blue",
                        border = "white", 
                        main = "Figure 3: Histogram of Total Steps per Day",
                        xlab = "Number of Steps (in thousands)")
        abline(v = mean(total.steps, na.rm = T), col = "blue", lwd = 2,
               lty = 4)
        text(16, 8, 
             labels = paste("Mean Total Steps =",
                                   mean(total.steps, na.rm = T) %>%
                                           round(1),
                                   "(in thousands)"),
             cex = 0.8)
})
```

Mean and median number of steps taken per day
```{r stepsTableNonMissing, results = "asis"}
steps.by.day %>% 
        kable(digits = 2, caption = "Table 2. Steps taken per day") %>%
        print
```

#### Comparing new and original datasets
Values from the new dataset (with no missing values) differ from the original (with missing values) in that the variability has been reduced.

## Are there differences in activity patterns between weekdays and weekends?
### 1. Create a new factor variable indicating week and weekdays
```{r weekends}
dat6 = mutate(dat4, weekday = ifelse(weekdays(date) == "Saturday" |
                              weekdays(date) == "Sunday", 
                              "weekend",
                              "weekday"))
```

### 2. Panel plot of daily average 5-min interval by weekday
```{r weekendsPlot}
dat7 = group_by(dat6, weekday, interval)
steps.by.interval = summarise(dat7, mean.steps = mean(steps, na.rm = TRUE))
p = ggplot(data = steps.by.interval,
           aes(y = mean.steps, 
               x = interval,
               group = weekday, color = weekday)
)
p + geom_line(color = "light blue", lwd = 1.25) + 
        facet_grid(. ~ weekday) + theme_bw() +
        ggtitle("Figure 4: Average Number of Steps per Interval") +
        labs(x = "Interval",
             y = "Average steps taken")
```
