---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r preprocessing}
# Load the csv data from the zip file into a data frame
step_data <- read.table(unz("activity.zip", "activity.csv"), header=T, quote="\"", sep=",")
# ensure that the date column is a date data type
step_data$date <- strptime(step_data$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?
```{r mean total steps}
library(plyr)
complete_steps <- step_data[complete.cases(step_data$steps),]
totaldailysteps <- ddply(complete_steps, .(date), summarise, total = max(steps))
hist(totaldailysteps$total, main = "Total number of steps each day", xlab = "Number of Steps")

meansteps <- mean(totaldailysteps$total,na.rm = TRUE)
mediansteps <- median(totaldailysteps$total,na.rm = TRUE)
```
The mean number of total daily steps is `r meansteps`. The median number of total daily steps is `r mediansteps`.

## What is the average daily activity pattern?

```{r avg by interval}
avgsteps <- ddply(complete_steps, .(interval), summarise, average_steps = mean(steps))
plot(x = avgsteps$interval, y = avgsteps$average_steps, type = "l", main = "Average Daily Activity Pattern", xlab = "Daily Interval", ylab = "Average Number of steps")
maxinterval <- avgsteps[which.max(avgsteps$average_steps),]$interval
```
The daily 5-minute interval containing the max number of steps on average is `r maxinterval`.

## Imputing missing values

```{r missing vals}
missingrows <- sum(is.na(step_data$steps))
```
The Total number of missing values (coded as NA) is `r missingrows`.

```{r impute missing vals}
#Create a copy of the step_data dataset which will be used to fill in missing values
step_dataimputed <- step_data
#Impute the missing values by using the average steps by interval
step_dataimputed$steps[is.na(step_dataimputed$steps)] <- ave(step_dataimputed$steps, step_dataimputed$interval, FUN = function(x)mean(x, na.rm=TRUE))[c(which(is.na(step_dataimputed$steps)))]
```

```{r mean total steps imputed}
library(plyr)
totaldailysteps_imputed <- ddply(step_dataimputed, .(date), summarise, total = max(steps))
hist(totaldailysteps_imputed$total, main = "Total number of steps each day", xlab = "Number of Steps")

# calculate the mean with the imputed values
meansteps_imputed <- mean(totaldailysteps_imputed$total,na.rm = TRUE)
# calculate the difference between this mean and the one calcualted above with missing values ignored
meansteps_imputed_diff <- meansteps_imputed - meansteps

if (meansteps_imputed_diff == 0) {
  meanresult <- "no"
} else if (meansteps_imputed_diff < 0) {
  meanresult <- "a negative effect"
}else 
  meanresult <- "a positive effect"

# calculate the median with the imputed values
mediansteps_imputed <- median(totaldailysteps_imputed$total,na.rm = TRUE)
# calculate the difference between this median and the one calcualted above with missing values ignored
mediansteps_imputed_diff <- mediansteps_imputed - mediansteps

if (mediansteps_imputed_diff == 0) {
  medianresult <- "no"
} else if (mediansteps_imputed_diff < 0) {
  medianresult <- "a negative effect"
}else 
  medianresult <- "a positive effect"
```
The mean number of total daily steps (with missing values imputed) is `r meansteps_imputed`. The median number of total daily steps (with missing values imputed) is `r mediansteps_imputed`. This is a difference of `r meansteps_imputed_diff` for the mean and `r mediansteps_imputed_diff` for the median from the calculations with missing values ignored.

Imputing the missing data has `r meanresult` on the mean and a `r medianresult` on the median.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekend diffs}
step_dataimputed$weekday <- weekdays(step_dataimputed$date,abbreviate = TRUE)
step_dataimputed$DayType <- "weekday"
step_dataimputed$DayType[step_dataimputed$weekday == "Sat" | step_dataimputed$weekday == "Sun"] <- "weekend"
step_dataimputed$DayType <- factor(step_dataimputed$DayType)

library(lattice)
avgsteps_imputed <- ddply(step_dataimputed, .(interval, DayType), summarise, average_steps = mean(steps))
xyplot(average_steps~interval | as.factor(DayType), data = avgsteps_imputed, layout = c(1,2), type="l", xlab = "Interval", ylab = "Number of steps" )
```