---
title: "Reproducible Research: Peer Assessment 1"

    keep_md: true
---


## Loading and preprocessing the data

```{r}
unzip("activity.zip")
mydata <- read.csv("activity.csv")
```
## What is mean total number of steps taken per day?
###Make a histogram of the total number of steps taken each day
```{r}
totalsteps <- aggregate(steps ~ date, data=mydata, FUN=sum)
qplot(date, data=totalsteps, geom="bar", weight=steps, ylab="steps")
```

###Calculate and report the mean and median total number of steps taken per day
```{r}
mean(totalsteps$steps)
median(totalsteps$steps)
```
## What is the average daily activity pattern?

###Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
intervalsteps <- aggregate(steps ~ interval, data=mydata, FUN=mean)
ggplot(intervalsteps, aes(interval, steps)) + geom_line() + xlab("") + ylab("Interval steps")
```
###Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
intervalsteps$interval[which.max(intervalsteps$steps)]
```
## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
