---
title: "Reproducible Research: Peer Assessment 1"
author: Alejandro
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load the necessary R packages.
```{r loadpackages}
library(dplyr)
library(xtable)
```

1. Load data from zipfile.
```{r loaddata}
zip_file = unzip("activity.zip")
dat = read.csv(zip_file)
```

2. Transfrom data for analysis.
```{r transformdata}
dat = transform(dat, date = as.Date(date))
```

## What is mean total number of steps taken per day?
Compute total and average numbers of steps per day using *group_by()* and *summarize()* functions from *dplyr* package.
```{r meanstepsbydaytable, results='asis'}
dat = group_by(dat, date)
steps.by.day = summarise(dat, 
                              total.steps = sum(steps) / 1000,
                                mean.steps = mean(steps, na.rm = TRUE),
                                median.steps = median(steps, na.rm = TRUE))
```
1. Make a histogram of the total number of steps by day
```{r totalstepshistogram}
with(steps.by.day, hist(total.steps, breaks = 30, col = "light blue",
                        border = "white", 
                        main = "Histogram of Total Steps per Day",
                        xlab = "Number of Steps (in thousands)"))
```

2. Create a table to report the mean, median, and total number of steps taken per day
```{r stepstable, results='asis'}
steps.by.day %>% xtable(type = "html") %>% print(comment = FALSE)
```

## What is the average daily activity pattern?
dat = group_by(dat, interval)
steps.by.interval = summarise()


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
