Peer-graded Assignment: Course Project 1
===========================================
  Loading and preprocessing the data

Reading the activity .csv file:
  
```{r}
activity <- read.csv("activity.csv", header = TRUE)
```

Converting and the testing of the date format:
```{r}
activity$date <- format(as.Date(activity$date, "%m/%d/%Y"), "%Y-%m-%d")
str(activity$date)
activity$date <- as.Date(activity$date)
str(activity$date)
```

What is mean total number of steps taken per day?

Creating dataframe of daily cumulative steps:
```{r}
activity_steps_day <- with(activity, aggregate(steps, by = list(date), sum))
colnames(activity_steps_day) <- c("date","steps")
```

Histogram of the daily cumulative steps:
```{r}
hist(activity_steps_day$steps, xlab="Daily Steps", main="Daily Steps Histogram")
```

Computing the daily mean of the number of steps:
```{r}
mean(activity_steps_day$steps, na.rm = TRUE)
```

Computing the daily median of the number of steps
```{r}
median(activity_steps_day$steps, na.rm = TRUE)
```

Setting up average number of step per interval
```{r}
activity_steps_interval <- aggregate(steps ~ interval, activity, mean)
```

Line plot of average number of steps per 5 minute interval
```{r}
plot(activity_steps_interval$interval,
     activity_steps_interval$steps, 
     type = "l", 
     xlab="Time of Day", 
     ylab = "Steps",
     main = "Average Number of Steps")
```

Sorting using arrange to see the highest number of average steps:
```{r}
library(dplyr)

arrange(activity_steps_interval, desc(activity_steps_interval$steps)) [1,]
```
Double checking by getting max value:
```{r}
max(unlist(activity_steps_interval$steps), na.rm = TRUE)
```
The value matches the value from the first output.

Imputing missing values

Counting the number of NAs:
```{r}
length(activity$steps[is.na(activity$steps)])
length(activity$date[is.na(activity$date)])
length(activity$interval[is.na(activity$interval)])
```

Because there are not any NAs in the date and interval fields, the NA count from the steps field is enough.


Creating the imputed data dataframe:
```{r}
activity_impute <- activity
```

Merging the imputed dataframe and the earlier average activity step interval created earlier, thus using the average by interval.
```{r}
activty_merge <- merge(activity_impute, activity_steps_interval, by="interval")
```

Creating a list where steps are NA:
```{r}
activty_na_list <- which(is.na(activity_impute$steps))
```
Replacing the NAs with the average interval values:
```{r}
activity_impute[activty_na_list,"steps"] <- activty_merge[activty_na_list,"steps.y"]
```
Imputed data frame for graph and the graph:
```{r}
activity_imputed_steps_day <- with(activity_impute, aggregate (steps, by = list(date), sum))
colnames(activity_imputed_steps_day) <- c("date","steps")
hist(x=activity_imputed_steps_day$steps, xlab="Daily Steps", main="Daily Steps Histogram (Imputed)")
```
Mean of the steps with NAs removed:
```{r}
mean(activity_imputed_steps_day$steps)
```
Median of the steps with NAs removed:
```{r}
median(activity_imputed_steps_day$steps)
```
Setting up week dataframe:
```{r}
activity_week <- activity
```
Adding the day of week column:
```{r}
activity_week$day_of_week <- weekdays(activity$date)
```
Adding the weekday/weekend column:
```{r}
activity_week$weekday_weekend <- ifelse(activity_week$day_of_week %in% c("Saturday","Sunday"), "Weekend","Weekday")
```
Opening lattice package to enable the request graph:
```{r}
library(lattice)
```
Setting up the data for the graph:
```{r}
activity_week_data <- aggregate(steps ~ weekday_weekend + interval, activity_week, mean)
```

Plot of the data:
```{r}
xyplot(steps ~ interval | activity_week_data$weekday_weekend,
       data = activity_week_data,
       type = "l",
       layout = c(1,2),
       main = "Average Steps per Interval by Weekend/Weekday",
       xlab = "Interval",
       ylab = "Averge Steps")