---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Introduction

This **R markdown document** processable by knitr encompasses the complete assignment specified as part of the [Course Project 1](https://www.coursera.org/learn/reproducible-research/peer/gYyPt/course-project-1), offered through the **Reproducible Research** course offered through Coursera platform, and associated to the [Data Science specialization](https://www.coursera.org/specializations/jhu-data-science) provided by the by the **Johns Hopkins University**.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading and preprocessing data

### 1. Load the data (i.e. read.csv())

```{r, echo=FALSE}
# Packages used
library(ggplot2)
```

After unzipping the file, the dataset is stored in a comma-separated-value (CSV) file named **activity.csv** and there are a total of 17,568 observations in this dataset. If it has not already been loaded, it is loaded into memory.


```{r}
unzip("activity.zip")
if (!exists("rawdata")){
    rawdata <- read.csv("activity.csv")
}
head(rawdata)
```

As can be seen above, the variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* date: The date on which the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken

### 2. Process/transform the data (if necessary) into a format suitable for your analysis

For a further analysis, the initial dataset will need to be processed/transformed, since it will be required to aggregate number of steps by date. Additionally we also get rid of rows containing missing values by obtaining a subset and saving it to a new data frame:

```{r}
df_day <- aggregate(steps ~ date, rawdata, sum, na.rm = TRUE)
```

## What is mean total number of steps taken per day?

### 1. Calculate the total number of steps taken per day

```{r}
ggplot(df_day,aes(steps)) + geom_histogram(bins=ceiling(max(df_day$steps)/1000),col="white") + 
    xlim(0,ceiling(max(df_day$steps)/1000)*1000) + 
    xlab("Steps") + ylab("Frequency") +
    ggtitle("Histogram of Total Steps per Day\n(Bin width = 1000 steps)")
```

### 2. Calculate and report the mean and median of the total number of steps taken per day

```{r}
mean(df_day$steps)
```

```{r}
median(df_day$steps)
```

So that the **mean** is **10766** steps and the **median** is **10765** steps.

## What is the average daily activity pattern?

### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

To make this new time series plot, we must go back to the initial dataset and aggregate but this time by the interval variable and using mean() function instead of sum() one:

```{r}
df_int_mean <- aggregate(steps ~ interval, rawdata, mean, na.rm = TRUE)

ggplot(df_int_mean,aes(x=interval,y=steps)) + geom_line() +
    xlab("Interval No.") + ylab("Steps") +
    ggtitle("Average Steps per Five Minutes Interval across all days")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max_steps   <- max(df_int_mean$steps)
df_int_mean[df_int_mean$steps == max_steps,]
print(paste("The 5-minute interval containing the max number of steps is the",
            df_int_mean[df_int_mean$steps == max_steps,1], "with a total of",
            round(df_int_mean[df_int_mean$steps == max_steps,2],2), "steps"))
```

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as **NA**). The presence of missing days may introduce bias into some calculations or summaries of the data.

### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
missingValues <- sum(is.na(rawdata$steps))
print(paste("There are", missingValues, "missing values in the dataset" ))
```

### 2. Devise a strategy for filling in all of the missing values in the dataset.

One of the most simple strategies can consist on replacing NA values by the mean already calculated above for that 5-minute interval.

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r}
df_wo_na <- rawdata
na_index <- is.na(df_wo_na$steps)
int_mean <- tapply(rawdata$steps, rawdata$interval, mean, 
                   na.rm=TRUE, simplify = TRUE)
df_wo_na$steps[na_index] <- int_mean[as.character(df_wo_na$interval[na_index])]

print(paste("The amount of NA entries for Steps variable is now", sum(is.na(df_wo_na$steps))))
```

### 4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. 

```{r}
df_wo_na_day <- aggregate(steps ~ date, df_wo_na, sum, na.rm = TRUE) 
ggplot(df_wo_na_day,aes(steps)) + geom_histogram(bins=ceiling(max(df_day$steps)/1000),col="white") + 
    xlim(0,ceiling(max(df_day$steps)/1000)*1000) + 
    xlab("Steps") + ylab("Frequency") +
    ggtitle("Histogram of Total Steps per Day (with missing data imputed)\n(Bin width = 1000 steps)")
```

```{r}
mean(df_wo_na_day$steps)
```

```{r}
median(df_wo_na_day$steps)
```

The new **mean** and **median** values are now the same, **10766.19**. Compared to initial values the mean does not change, while the median varies a little, becoming identical to the mean. 

## Are there differences in activity patterns between weekdays and weekends?

### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
#function to decide if a given day is a week day or not
isWeekDay <- function(date){
    wd <- weekdays(date)
    ifelse(wd == "Saturday" | wd == "Sunday", "weekend", "weekday")
}

day_type            <- sapply(as.Date(df_wo_na$date), isWeekDay)
df_wo_na$day_type   <- as.factor(day_type)
head(df_wo_na)
```

### 2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
df_final <- aggregate(steps ~ interval + day_type, df_wo_na, mean, na.rm = TRUE)

ggplot(df_final,aes(x=interval,y=steps)) + geom_line() + facet_grid(day_type ~ .) +
    xlab("Interval No.") + ylab("Steps") +
    ggtitle("Average Steps per Five Minutes Interval across all days")
```

From above panel plot the following findings are observed:

* Activity arise earlier at weedays compared to weekends.
* From 10 AM to 5 PM, the weekends have higher activity levels than the weekdays.
