---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r loadingpackages}
library(dplyr)
library(ggplot2)
setwd("D:/zhengcl/Datascience/reproducible1")
rawdata <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

### Here is the histogram for mean total number of steps taken per day:
```{r}
days <- group_by(rawdata, date)
daystep <- summarize(days,  steps=sum(steps, na.rm=TRUE))
q <- ggplot(daystep, aes(steps)) + geom_histogram(bins=10)
q   

avestep <- mean(daystep$steps)
medianstep <- median(daystep$steps)
summary(daystep$steps)
```

The average total number of steps taken per day is `r avestep` and the median is `r medianstep`. 

## What is the average daily activity pattern?

This time series plot shows the average daily activity pattern:
```{r}
intervals <- group_by(rawdata, interval)
timeser <- summarize(intervals,  steps=mean(steps, na.rm=TRUE))
q <- ggplot(timeser, aes(interval, steps)) + geom_line()
q
maxstep <- max(timeser$steps)
maxinterval <- timeser[match(maxstep,timeser$steps),1]
maxinterval
```
The `r maxinterval` 5-min interval contains maxmim number of steps.

## Imputing missing values

### Here is the histogram showing daily activity distribution after missing values are filled in:
```{r }
numofmissing <- sum(is.na(rawdata))
# this function replace NA in a data frame with mean value in that interval and return a data frame with 
# the missing data filled in
replace <- function(df){
  inter <- group_by(rawdata, interval)
  timeser1 <- summarize(intervals,  steps=mean(steps, na.rm=TRUE))
  df$steps[is.na(df$steps)] <- timeser1$steps[match(df$interval, timeser1$interval)][is.na(df$steps)]
  df
}
  
rawdata1 <- replace(rawdata)  # replace missing value
# draw histogarm for daily steps distribution  
days1 <- group_by(rawdata1, date)
daystep1 <- summarize(days1,  steps=sum(steps))
q <- ggplot(daystep1, aes(steps)) + geom_histogram(bins=10)
q
avestep1 <- mean(daystep1$steps)
medianstep1 <- median(daystep1$steps)
summary(daystep1$steps)
```
We can see the original data have `r numofmissing` missing value. After they are filled in, both numbers of mean and median are increased, however, imputing missing value has biger effect on the calculationg of mean value.   

## Are there differences in activity patterns between weekdays and weekends?
```{r}
rawdata1$date <- as.Date(rawdata1$date)
weekday <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
rawdata2 <- mutate(rawdata1,wday=factor((weekdays(rawdata1$date) %in% weekday),levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))) 

intervals1 <- group_by(rawdata2, wday,interval)
timeser2 <- summarize(intervals1,  steps=mean(steps))
g <- ggplot(timeser2, aes(interval, steps)) + geom_line() + facet_grid(wday~ .)
g
```

From the plot, we can see the person in weekends is more active than weekdays.