---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#remove scientific notation
options("scipen"=999)
#create figure directory
dir.create("figures")
```

## Loading and preprocessing the data

```{r}
library(dplyr)
library(stringr)

#Read in activity data
raw_activity_data <- read.csv("activity.csv")
head(raw_activity_data)
summary(raw_activity_data)

#convert dates to datetime
activity_data <- raw_activity_data %>% mutate(date = as.POSIXct(date, format = "%Y-%m-%d", tz=Sys.timezone()))
head(activity_data)
```
## What is mean total number of steps taken per day?
```{r}
library(ggplot2)

#Summarise number of steps taken each day
daily_steps <- activity_data %>% group_by(date) %>%
                  summarise(steps = sum(steps, na.rm =TRUE))
head(daily_steps)

#Plot number of steps taken each day
q <- ggplot(daily_steps, aes(date,steps)) + 
                    labs(title = "Daily Steps taken by Wearer of Activity Monitoring Device",
                         y = "Number of Steps", x = "Date")

q + geom_bar(stat="identity")
#Save plot to file
dev.copy(png, file="figures/daily_steps.png")
dev.off() #Close PNG device

avg_daily_steps <- round(mean(daily_steps$steps, na.rm = TRUE),0)
median_daily_steps <- median(daily_steps$steps, na.rm = TRUE)
```
- The mean number of steps taken each day is `r round(avg_daily_steps,0)`. 
- The median number of steps taken each day is `r median_daily_steps`.

## What is the average daily activity pattern?
```{r}
#Group number of steps by interval and calculate mean
interval_steps <- activity_data %>% group_by(interval) %>%
                  summarise(avg_steps = mean(steps, na.rm = TRUE))
head(interval_steps)

#Plot average number of steps for each interval
plot(interval_steps$interval, interval_steps$avg_steps, type = "l",
     main = "Average number of steps for each 5-minute interval",
     xlab = "Interval",
     ylab = "Number of Steps")

#Save plot to file
dev.copy(png, file="figures/avg_steps_per_interval.png")
dev.off() #Close PNG device

#Find interval with maximum average of steps
max_interval <- interval_steps$interval[which.max(interval_steps$avg_steps)]

#Convert interval into time format
interval_in_minutes <- vector()
interval_in_minutes[1] <- interval_steps$interval[which.max(interval_steps$avg_steps)]
interval_in_minutes[2] <- interval_steps$interval[which.max(interval_steps$avg_steps)+1]

interval_in_minutes <- paste0(str_pad(interval_in_minutes, 4, pad = 0),"00") %>%
                        strptime("%H%M%S") %>% format("%H:%M")
```

- The  interval with the highest average number of steps is `r max_interval` which corresponds to the 5-minute
time interval `r interval_in_minutes[1]` - `r interval_in_minutes[2]`.

## Imputing missing values
```{r}
#Find number of missing rows
no_of_missing_rows <- sum(!complete.cases(activity_data))
```
The total number of rows with missing numbers is `r no_of_missing_rows`.

```{r}
#Replace missing values with average for the 5-minute interval
activity_data_clean <- activity_data %>%
                        mutate(steps = ifelse(is.na(steps),
                         interval_steps$avg_steps[interval_steps$interval == interval],
                          steps))
head(activity_data_clean)

#Calculate daily steps in imputated data
daily_steps_clean <- activity_data_clean %>% group_by(date) %>%
                  summarise(steps = sum(steps, na.rm =TRUE))
head(daily_steps)

#Plot imputed data
q2 <- ggplot(daily_steps_clean, aes(date,steps)) + 
                    labs(title = "Daily Steps taken by Wearer of Activity Monitoring Device",
                         y = "Number of Steps", x = "Date")

q2 + geom_bar(stat="identity")
#Save plot to file
dev.copy(png, file="figures/daily_steps_imputed_data.png")
dev.off() #Close PNG device

#Mean number of daily steps
avg_daily_steps_clean <- mean(daily_steps_clean$steps)

#Median number of daily steps
median_daily_steps_clean <- median(daily_steps_clean$steps)

#Relative percentage change in mean number of daily steps between raw and imputed data
rel_change_mean <- 100*(avg_daily_steps_clean - avg_daily_steps)/avg_daily_steps

#Relative percentage change in median number of daily steps between raw and imputed data
rel_change_median <- 100*(median_daily_steps_clean - median_daily_steps)/median_daily_steps
```

- The mean number of daily steps in the imputated activity data is `r round(avg_daily_steps_clean,0)`.
- The median number of daily steps in the imputated activity data is `r median_daily_steps_clean`.

- The mean daily steps in the imputated activity data is `r round(rel_change_mean,2)`% higher than the mean in the raw activity data.
- The median daily steps in the imputated activity data is `r round(rel_change_median,2)`% higher than the median in the raw activity data.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
# Define vector of Weekend days
weekend = c("Saturday","Sunday")

# create dataset of avg. steps per interval grouped by Weekday/Weekend
interval_steps_by_day_type <- activity_data_clean %>%
                     mutate(type_of_day = weekdays(date),
                            type_of_day = ifelse(type_of_day %in% weekend,"weekend","weekday")) %>%
                    group_by(type_of_day,interval) %>%
                    summarize(steps = mean(steps, na.rm = TRUE))
head(interval_steps_by_day_type)

library(lattice)

#Plot average number of steps vs. interval for Weekdays and Weekends
xyplot(interval_steps_by_day_type$steps ~ interval_steps_by_day_type$interval | interval_steps_by_day_type$type_of_day,
       type = "l",
       main = "Avg. Steps during 5 minute intervals on Weekdays and Weekends",
       xlab = "Interval",
       ylab = "Number of Steps",
       layout = c(1, 2)) ## Plot with 2 panels

#Save plot to file
dev.copy(png, file="figures/avg_steps_weekdays_weekends.png")
dev.off() #Close PNG device
```