``` {r setoptions, echo=FALSE}
opts_chunk$set(cache=TRUE)
```
# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r}
data <- read.csv("activity.csv", header=TRUE)
clean_data <- data[which(data$steps != "NA"), ]
```
## What is mean total number of steps taken per day?
```{r}
library(plyr)
total_by_day <- ddply(clean_data, .(date), summarise, steps=sum(steps))
hist(total_by_day$steps, main="Number of Steps", 
     xlab="Total number of steps taken each day", col="light blue")

# mean and median total number of steps taken per day
mean(total_by_day$steps)
median(total_by_day$steps)
```
## What is the average daily activity pattern?
```{r}
average_by_interval <- ddply(clean_data, .(interval), summarise, steps=mean(steps))
plot(average_by_interval$interval, average_by_interval$steps, type="l", 
     col="blue",
     xlab="5-minute interval", 
     ylab="Average number of steps taken",
     main="Average daily activity pattern")

# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
average_by_interval[average_by_interval$steps==max(average_by_interval$steps),]
colnames(average_by_interval)[2] <- "intervalAvg"
```
## Imputing missing values
```{r}
# Total number of missing values in the dataset
sum(is.na(data$steps))
# Fill NA's with average for that 5-min interval
merged <- arrange(join(data, average_by_interval), interval)
# Create a new dataset that is equal to the original dataset but with the missing data filled in.
merged$steps[is.na(merged$steps)] <- merged$intervalAvg[is.na(merged$steps)]
# Histogram
new_total_by_day <- ddply(merged, .(date), summarise, steps=sum(steps))
hist(new_total_by_day$steps, main="Number of Steps", 
     xlab="Total number of steps taken each day", col="light blue",)
# mean and median total number of steps taken per day
mean(new_total_by_day$steps)
median(new_total_by_day$steps)
total_steps1 <- sum(clean_data$steps)
total_steps2 <- sum(merged$steps)
total_diff <- total_steps2 -total_steps1 []
```
Mean values didn't change, because I supplied interval averge for the given interval, so total average didn't change. Histogram and median changed somewhat. filling values added total of `r total_diff` steps.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
library(lattice)
weekdays <- weekdays(as.Date(merged$date))
data_with_weekdays <- transform(merged, day=weekdays)
data_with_weekdays$wk <- ifelse(data_with_weekdays$day %in% c("Saturday", "Sunday"),"weekend", "weekday")
average_by_interval_wk <- ddply(data_with_weekdays, .(interval, wk), summarise, steps=mean(steps))

xyplot(steps ~ interval | wk, data = average_by_interval_wk, layout = c(1, 2), type="l")
```
