---
title: "Reproducible Research - Peer assesment 1"
author: "Luis F Botero"
date: "Tuesday, February 10, 2015"
output: html_document
---
Loading andpreprocessing data

```{r, echo=TRUE}
require(ggplot2)
library(plyr)
library(dplyr)
#setwd("./R/reproducible_research/peer_assesment1")

data<-read.csv("activity.csv")

q0<-data.frame(data)
q1<-subset(q0,steps!="NA")

# Asigns weekday names
q1$day <- weekdays(as.Date(q1$date))

```

What is the mean total number of steps taken per day
1. Calculate the total number of steps taken per day

```{r, echo=TRUE}
q2 <- aggregate(q1[c("steps")], list(date = q1$date), sum)
```

2. Histogram of the total number of steps taken every day

```{r, echo=TRUE}
ggplot(q2,aes(date,steps))+
    geom_histogram(stat="identity")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
    labs(title="Total Steps per date")
````

3. Calculate and report mean an median of the total number of steps taken per day

Mean

```{r, echo=TRUE}
# Agregates by weekday
q3_mean<-aggregate(steps~date,q2,mean)
q3_mean[order(q3_mean$date),]
```

Median

```{r, echo=TRUE}
q3_median<-aggregate(steps~date,q2,median)
q3_median[order(q3_mean$date),]
```

What is the average daily activity pattern?

1. Make a Time series plot of the 5-minute interval and the avg number of steps,averaged across all days (y-axis)

```{r, echo=TRUE}

r<-aggregate(steps~interval,q1,mean)
names(r)<-c("interval","rmean")

plot(r$interval,r$rmean,type="l",xlab="Interval",ylab="Avg steps",main="Average daily steps")
max_steps=max(r$rmean)
abline(h=max_steps,col="gray60")
v=r[r$rmean==max_steps,]
abline(v=v$interval,col="gray60")

text(120,max_steps-20,floor(max_steps*100)/100)
text(v$interval-50,10,v$interval)
```

2. Which 5 minute interval contains the maximum number of steps ?

```{r, echo=TRUE}
c("Interval",v$interval)
c("Avg number of steps",max_steps)

```

Imputing Missing Values

1. Calculate and report total number of missing values in dataset (NA)

```{r, echo=TRUE}
length(which(is.na(q0$steps)))
```

2. Devise a strategy to fill the missing values in dataset

Filling with average steps during that interval

```{r, echo=TRUE}
# r  frame contains avg steps by interval,  r$rmean
s<-q0  # to preserve raw data
# merges with R, which contains average steps by interval
s1<-merge(s,r)
#hom_steps is steps when not NA and  rmean when NA
s1$hom_steps<-ifelse(is.na(s1$steps)==TRUE,s1$rmean,s1$steps)
```

3. Create a new dataset that is equal to the original but with the missing data filled in
```{r, echo=TRUE}
t<-data.frame(s1$hom_steps,s1$date,s1$interval)
names(t)<-c("steps","date","interval")
```

4. Make a histogram with the total number of stepstaken each day
calculate mean and median steps per day

```{r, echo=TRUE}
t1<-t
t1$day <- weekdays(as.Date(t1$date))
t2 <- aggregate(t1[c("steps")], list(date = t1$day), sum)

t2$date <- factor(t2$date, levels= c("Sunday", "Monday", 
    "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
t3<-t2[order(t2$date),]

ggplot(t3,aes(date,steps))+
    geom_histogram(stat="identity")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
    labs(title="Total Steps per date")
````

Mean: Comparison between original without NAs and Filled for NAs

```{r, echo=TRUE}
# Agregates by weekday
t2b<-aggregate(steps ~ date,t1,sum)
t4_mean<-aggregate(steps ~ date,t2b,mean)
names(t4_mean)<-c("date","filled steps")
t6<-merge(t4_mean,q3_mean,all=TRUE)
t6

```

Median: Comparison between original without NAs and filled for NAs

```{r, echo=TRUE}
t4_median<-aggregate(steps ~ date,t2b,median)
names(t4_median)<-c("date","filled_steps")
t7<-merge(t4_median,q3_median,all=TRUE)
t7
```

Are there differences in activity between weekdays and weekends ?

1. Create a new factor variable with two labels= weekday and weekdend

```{r, echo=TRUE}

m<-t1
m$wday<-ifelse(m$day =="Sunday" | m$day =="Saturday","Weekend","Weekday")
m$wday<-as.factor(m$wday)
```

2. Make a panel plot showind differences in average steps by interval during weekends and weekdadys

```{r, echo=TRUE}
m2 <- aggregate(steps~wday+interval,data=m,FUN=mean)

ggplot(m2,aes(interval,steps))+
  geom_line()+
  labs(title = "Average steps per interval")+ 
  labs(x="Interval")+
  labs(y="Number of steps")+
  facet_grid(wday~.)   
  
  


