#Reproducible Research : Peer Assignment 1
========================================================

##setting the current working directory 

```{r}
setwd("C:\\Users\\ravitejaj\\Desktop\\")
getwd()
```

1.Loading and preprocessing the data
##Loading the data using read.csv() function and converting it to a dataframe

```{r}
data=read.csv("activity.csv")
data=as.data.frame(data)
```


1.What is mean total number of steps taken per day?

2.Calculate and report the mean and median total number of steps taken per day

### For calculation of mean firstly sum should be calculated so aggregate function is used

```{r}
l <- aggregate(steps ~ date, data=data, FUN=sum)
plot(l$steps,xlab="",ylab="Total number of Steps",main="Total No of Steps per day",col="Green")
mean(l$steps)
median(l$steps)

```

##What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

### calculating the maximum mean value using max function

```{r}
y<-aggregate(steps~interval,data=data,FUN=mean)
plot(y$interval,y$steps,xlab="Intervals",ylab="Average Steps",type="l")
z=y[y$steps==max(y$steps),]
z[1] #it prints the interval is 835
```

## Imputing missing values

## Calculating the no of NA's using the complete.cases

```{r}
nrow(data[!complete.cases(data),])


```

###Strategy:
###I will replace the NA values with 5 min interval means that is the dataset 'y'

##3. Create a new dataset that is equal to the original dataset but with the missing data filled in.


```{r}
data <- merge(data, y, by="interval", suffixes=c("",".y"))
nas <- is.na(data$steps)
data$steps[nas] <- data$steps.y[nas]
data <- data[,c(1:3)]
l1<- aggregate(steps ~ date, data=data, FUN=sum)
plot(l1$steps,xlab="",ylab="Total number of Steps",main="Total No of Steps per day",col="Green")
mean(l1$steps)
median(l1$steps)

```

##There is slight difference in the things mean and median

## Are there differences in activity patterns between weekdays and weekends?

## Converting the dates to weektypes that is weekend r weekday


```{r,cache=TRUE}
weektype <- function(date) {
    if (weekdays(as.Date(date)) %in% c("Saturday", "Sunday")) {
        "weekend"
    } else {
        "weekday"
    }
}
data$day<- as.factor(sapply(data$date, weektype))
```


## Panel Plot

```{r,echo=TRUE}
par(mfrow=c(2,1))
{
  for(type in c("weekend","weekday"))
    {
    x<-aggregate(steps ~ interval,data=data,subset=data$day==type,FUN=mean)
    plot(x,type="l",main=type,col="blue")
    }
}

```
