---
title: "Assignment1"
author: "sabineX"
date: "27 August 2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

 

```{r }
  library(dplyr)
  library(ggplot2)
  df <- read.csv("C:/Users/Sabine/Downloads/ass1/activity.csv")
  df$date <- as.Date(df$date, format= "%Y-%m-%d")
  my_df <- subset(df,  !(is.na(steps) |  is.na(interval)  ))
```



#What is mean total number of steps taken per day?
```{r }
  steps_per_day <- with(my_df, aggregate(steps, by = list(date), FUN = sum, na.rm = TRUE))
  hist(steps_per_day$x, ylim = c(0,30), xlab= "steps",main= "Total number of steps taken per day")
  abline(v=mean(steps_per_day$x), col="red", lwd=5)
  abline(v=median(steps_per_day$x), col="black",lty="longdash", lwd=3)
```



#What is the average daily activity pattern?
```{r }
  average_daily_activity <- aggregate(df$steps, by=list(df$interval), FUN=mean, na.rm=TRUE)
  names(average_daily_activity) <- c("interval", "mean")
  plot(average_daily_activity$interval, average_daily_activity$mean,type = "l",xlab="Interval", ylab="Average number of steps", main="Average number of steps per intervals")
 
  val<-subset(average_daily_activity, average_daily_activity$mean==max(average_daily_activity$mean))
  abline(v=val$interval, col="red", lwd=2)
```



#Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r }
size_df<- as.numeric(count(df))
  size_my_dfx<- as.numeric(count(subset(df,  (is.na(steps) |  is.na(interval)  ))))
  MissingValues<-c(size_my_dfx,size_df)
  barplot(MissingValues, ylim = c(0,20000), main ="Number of datasets with missing values  /  Number of complete datasets")
```



#Devise a strategy for filling in all of the missing values 
#Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r }
  df_new<-df
  df_new$input_steps <- average_daily_activity$mean[match(df_new$interval, average_daily_activity$interval)]
  ind <- is.na(df_new$steps)
  df_new$steps[ind] <- df_new$input_steps[ind] 
  df_new$input_steps<-NULL
  
```



#Make a histogram of the total number of steps
```{r }
  steps_per_day_NA_replaced <- with(df_new, aggregate(steps, by = list(date), FUN = sum, na.rm = TRUE))
  hist(steps_per_day_NA_replaced$x,xlab =  "steps" ,main= "Total number of steps  ( NA replaced )")
```



#Do these values differ from the estimates from the first part of the assignment?
```{r }
  par(mfrow=c(1,2))
  hist(steps_per_day$x, ylim = c(0,35), xlab = "steps" , main= "Total nr. of steps ")
  abline(v=mean(steps_per_day$x),col="red", lwd=2)
  hist(steps_per_day_NA_replaced$x, ylim = c(0,35), xlab = "steps", main= "Total nr. of steps(NA replaced)")
  abline(v=mean(steps_per_day_NA_replaced$x),col="red", lwd=2)
```  



#Are there differences in activity patterns between weekdays and weekends?
```{r }
  weekday<-subset(df_new, ((weekdays(date)=="Montag")|(weekdays(date)=="Dienstag")|(weekdays(date)=="Mittwoch")|(weekdays(date)=="Donnerstag")|(weekdays(date)=="Freitag")))
  weekend<-subset(df_new, ((weekdays(date)=="Samstag")|(weekdays(date)=="Sonntag")))
  w1<-subset(df_new, ((weekdays(date)=="Sonntag")))
  
  par(mfrow=c(2,1),mar=c(0.0,4.1,4.1,0.0))
    average_weekday <- aggregate(weekday$steps, by=list(weekday$interval), FUN=mean, na.rm=TRUE)
    names(average_weekday) <- c("interval", "mean")
    plot(average_weekday$interval, average_weekday$mean,type = "l",xlab="", ylab= " Weekday", main="Average number of steps")
    val<-subset(average_weekday, average_weekday$mean==max(average_weekday$mean))
 
    average_weekend <- aggregate(weekend$steps, by=list(weekend$interval), FUN=mean, na.rm=TRUE)
    names(average_weekend) <- c("interval", "mean")
    plot(average_weekend$interval, average_weekend$mean,type = "l",xlab="Interval", ylab="Weekend", main="")
    val<-subset(average_weekend, average_weekend$mean==max(average_weekend$mean))
```  
