---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Initially, missing entries will be removed using the complete.cases function

```{r}
data <- read.csv("activity.csv")

clean <- complete.cases(data)
cleandata <- data[clean,]
```

## What is mean total number of steps taken per day?

The total steps on a given day is produced in a data table as follows, using the clean data (NA removed):

```{r}
library(data.table)

cleandata.dt <- data.table(cleandata)
totalsteps <- cleandata.dt[,list(total.steps = sum(steps)), by='date']
```

A histogram of the data is shown below:

```{r}
hist(totalsteps$total.steps, breaks = 20)
```

The mean and median total number of steps are therefore:

```{r}
mean(totalsteps$total.steps)
median(totalsteps$total.steps)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
