---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First, the data is loaded:

```{r}
tab=read.csv("D:\\celina\\reproduce\\activity.csv")
```


## What is mean total number of steps taken per day?
The average steps per day are

```{r}
aggregate(steps~date,data=tab,mean)   
```


## What is the average daily activity pattern?
The median steps per day are

```{r}
aggregate(steps~date,data=tab,median)   
```

As for the  average daily activity pattern, according to the line chart, the trend is similar to inverse bell. The maximum number of steps is found as below.

```{r max,fig.align='center'}
r2=aggregate(steps~interval,data=tab,mean)    ## avg step/interval
plot(r2$steps~r2$interval,type='l',col='blue',xlab="interval",ylab="steps")
r2[which(r2$steps==max(r2$steps)),] 
```


## Imputing missing values
As the data shown above, the tracking data is totaly missing on some specifique day such as "2012-10-01". So roughly we replace the missing data with the mean of 5-minute interval. 


```{r}
c=tab[is.na(tab),]        ## na data
dim(c)
summary(c)
d=rep(r2$steps,8)
c$steps=d
tab2=merge(c,tab,by=c("date","interval"),all.y=T)
tab2$steps.x[is.na(tab2$steps.x)]=0
tab2$steps.y[is.na(tab2$steps.y)]=0
tab2$steps.new=tab2$steps.x+tab2$steps.y
```
Now the new steps/day is as below, also the new mean and median. 

```{r}
a2=aggregate(steps.new~date,data=tab2,sum)  
barplot(a2$steps.new,col="blue")
aggregate(steps.new~date,data=tab2,median )
aggregate(steps.new~date,data=tab2,mean )
```

Obvously we have the complete dataset because the NA is overlooked automatically by R. So the mean of data is larger than before.

```{r,fig.align='center'}
par(mfrow=c(1,2))
boxplot(tab$steps,main="Boxplot of original data")
boxplot(tab2$steps.new,main="Boxplot of after filling missing data")
```

## Are there differences in activity patterns between weekdays and weekends?
When the weekday is input into data, we find the steps/interval is totally different. Overall the steps/interval is smaller on weekend than on weekday.

```{r,fig.align='center'}
tab2$weekday=weekdays(as.Date(tab2$date))
tab2$wd=1
for (i in 1:17568){
if (tab2$weekday[i] =="Sunday") tab2$wd[i]=0
if (tab2$weekday[i] =="Saturday") tab2$wd[i]=0
}
b2=aggregate(steps.new~interval+wd,data=tab2,mean)    ## avg step/interval
b2.wd=b2[which(b2$wd==1),]
b2.wkend=b2[-which(b2$wd==1),]
plot(b2.wd$steps~b2.wd$interval,type='l',col='blue',main="Weekdays of steps/interval")
plot(b2.wkend$steps~b2.wkend$interval,type='l',col='blue',main="Weekends of steps/interval")

```
