---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## 0. Data Source

The data for this assignment can be downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

## 1. Loading and preprocessing the data

* Read the data into a data frame
```{r}
data <- read.csv("activity.csv",header=TRUE)
```

## 2. What is mean total number of steps taken per day?

**2.1** Before making a histogram of the total number of steps taken each day, we'll need to sum all the steps per day (ignore all the missing values)
```{r}
steps <- tapply(data$steps,data$date,na.rm=TRUE,FUN = sum)

steps_by_day <- aggregate(steps ~ date, data = data,sum)
```

Run the summary command to list the min/max, median and mean values of the data set
```{r}
summary(steps_by_day)
```

**2.2** Make a histogram of the total number of steps taken each day
```{r}
hist(dataSteps, breaks=10, ylim = c(0,20), xlim = c(0,25000), 
     xlab = "Total # of Steps", main = "Histogram Of The Total Number Of Steps Taken Each Day")
```
Calculate and report the **mean** and **median** total number of steps taken per day


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
