---
title: "PA_Template"
author: "Scott Semel"
date: "October 17, 2015"
output: html_document
---

## Loading and preprocessing the data

```{r}

library(dplyr)
library(knitr)

dat = read.csv("activity.csv")


by_date = group_by(dat, date)
a = summarize(by_date, sum(steps))
names(a) = c("date", "totalSteps")
#x =  a$date
#y =  as.numeric(a$totalSteps)
#b = cbind(x,y)
#names(b) = c("date", "totalSteps")
totalStepsPerDay = as.numeric(a$totalSteps[!is.na(a$totalSteps)])
hist(totalStepsPerDay, col ="green")
```
##What is mean total number of steps taken per day?
```{r}
mean(totalStepsPerDay)
median(totalStepsPerDay)
```


## What is the average daily activity pattern?
```{r}

mat = matrix(data = dat[,1], ncol = 288, byrow = TRUE)
av = colMeans(mat,na.rm=TRUE)
plot(av, type="l", xlab="5 minute intervals of the day", ylab = "Average number of steps per interval", col=2) 
lines(av, col="blue")
```
## Highest average is 206 steps at interval 104 around 8:40am.

## Find the number of rows with NA. 

```{r}
numOfNAs = sum(is.na(dat$steps))
```
## [1] 2304

## Imputing missing values. 
## Replace each NA with the average of that 5 minute interval for the whole set.

```{r}
allavs = rep(av,length.out = 17568)

newTotalStepsPerDay = pmax(dat$steps, allavs, na.rm = TRUE)
dat$steps = newTotalStepsPerDay

by_date = group_by(dat, date)
a = summarize(by_date, sum(steps))
names(a) = c("date", "totalSteps")
#x =  a$date
#y =  as.numeric(a$totalSteps)
#b = cbind(x,y)
#names(b) = c("date", "totalSteps")
totalStepsPerDay = as.numeric(a$totalSteps[!is.na(a$totalSteps)])
hist(totalStepsPerDay, col ="green")

mean(totalStepsPerDay)
median(totalStepsPerDay)
```

## It increased the average number of steps per day from 10000 to 17000
```{r}

theday = weekdays(as.Date(dat$date,'%Y-%m-%d'))
ind = which(theday =="Saturday" |theday =="Sunday")
typeday = rep("weekday", 17568)
typeday[ind] = "weekend"
dat = cbind(dat,typeday)
names(dat) = c("steps", "date", "interval", "typeday")
by_type = group_by(dat, typeday)

```
## Are there differences in activity patterns between weekdays and weekends?
## Find the average number of steps per 5 min interval per day for weekdays vs weekends.
```{r, echo=FALSE}


newdat = dat[dat$typeday=="weekday",]
mat = matrix(data = newdat[,1], ncol = 288, byrow = TRUE)
av = colMeans(mat)
plot(av, type="l", xlab="5 minute intervals of the day", ylab = "Average number of steps per day", col=2) 
lines(av, col="blue")

newdat2 = dat[dat$typeday=="weekend",]
mat2 = matrix(data = newdat2[,1], ncol = 288, byrow = TRUE)
av2 = colMeans(mat2)
lines(av2, col="red")
legend("topright","(x,y)", c("Weekday", "Weekend"), cex=0.8, col= c("blue", "red"), lty=1)
```
## There's less activity in the morning but more at night on the weekends.

