---
title: "Reproducible Research  -- Activity Monitoring"
author: "SMenon"
date: "March 18, 2017"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read Activity Dataset and Inspect  

Activity data is in a csv file with headers. The file contains three variables:  

* *steps*: Number of steps taking in a 5-minute interval (missing values are coded as **NA**)  
* *date*: The date on which the measurement was taken in YYYY-MM-DD format  
* *interval*: Identifier for the 5-minute interval in which measurement was taken  

```{r}
activityData <- read.csv(file="C:/Users/samenon/Documents/Coursera/ReproResearch/activity.csv",header=TRUE,sep=",")
head(activityData)
```

## 1. Data preparation

Here data is processed for charting and analysis  

### a.First step to remove NA from the steps readings:  

Read and omit missing values  
```{r}
activityDataNoNA = na.omit(activityData)
head(activityDataNoNA)
```

### b. create data set where steps are aggregated by day

Data aggregation  

```{r}
library(dplyr)
activityDataByDay = group_by(activityDataNoNA,date)
activityDataByDay = summarise(activityDataByDay,Sum_Steps=sum(steps))
head(activityDataByDay)
```

### c. create data set where steps are aggregated by interval

Data aggregation  

```{r}
library(dplyr)
activityDataByInt = group_by(activityDataNoNA,interval)
activityDataByInt = summarise(activityDataByInt,Avg_Steps=mean(steps))
head(activityDataByInt)
```

## 2. Histogram of daily steps & print mean/median
Charting  

```{r}
library(ggplot2)
qplot(activityDataByDay$Sum_Steps,geom = "histogram",xlab="Num of steps", ylab="frequency")
print (paste("Median daily steps is:",median(activityDataByDay$Sum_Steps)))
print (paste("Mean daily steps is:",mean(activityDataByDay$Sum_Steps)))
```

## 3 Time series plot of steps by interval & max 5-min interval
Time series plot  

```{r}
 plot.ts(activityDataByInt$interval,activityDataByInt$Avg_Steps,main="Steps by Interval",type="l",xlab="Interval", ylab="Average Steps")
maxIndex <- which.max(activityDataByInt$Avg_Steps)
print(paste("Max Avg. steps happen at interval:",activityDataByInt[maxIndex,1]))
```
## 4 Missing Values

First populate missing values with average of the interval  

```{r activityData}
activityDataAugment <- merge(activityData,activityDataByInt,by="interval")
  activityDataAugment[is.na(activityDataAugment[,2]),2] <- activityDataAugment[is.na(activityDataAugment[,2]),4]
```
  Now create the histogram and estimate mean/medians
```{r}
library(ggplot2)
New_activityDataByDay = group_by(activityDataAugment,date)
New_activityDataByDay = summarise(New_activityDataByDay,Sum_Steps=sum(steps))
qplot(New_activityDataByDay$Sum_Steps,geom = "histogram",xlab="Num of steps", ylab="frequency")
print (paste("Median daily steps is:",median(New_activityDataByDay$Sum_Steps)))
print (paste("Mean daily steps is:",mean(New_activityDataByDay$Sum_Steps)))
```
Here, median has changed since the missing values are populated. The Mean has not changed since the default value was the mean for the interval

## 5 Weekday vs. Weekend analysis

Are there any difference by weekend and weekdays  

```{r}
activityDataAugment$DOW <- weekdays(as.Date(activityDataAugment$date))
activityDataAugment$DOW <- ifelse(activityDataAugment$DOW %in% c("Saturday","Sunday"),"Weekend","Weekday")
activityDataByNewInt = group_by(activityDataAugment,interval,DOW)
activityDataByNewInt = summarise(activityDataByNewInt,Avg_Steps=mean(steps))
ggplot(data=activityDataByNewInt,aes(interval,Avg_Steps))+stat_summary(fun.y=mean,geom="line") + facet_grid(~DOW)
```
  
Here, we see higher volatility in weekday activity
