---
title: "PA1_template"
author: "Lautier"
date: "10 April 2016"
output: html_document
---

Reproducible Research Assignment 1

First step is reading the file (activity monitoring data) into R, which can be done by means of the following:

```{r, echo=TRUE}
activity <- read.csv("activity.csv")
```

Date variable is originally of factor class, so it is useful to change its class to Date:

```{r, echo=TRUE}
activity$date <- as.Date(as.character(activity$date))
```


Q1: What is mean total number of steps taken per day?

Firstly, we have to calculate total number of steps taken per day. Aggregate function comes in handy:

```{r, echo=TRUE}
sum_per_day<- aggregate(steps ~ date, data=activity, sum)
```

Even though there are 61 days in total, the step above provided us with results only for 53 days, because of the NA values. 

We can obtain a histogram of total number of steps taken each day with lines marking the mean and the median by:

```{r, echo=TRUE}
hist(sum_per_day$steps, xlab="Total number of steps per day", ylab="Frequency", main="Total number of steps per day - histogram", freq=TRUE)

abline(v=mean(sum_per_day$steps),col="red")
abline(v=median(sum_per_day$steps),col="green")


legend(x = "topright",
       c("Mean", "Median"),
       col = c("red", "green"),
       lwd = c(2, 2, 2))


```

It seems that median and mean values are located very close to each other and because of that we cannot discern the lines from each other.

To check what is the mean and median, we can use the following code:


```{r, echo=TRUE}
summary(sum_per_day$steps)
```

Mean is equal to 10770 steps per day and median to 10760 steps per day. As noted above, it is true that those values are very close to each other on the histogram.


Q2: What is the average daily activity pattern?

To check what is the average daily activity pattern, first we have to calculate the mean number of steps taken per every 5-minute interval averaged across all days:


```{r, echo=TRUE}
mean_int <- aggregate(steps ~ interval, data=activity, mean)
```

Then we can plot the 5-minute interval and the average number of steps taken, averaged across all days.

```{r, echo=TRUE}
library(lattice)
xyplot(steps~interval,
       data = mean_int,
       type = "l",
       main="Average number of steps per interval",
       xlab="5-minute intervals",
       ylab="Average number of steps")
```

To check which 5-minute interval contains the maximum number of steps, we can use the following code:

```{r, echo=TRUE}
mean_int[which.max(mean_int$steps), ]
```

Interval 835 contains the maximum number of steps, which is equal to 206,2 steps.


Imputing missing values

To check how many missing values there are in the dataset, we can use is.na function:

```{r, echo=TRUE}
NAs <-is.na(activity)
summary(NAs)
```

Then we have to call the summary function of the newly created object:

```{r, echo=TRUE}
summary(NAs)
```

It seems that there are 2304 missing values for steps variable (TRUE represents number of values that were considered NA by is.na function). 