---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Reading using `read.csv()` and convert date field
```{r ,results='hide'}
# Make sure to have your dataset unzipped in `data` directory
# Make sure to have `lubridate` installed
library(lubridate,quietly = TRUE,warn.conflicts = FALSE)
```
```{r ReadingData , echo=TRUE}
dt <- read.csv('data/activity.csv')
dt$date <- ymd(dt$date)
dt$dateTime <- dt$date + minutes(dt$interval)
head(dt)
```

## What is mean total number of steps taken per day?
From the code chunk above we can see that there are some missing values in the dataset. We should make sure to set `na.rm=TRUE`

```{r meanSteps,echo=TRUE,results='hide'}
sumStepsPerDay<-tapply(dt$steps, dt$date, sum,na.rm=TRUE)
meanStepsPerDay <- mean(sumStepsPerDay);
medianStepsPerDay <- median(sumStepsPerDay)
hist(sumStepsPerDay,main = 'Histogram of steps per day',xlab = 'Total Steps per Day')

dev.copy(png,'figure/stepsPerDayHist.png')
dev.off();
```


The mean of total steps taken per day is `r meanStepsPerDay`,
the median is `r medianStepsPerDay`.

## What is the average daily activity pattern?

We should average the number of steps by date.
Some days have no value at all hence they should be excluded from the plot.
```{r dailyPattern, echo=TRUE,results='hide'}
averageStepsPerDayInterval <-tapply(dt$steps, dt$interval, mean,na.rm=TRUE)


plot(names(averageStepsPerDayInterval),
     averageStepsPerDayInterval,
     type='l',
     xlab = 'Interval',
     ylab = 'Average steps per day')
dev.copy(png,'figure/stepsPerDayTimeSeries.png')
dev.off()
maxInterval <- which.max(averageStepsPerDayInterval)
```
Interval `r maxInterval` has the most number of steps

## Imputing missing values(NAs)
Data has `r sum(is.na(dt$steps))` `NA`s. We will impute the missing values by filling them with the mean of the 5-minute interval.


```{r imputeNAs,echo=TRUE}
meanStepsPerInterval <- tapply(dt$steps,dt$interval,mean,na.rm=TRUE)
newDt <- dt
intervals <- newDt[is.na(newDt$steps),'interval']

## Make sure to use `as.character()` with intervals

meanStepsPerInterval <- meanStepsPerInterval[as.character(intervals)]

newDt[is.na(newDt$steps),'steps'] <- meanStepsPerInterval
head(newDt)
```

```{r meanStepsNoNAs,echo=TRUE,results='hide'}
sumStepsPerDayNoNAs<-tapply(newDt$steps, newDt$date, sum)
meanStepsPerDayNoNAs <- mean(sumStepsPerDayNoNAs)
medianStepsPerDayNoNAs <- median(sumStepsPerDayNoNAs)
hist(sumStepsPerDayNoNAs,
     main = 'Histogram of steps per day with no missing values',
     xlab = 'Total Steps per Day')
dev.copy(png,'figure/stepsPerDayHistNoNas.png')
dev.off()
```

The mean of total steps taken per day, with no missing values, is `r round(meanStepsPerDayNoNAs,2)`,
the median is `r medianStepsPerDayNoNAs`.

The results differ slightly from the original data.  

The median is the the same as the mean because all the missing values are filled with the mean, making the mean the most frequent value in the new data.


## Are there differences in activity patterns between weekdays and weekends?
```{r patternByWeekdayWeekend, echo=TRUE,results='hide'}
library(lattice)

newDt$dayType <- ifelse(weekdays(newDt$date) %in% c('Saturday','Sunday'),'weekend','weekday')
newDt$dayType <- as.factor(newDt$dayType)

weekends <- subset(newDt,dayType=='weekend')
weekdays <- subset(newDt,dayType!='weekend')
averageStepsPerWeekend <- tapply(weekends$steps,weekends$interval,mean)

averageStepsPerWeekday <- tapply(weekdays$steps,weekdays$interval,mean)

finalDt <- data.frame(steps=c(averageStepsPerWeekend,averageStepsPerWeekday),
                      dayType = c(rep('weekend',length(averageStepsPerWeekend)),
                                  rep('weekday',length(averageStepsPerWeekday))) )

finalDt$invervals <- as.numeric(row.names(finalDt))

xyplot(steps~intervals|dayType ,
       data = finalDt,
       type='l',
       layout= c(1,2),
       ylab = 'Number of steps',
       xlab='Interval')
dev.copy(png,'figure/stepsWeekdayWeekend.png')
dev.off()
```

We Notice that the number of steps increases at around the 700th interval.
