
# Reproducible Research: Peer Assessment 1



## Loading and preprocessing the data
```{r, echo=TRUE}
library(readr)
library(dplyr)
activity <- read_csv("./activity.zip")
steps_by_day <- group_by(activity, date) %>% summarize(sum(steps))

```


## What is mean total number of steps taken per day?
```{r, echo=TRUE}

library(ggplot2)
ggplot(steps_by_day, aes(x = date, y = steps_by_day$`sum(steps)`)) + 
        geom_col() +
        theme_light() +
        labs(x = "Date", y = "Steps", title = "Steps by day")

summary(steps_by_day$`sum(steps)`)
```

mean = 10766
median = 10765


## What is the average daily activity pattern?
```{r}
steps_by_interval <- group_by(activity, interval) %>% summarize(mean(steps, na.rm = TRUE))
```

Time period that contains maximum number of steps on average is 835
```{r}
steps_by_interval[which.max(steps_by_interval$`mean(steps, na.rm = TRUE)`),]
```


library(ggplot2)
ggplot(steps_by_interval, aes(x = interval, y = steps_by_interval$`mean(steps, na.rm = TRUE)`)) +
        geom_line() +
        labs(title = "Steps by interval") +
        ylab("Steps") +
        theme_bw()
````


## Imputing missing values

```{r,echo=TRUE}
sum(is.na(activity$steps))
```
NAs are 2304

Imputating NAs with the mean of each interval

```{r}
listmean <- as.list(steps_by_interval$`mean(steps, na.rm = TRUE)`)

db_no_NA <- activity %>% mutate(steps_mean = ifelse(is.na(steps), listmean, steps))
db_no_NA <- transform(db_no_NA, steps_mean = as.numeric(steps_mean))

ggplot(db_no_NA, aes(date, steps_mean)) +
        geom_col() +
        theme_light() +
        labs(x = "date", y = "steps by day", title = "steps by day (no NAs)")
```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
db_wd <- db_no_NA %>% mutate(weekday = ifelse(weekdays(date) == 
        "Sabato" | weekdays(date) == "Domenica", yes = "Weekend", no = "Weekday"))


ggplot(db_wd, aes(interval, steps)) +
        geom_line() +
        facet_grid(.~weekday) +
        labs()
```

The plot shows that there is a difference in pattern between weekday and weekend



