---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

I will first read in the data and show a summary of the information.  
```{r}
library(data.table)
data <- read.table("./activity.csv", header = T, sep = ",")
data$date <- as.Date(data$date, format = "%Y-%m-%d")
summary(data)
```

Making a histogram of the total number of steps taken each day
```{r}
library(ggplot2)
library(plyr)
library(dplyr)
total <- data[,1:2]
total <- total %>% group_by(date) %>% summarise(sum = sum(steps, na.rm = T))
g <- ggplot(total, aes(sum))
g + geom_histogram()
```


```{r, echo = F}
mean <- mean(total$sum)
median <- median(total$sum)
```
`r mean` and `r median` is the mean and median, respectively, steps taken each day.  

This is the average number of steps taken for each interval across all days. 

```{r}
average <- data[,c(1,3)]
average <- average %>% group_by(interval) %>% summarise(mean = mean(steps, na.rm = T))
g2 <- ggplot(average, aes(interval, mean))
g2 + geom_line()
```

```{r, echo= F}
max <- average$interval[which.max(average$mean)]
```
This interval with the maximum number of steps on average is `r max`. 

We are now inputting the mean steps for each interval into observations so that have missing values. 
```{r}
new <- data
for (i in average$interval) {
  new[new$interval == i & is.na(new$steps), ]$steps <- average$mean[average$interval == i]
}

summary(new)
```

We can make an updated historgram now that the missing values are filled in. 
```{r}
total2 <- new [,1:2]
total2 <- total2 %>% group_by(date) %>% summarise(sum = sum(steps, na.rm = T))
g3 <- ggplot(total2, aes(sum))
g3 + geom_histogram()
```

Now, we will subset the data into two datasets, a weekday data set and a weekend data set. We will then create plots to showcase the mean number of steps taken between Monday and Friday for each interval and the mean number of steps taken on Saturday and Sunday for each interval. 
```{r}
weekday <- new[wday(new$date) %in% 1:5,]
weekends <- new[wday(new$date) %in% 6:7,]

averageweekday <- weekday[,c(1,3)]
averageweekday <- averageweekday %>% group_by(interval) %>% summarise(mean = mean(steps, na.rm = T))

averageweekends <- weekends[,c(1,3)]
averageweekends <- averageweekends %>% group_by(interval) %>% summarise(mean = mean(steps, na.rm = T))

par(mfrow = c(2,1), mar = c(2,2,2,2))
with(averageweekday, plot(interval, mean, type = "l", main = "Weekday"))
with(averageweekends, plot(interval, mean, type = "l", main = "Weekend"))
```