# Peer Assessment 1 of Reproduce Research
========================================================

```{r}
library(lattice)
```

## Loading and preprocessing the data
```{r}
activity <- read.csv("activity.csv",header=TRUE,colClasses=c("integer","Date","integer"))
head(activity)
tail(activity)
```

## What is mean total number of steps taken per day?
- Make a histogram of the total number of steps taken each day
```{r}
totalStepsByDay <- sapply(with(activity,split(steps,date)),sum,na.rm=T)
hist(totalStepsByDay,xlab="Total Steps Per Day",main="Histogram of Total Steps Per Day")
```
- Calculate and report the mean and median total number of steps taken per day
```{r}
meanStepsByDay <- mean(totalStepsByDay)
medianStepsByDay <-median(totalStepsByDay)
```
## What is the average daily activity pattern?
- Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
avgStepsByInterval <- sapply(with(activity,split(steps,interval)),mean,na.rm=T)
intervals <- names(avgStepsByInterval)
plot(intervals,avgStepsByInterval,type="l")

names(which.max(avgStepsByInterval))
```
## Inputing missing values
- Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
sum(is.na(activity))
```
- Filling in all of the missing values in the dataset with the mean for that 5-minute interval.
- Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
missingIndices <- which(is.na(activity))
new_activity <- activity
for(i in missingIndices){
  if(is.na(activity[i,]$steps)){
    new_activity[i,]$steps = avgStepsByInterval[as.character(activity[i,]$interval)]
  }
}
```

- Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
```{r}
totalStepsByDayNew <- sapply(with(new_activity,split(steps,date)),sum,na.rm=T)
hist(totalStepsByDayNew,xlab="Total Steps Per Day",main="Histogram of Total Steps Per Day New")
newMeanStepsByDay <- mean(totalStepsByDayNew)
newMedianStepsByDay <-median(totalStepsByDayNew)
```
The new mean of steps is **`r newMeanStepsByDay `** and the old one is **`r meanStepsByDay `**  
The new median of steps is **`r newMedianStepsByDay `** and the old one is **`r medianStepsByDay `**

## Are there differences in activity patterns between weekdays and weekends?
- Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
new_activity$dayType <- with(new_activity,ifelse(weekdays(date)=="Saturday" | weekdays(date)=="Sunday","Weekend","Weekday"))
```

- Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
avgStepsByInterval_DT <- with(new_activity,aggregate(steps,by=list(interval,dayType),mean))
names(avgStepsByInterval_DT) <- c("interval","dayType","steps")
xyplot(steps~interval|dayType, data=avgStepsByInterval_DT,type="l",layout=c(1,2),ylab="Number of Steps")
```
