---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data
```{r echo=TRUE}
#import necessary libraries
library(data.table)
library(lubridate)
library(ggplot2)
library(lattice)


unzip('activity.zip')
data <- fread('activity.csv',header = TRUE)
data[,date:=as.Date(date)]
```


## What is mean total number of steps taken per day?

#### 1. Calculate the total number of steps taken per day
```{r echo=TRUE}
#select and sum steps where not equal NA
sumsOfSteps <- data[!is.na(steps),sum(steps),by=date]

#change variable name V1 to sumOfSteps
sumsOfSteps <- sumsOfSteps[,sumOfSteps:=as.numeric(V1)]
#remove V1 and display sumsOfSteps
sumsOfSteps[,V1:=NULL]
```

#### 2. Make a histogram of the total number of steps taken each day
```{r echo=TRUE}
#draw histogram
hist(sumsOfSteps[,sumOfSteps],breaks = 30,main='Sum of Steps per Day',xlab = 'Sum of Steps',col='red',density = 20)
```

#### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r echo=TRUE}
#calculate mean by ignoring NA values
meanOfSteps <- data[!is.na(steps),mean(steps),by=date]
meanOfSteps <- meanOfSteps[,mean:=V1]
#report mean per day
meanOfSteps[,V1:=NULL]

#calculate median by ignoring NA values
medianOfSteps <-  data[!is.na(steps),as.double(median(steps)),by=date]
medianOfSteps <- medianOfSteps[,median:=V1]
#report median per day
medianOfSteps[,V1:=NULL]
```


## What is the average daily activity pattern?
#### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r echo=TRUE}
avgPattern <- data[!is.na(steps),mean(steps),by=interval]
avgPattern <- avgPattern[,averageSteps:=V1]

#report avgPattern
avgPattern[,V1:=NULL]

#draw the plot
g <- ggplot(avgPattern,aes(y=averageSteps,x=interval))
g+geom_line(aes(colour=averageSteps))+scale_colour_gradient(low='blue')+ggtitle('Average Daily Activity Pattern')+xlab('interval (5-min)')+ylab('Average Steps (per interval)')
```

#### 2. The interval which has maximum mean value
```{r echo=TRUE}
#find maximum 
max <- avgPattern[,max(averageSteps)]
avgPattern[averageSteps==max]
```
On average across all the days in the dataset 5-min interval 835 contains the maximum number of steps with mean 206.17

## Imputing missing values
#### 1. Calculate and report the total number of missing values in the dataset
```{r echo=TRUE}
#calculate NA values 
data[is.na(steps),.N]
```
Total number of NA values is 2304 in the dataset.

#### 2&3. Filling in all of the missing values in the dataset 
 - The strategy for this part is filling NA values in with the mean for corresponding 5-minute interval
```{r echo=TRUE}
#change type of steps integer to double and copy data.table to a new data set
newDataSet <- copy(data[,steps := as.double(steps)]) 
#assign average steps by interval to NA values 
newDataSet[is.na(steps) & interval==avgPattern[,interval],steps:=avgPattern[,averageSteps]]
```
#### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r echo=TRUE}
newSumsOfSteps <- newDataSet[,sum(steps),by=date]
hist(newSumsOfSteps[,V1],breaks = 30,main='Sum of Steps per Day',xlab = 'Sum of Steps',col='blue',density = 20)
```
##### Mean of steps in imputed data set:
```{r echo=T}
newMeanOfSteps <- newDataSet[,mean(steps),by=date]
newMeanOfSteps <- newMeanOfSteps[,mean:=V1]
newMeanOfSteps[,V1:=NULL]
```
##### Median of steps in imputed data set:
```{r echo=T}
newMedianOfSteps <- newDataSet[,median(steps),by=date]
newMedianOfSteps <- newMedianOfSteps[,median:=V1]
newMedianOfSteps[,V1:=NULL]
```
As one can see the results from part1 vs part2 of sum, mean and median values differ in some ways:

 - First of all sums of steps increased dramatically because we just replaced 2304 NA values to corresponding mean by interval

 - Mean values doesn't changed for valid rows, but there are some days have only NA values. These days' mean change NA to 37.3825996
 
 - Median values stay as the same but again the days with only NA values changed to 34.11321

## Are there differences in activity patterns between weekdays and weekends?
#### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r echo=TRUE}
Sys.setlocale('LC_TIME','en_US')
newDataSet <- newDataSet[,day:='weekday']
newDataSet <- newDataSet[weekdays(date)=='Sunday' | weekdays(date)=='Saturday', day:='weekend']
newDataSet <- newDataSet[,mean:=mean(steps),keyby=interval]
newDataSet <- newDataSet[,day:=as.factor(day)]
```
#### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r echo=TRUE}
xyplot(mean~interval|day,newDataSet,type = "l",lwd=1,  layout = c(1, 2),xlab = "Interval", ylab = "Avg. Number of steps")
```

