---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

1) Load the data into R as a data frame and look at it.
```{r}
activity = read.csv('activity.csv')
str(activity)
```
2) Let's convert the date variable from a factor to a date.
```{r}
activity$date <- as.Date(activity$date)
str(activity)
summary(activity)
```

Make a new data frame for total daily steps.
```{r}
library(plyr)
dailysteps <- ddply(activity, .(date), summarise, totalsteps = sum(steps))
str(dailysteps)
summary(dailysteps)

```

## What is mean total number of steps taken per day?
1) Make a histogram of the total number of steps taken each
day.
```{r}
hist(dailysteps$totalsteps, main = "Histogram of the total
     number of steps taken each day", xlab="Steps")
```
2) Calculate the mean and median total steps per day.
```{r}
mean(dailysteps$totalsteps, na.rm=T)
median(dailysteps$totalsteps, na.rm=T)
```
The mean number of daily steps is 10766, and the median
number of daily steps is 10765.

## What is the average daily activity pattern?
Let's manipulate the activity data frame so that it is
easier to work with.
```{r}
dailyactivity <- ddply(activity, .(interval), summarise, avesteps = mean(steps,na.rm=T))
str(dailyactivity)
summary(dailyactivity)
```
1) Make a time series plot of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across
all days (y-axis).
```{r}
library(ggplot2)
myplot <- ggplot(dailyactivity, aes(x=interval, y=avesteps))
myplot <- myplot + geom_line()
myplot <- myplot + labs(x="5- minute interval", y = "Average number of steps", title = "Number of steps taken in 5-minute intervals, averaged across all days")
myplot
```
2) Determine which 5-minute interval contains the maximum number of steps.
```{r}
maxsteps <- which.max(dailyactivity$avesteps)
dailyactivity$avesteps[maxsteps]
dailyactivity$interval[maxsteps]
```
The interval number 835 contains the maximum number of steps.  That looks consistent with our plot.


## Imputing missing values
1) Find the total number of missing values in the dataset.

```{r}
summary(activity)
```
The steps variable is the only variable with missing values.
There are 2304 missing values in the dataset.

2) Devise a strategy to replace missing values.  I will replace the missing values with the average for that interval, since the pattern varies significantly over the
course of a day.  First I will merge the daily pattern 
data frame with the activity data frame.
```{r}
activity2 <- merge(activity, dailyactivity)
str(activity2)
summary(activity2)
```
3) Now create a new dataset equal to the original but with the
missing data filled in.
```{r}
activity2$steps[is.na(activity2$steps)] <- activity2$avesteps
activity2$avesteps <- NULL
str(activity2)
summary(activity2)
```
4) Make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps taken per day.
First I will make a new data frame for the total daily steps.
```{r}
dailysteps2 <- ddply(activity2, .(date), summarise, totalsteps = sum(steps))
str(dailysteps2)
summary(dailysteps2)
```
Now make the histogram.
```{r}
hist(dailysteps2$totalsteps, main = "Histogram of the total number of steps\n taken each day (imputed data)", 
    xlab = "Steps")
mean(dailysteps2$totalsteps)
median(dailysteps2$totalsteps)
```
Now the mean and median total steps are 9371 and 10395,
respectively.  Imputing missing data has changed these
estimates, and in this case, lowers both the mean and 
median, with a more significant effect on the mean.

## Are there differences in activity patterns between weekdays and weekends?
1) Create a new factor variable in the dataset with 
two levels "weekday" and "weekend."
```{r}
activity2$dayofweek <- weekdays(activity2$date) 
head(activity2)
activity2$dayofweek[activity2$dayofweek == 'Saturday' |
                    activity2$dayofweek == 'Sunday'] <- 'weekend'
activity2$dayofweek[activity2$dayofweek != 'weekend'] <- 'weekday'
activity2$dayofweek <- as.factor(activity2$dayofweek)
head(activity2)
str(activity2)
```

2) Make a panel plot containing a time series plot of the
5-minute interval (x-axis) and the average number of steps
taken, averaged across all weekday days or weekend days
(y-axis).

```{r}
activity3 <- ddply(activity2, .(interval,dayofweek), summarise, avesteps = mean(steps))
summary(activity3)
library(lattice)
xyplot(avesteps ~ interval | dayofweek, data=activity3,
       type='l',
       lwd=2,
       layout=c(1,2),
      ylab = 'Number of steps')
```
It does look like there are differences in activity
patterns between weekdays and weekends.  During weekdays,
there is a peak in activity early in the day (maybe when
people are walking to work) and then little activity 
the rest of the day.  During weekends, the activity is 
consistently slightly higher over the interval.
