---
title: "Reproducible Research: Peer Assessment 1"
author: "KW"
date: "December 20, 2015"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, results = "show")
```

## Loading and preprocessing the data
**Libraries**
```{r libraries}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(lubridate)
```

### Load data into dplyr data frame
```{r initload}
if (!file.exists("repdata-data-activity.zip")) {
        fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(fileUrl, destfile = "./activity.zip") }

data <- tbl_df(read.csv(unz("activity.zip", "activity.csv"), header = T))
```


## What is mean total number of steps taken per day?
**Group base data on days**
```{r groupbydays}
days <- group_by(data, date)
```

**Analyze steps over days**
```{r stepsoverday1}
sumsteps <- summarise(days, 
                      "meansteps" = mean(steps, na.rm = TRUE), 
                      "mediansteps" = median(steps, na.rm = TRUE), 
                      "totalsteps" = sum(steps, na.rm = TRUE))
```

### Histogram of daily steps
```{r unimputedhistogram}
unimphist <- ggplot(sumsteps, aes(x = totalsteps)) + 
        stat_bin(binwidth = 10) +
        geom_histogram(fill = "#880011", binwidth = diff(range(sumsteps$totalsteps))/30) +
        ggtitle("Steps/Day") +
        labs(x = "Steps")
unimphist
```

### Mean and median steps per day
```{r displaystepanalysis}
as.data.frame(select(sumsteps,1:3))
```


## What is the average daily activity pattern?
```{r activityplot}
intervals <- group_by(data, interval)
pp <- summarise(intervals, 
                "avgsteps" = mean(steps, na.rm = TRUE))

activityplot <- ggplot(pp, aes(interval, avgsteps)) + 
        geom_line(color = "blue4") +
        ggtitle("Average Steps Taken Per 5-Minute Interval") +
        labs(x = "Interval", y="Average Steps")
activityplot
```

### 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps
```{r maxstepint}
intsteps <- summarise(intervals, 
                      mean(steps, na.rm = TRUE))
colnames(intsteps) <- c("interval", "meansteps")
maxsteps <- as.integer(select(filter(intsteps, 
                                     meansteps == max(na.omit(meansteps))),
                              interval))
maxsteps
```

## Imputing missing values
### Number of missing values
```{r missingvals}
nrow(filter(data, is.na(steps)))
```

**New dataset filling missing values with interval mean**
```{r imputedvals}
dataf <- data %>% 
        group_by(interval) %>% 
        mutate(steps = ifelse(is.na(steps), 
                              as.integer(round(mean(steps, na.rm = TRUE))), 
                              steps))
```

**Replicate original dataset averages from imputed dataset**
```{r summaryimputed}
daysf <- group_by(dataf, date)
sumstepsf <- summarise(daysf, 
                       "meansteps" = mean(steps, na.rm = TRUE), 
                       "mediansteps" = median(steps, na.rm = TRUE), 
                       "totalsteps" = sum(steps, na.rm = TRUE))
```

**Prepare imputed histogram of daily steps**
```{r imputedhistogram}
imphist <- ggplot(data=sumstepsf, aes(x=totalsteps)) +
        geom_histogram(fill="#003366", 
                       binwidth = diff(range(sumstepsf$totalsteps))/30) +
        ggtitle("Steps/Day (Post-Imputing)") +
        labs(x="Steps")
```

### Difference between original and imputed datasets
The impact imputing had to the dataset reduced the number of days having 0 steps and increasing the frequency of steps per day near the 10,000 step per day mean.  

```{r panelhistograms}
grid.arrange(unimphist, imphist, nrow = 1)
```  

### Mean and median steps per day with imputed missing values
```{r imputedtable}
as.data.frame(sumstepsf,1:3)
```


## Are there differences in activity patterns between weekdays and weekends?

**Prepare dataset with weekday vs weekend levels over date**
```{r assigningdaytype}
dataj <- mutate(dataf, 
                "dayt" = { ifelse (wday(date) < 2 | wday(date) > 6, "weekend", "weekday") })

intj <- group_by(dataj, interval, dayt)
ppj <- summarise(intj, 
                 "avgsteps" = mean(steps, na.rm = TRUE))
```

**Prepare plot of weekend vs weekday average steps/interval**
```{r daytypecompare}
daytypeplot <- ggplot(ppj, aes(interval, avgsteps)) +
        geom_line(color = "blue4") +
        facet_wrap(~ dayt, ncol = 1) +
        ggtitle("Average Steps/Interval between Weekdays and Weekends")
```
### Average steps/interval comparing weekday and weekend levels
```{r weekpartplot}
daytypeplot
```