---
title: "Reproducible Research Project 1"
author: "A. van der Kuech"
date: "May 17, 2015"
output: html_document
---



```{r}
library(ggplot2)
library(lubridate)
library(plyr)
library(dplyr)
```

Downloading file, creating and setting a working directory

```{r}
if (!file.exists("peerAsgn1")) {
        dir.create("peerAsgn1")
        setwd("peerAsgn1")
        download.file(fileUrl, destfile = "./activity.zip", method = "curl")
        download.date = date()
        list.files() 
}
```


Processing Data

```{r}
activM = read.csv("./peerAsgn1/activity.csv", sep = ",")
head(activM, n = 3)
dim(activM)
summary(activM)
str(activM)
```

Mean and median of total number of steps taken per day

```{r}
totalSteps = tapply(activM$steps, activM$date, FUN = sum, na.rm = T)

mean(totalSteps)
median(totalSteps)
```

Calculate the number of steps taken per day

```{r}
steper = activM %>%
        filter(!is.na(steps)) %>%
        group_by(date) %>%
        summarize(steps = sum(steps)) %>%
        print
```

Histogram of number of steps taken each day

```{r}
ggplot(steper, aes(x = date, y = steps)) + geom_histogram(stat = "identity") + xlab("Dates") + ylab("Steps") + labs(title = "Total Number of Steps per Day")
```

Average daily activity pattern

```{r}
activP = activM %>%
        filter(!is.na(steps)) %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) %>%
        print

plot(activP$interval, activP$steps, type = "l", xlab = "Interval", ylab = "Steps", main = "Average Number of Steps Taken Across All Days", col = "magenta")
```

Maximum number of steps

```{r}
activP[which.max(activP$steps), ]$interval
```

Calculate the total missing values in the dataset

```{r}
totalNAs = sum(is.na(activM))
totalNAs
```

New "cloned" dataset with missing data filled in

```{r}
activM2 = activM %>%
        group_by(interval) %>%
        mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = T), steps))
summary(activM2)
```

Histogram of total number of steps per day

```{r}
activM2Group = activM2 %>%
        group_by(date) %>%
        summarize(steps = sum(steps)) %>%
        print

ggplot(activM2Group, aes(x = date, y = steps)) + geom_histogram(stat = "identity") + xlab("Dates") + ylab("Steps") + labs(title = "Total Number of Steps per Day")
```

Mean and median of total number of steps taken per day.

```{r}
sumSteps = tapply(activM2$steps, activM2$date, FUN = sum, na.rm = T)
mean(sumSteps)
median(sumSteps)
```

Do values differ?

```{r}
mean(totalSteps) == mean(sumSteps)

median(totalSteps) == median(sumSteps)

summary(totalSteps)

summary(sumSteps)
```


Impact of missing data on the estimates of the total daily number of steps. 
Estimates are increased by 41, 3041, 370, 1416, 0, 0.

```{r}
summary(sumSteps) - summary(totalSteps)


par(mfrow = c(2, 1))
hist(sumSteps, col = "magenta")
hist(totalSteps, col = "green")
```

Differences in activity patterns between weekdays and weekends

```{r}
dayWeek = function(date) {
        if (weekdays(as.Date(date)) %in% c("Saturday", "Sunday")) {
                "weekend"
        } 
        else {
                "weekday"
        }
}

activM2$dayType = as.factor(sapply(activM2$date, dayWeek))
```

Panel plot containing time series

```{r}
par(mfrow = c(2, 1))
for (type in c("weekend", "weekday")) {
        stepsType = aggregate(steps ~ interval, data = activM2, subset = activM2$dayType == type, FUN = mean)
        plot(stepsType, type = "l", main = type)
}
```

Session information:

```{r}
sessionInfo()
```




































