---
title: "Rep Research Assignment 1"
author: "Thomas Gannon"
date: "Saturday, April 16, 2016"
output: html_document
---

Load Libraries
```{r, warning = FALSE, message = FALSE}
library("ggplot2")
library("dplyr")
```

1. Code for reading in dataset and processing data
```{r, warning = FALSE, message = FALSE}
activity <- read.csv("activity.csv", header = TRUE, sep = ",", colClasses = c("numeric","character","numeric"))
activity$date <- as.Date(activity$date, format = "%m/%d/%Y")
```


2. Make a histogram of total steps per day
```{r, warning = FALSE, message = FALSE}

DailyActivity <- aggregate(steps ~ date, activity, sum)

#create histogram plot
ggplot(DailyActivity, aes(steps)) + 
  geom_histogram(color = "red", fill = "red") + 
  ggtitle("Distribution of total steps per day")

```

3. Mean and median # of steps per day
```{r, warning = FALSE, message = FALSE}
StepsMean <- mean(DailyActivity$steps)
StepsMedian <- median(DailyActivity$steps)
```

4. Time series plot of avg # of steps taken
```{r, warning = FALSE, message = FALSE}
DailyStepActivity <- aggregate(steps ~ interval, activity, mean)

#create plot
ggplot(DailyStepActivity, aes(interval, steps)) + 
  geom_line() + 
  ggtitle("Average daily activity pattern")  +
  labs(x = "interval",  y = "steps")
```

5. The five minute interval that on average contains max # of steps
```{r, warning = FALSE, message = FALSE}
MaxInterval <- which(DailyStepActivity$steps == max(DailyStepActivity$steps))
MaxInt <- DailyStepActivity[MaxInterval, 1]
MaxInt
```

6. Code to impute missing values
```{r, warning = FALSE, message = FALSE}
MissingValues <- sum(is.na(activity$steps))


RevisedActivity <- activity
RevisedActivity <- merge(activity, DailyStepActivity, by = "interval")
RevisedActivity$steps.x[is.na(RevisedActivity$steps.x)] <- RevisedActivity$steps.y
RevisedMissingValues <- sum(is.na(RevisedActivity$steps.x))

RevisedDailyActivity <- aggregate(steps.x ~ date, RevisedActivity, sum)
RevisedStepsMean <- mean(RevisedDailyActivity$steps.x)
RevisedStepsMedian <- median(RevisedDailyActivity$steps.x)
```

7. Histogram of total # of steps taken each day after missing values imputed
```{r, warning = FALSE, message = FALSE}
ggplot(RevisedDailyActivity, aes(steps.x)) + 
  geom_histogram(color = "green", fill = "green") + 
  ggtitle("Revised distribution of total steps per day")
```

8. Plot comparing avg steps taken per 5 minute interval across weekdays and weekends
```{r, warning = FALSE, message = FALSE}
WeekActivity <- RevisedActivity %>% 
  mutate(Weekday = weekdays(RevisedActivity$date)) %>%
  mutate(WeekdayClass = "Weekday")

weekend <- WeekActivity$Weekday %in% c("Saturday", "Sunday") 
WeekActivity$WeekdayClass[weekend == TRUE] <- "Weekend"

DailyStepActivityByWeekDayType <- aggregate(steps.x ~ interval + WeekdayClass, WeekActivity, mean)

#create plot
ggplot(DailyStepActivityByWeekDayType, aes(interval, steps.x)) + 
  geom_line() + 
  facet_wrap(~ WeekdayClass, nrow = 2, ncol = 1) +
  labs(x = "interval",  y = "steps")
```

9. All the R code need to reproduce results.  All code is contained above.