---
title: "PA1_template.Rmd"
author: "Emma Engler"
date: "2/24/2020"
output:
  html_document:
    df_print: paged
---

#Reproducible Research: Assignment 1 

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(knitr)
library(dplyr)
```

##Load and preprocess data 
###Loading data 
```{r, results='markup',warning=TRUE,message=TRUE}
if(!file.exists('activity.csv')){
  unzip('activity.zip')
}
data <-read.csv('activity.csv')
summary(data)
head(data)
```

## What is mean total number of steps taken per day? 
Missing values in the dataset are not used. 
```{r}
# Calculating the mean total number steps taken per day 
steps <-aggregate(data$steps, by=list(Date=data$date), FUN=sum)
library(ggplot2)
names(steps)[names(steps)=="x"] <-"Total"
temp <-as.Date(steps$Date, "%Y-%m-%d")
steps$Date <-format(temp, format = "%m-%d")
head(steps)
```
### Make a histogram of the total number of steps taken each day 
```{r}
hist1 <-ggplot(data=na.omit(steps), aes(Total)) +
  geom_histogram(binwidth=1500, colour="grey") +
  xlab("Total Steps Per Day") +
  ylab("Count") +
  ggtitle("Histogram of Total Steps Per Day")
print(hist1)
```
### Calculate and report mean and median of total number of steps per day 
```{r}
mean(na.omit(steps$Total))
median(na.omit(steps$Total))
```

##What is the average daily activity pattern? 
Time series plot of the average number of steps taken. Shows the 5-min interval that, on average, contains the max number of steps. 
```{r}
# Make a Time Series Plot 
steps_five <- aggregate(steps ~ interval, data=data, FUN=mean)
timeseries1 <-ggplot(data=steps_five, aes(x=interval, y=steps)) +
  geom_line() +
  xlab("Time Intervals (5 mins)") +
  ylab("Total Number of Steps") +
  ggtitle("Av. Number of Steps Per Interval")
print(timeseries1)
```

### Which interval contains max number of steps? 
```{r}
head(steps_five)
steps_five[which(steps_five$steps==max(steps_five$steps)),]
```
*Most steps: 'r most.steps'
## Imputing missing values 
There are many missing values in days/intervals. This introduces the possibility of bias into results or summaries. 

### Calculate and report total missing values in dataset 
```{r}
sapply(X=data, FUN=function(x) sum(is.na(x)))
```
*Number of missing values: 'r missing.values' 
### Devise a strategy to fill in all missing values in dataset 
```{r}
replace_bymean <-function(num) replace(num, is.na(num), mean(num, na.rm=TRUE))
day_mean <-(data %>% group_by(interval) %>% mutate(steps=replace_bymean(steps)))
head(day_mean)
sum(is.na(day_mean))
```

### Create new dataset equal to the original dataset but with the missing data filled in 
```{r}
new_data <-as.data.frame(day_mean)
head(new_data)
```
## Make a histogram of the total number of steps taken each day 
```{r}
new_steps <-aggregate(new_data$steps, by=list(new_data$date), FUN=sum)
names(new_steps)[names(new_steps)=="x"] <-"Total"
names(new_steps)[names(new_steps)=="Group1"] <-"Date"
hist2 <-ggplot(data=new_steps, aes(Total)) +
  geom_histogram(binwidth=1500, colour="grey") + 
  xlab("Total Steps Per Day") +
  ylab("Count") +
  ggtitle("Hist of Total Number of Steps Per Day-New Data")
print(hist2)
```

## Comparison of the two plots 
```{r}
library(grid) 
library(gridExtra)
grid.arrange(hist1, hist2, ncol=2)
```
## Comparing mean and median 
```{r}
mean(na.omit(steps$Total))
median(na.omit(steps$Total))
mean(new_steps$Total)
median(new_steps$Total)
```
While the means of the dataset have remained the same, the medians of each dataset are slightly changed. The new data version shows a larger than that of the original with the NA's included. 

## Comparing the average number of steps taken per 5-minue interval across weekdays and weekends 
```{r}
#Creating new variable in dataset of weekday and weekend 
new_data$WeekendorWeekday <-ifelse(weekdays(as.Date(new_data$date)) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "Weekday", "Weekend")
head(new_data)
```
### Making a panel plot to compare the average number of steps taken per interval across weekdays and weekends 
```{r}
new_data <-(new_data %>% group_by(interval, WeekendorWeekday) %>% summarise(Mean=mean(steps)))
ggplot(new_data, mapping=aes(x=interval, y=Mean)) +
  geom_line() +
  facet_grid(WeekendorWeekday~.) +
  xlab("Interval") + 
  ylab("Means of Steps") +
  ggtitle("Comparison of Av Number of Steps for Each Interval")
```

It can be seen that there is different patterns between weekdays and weekends. 
