---
<<<<<<< HEAD
title: "PA1_template.Rmd"
author: "Kim Hollis"
date: "January 11, 2016"
output: html_document
---

##Load R Packages

```{r}
library(plyr)
library(ggplot2)
library(lattice)
```

##Process/Tranform Data
```{r}
setwd("C:/Users/khollis/Documents/R/data/repdata_data_activity")
data_activity <- read.table("c:activity.csv", header=TRUE, sep=",")
sum_steps <- tapply(data_activity$steps, data_activity$date, sum)
notna <- data_activity[!is.na(data_activity['steps']),]
dailysteps <- ddply(notna, .(date), summarize, totalsteps=sum(steps))
```

##Histogram of Total Steps
```{r}
hist(sum_steps, breaks = 5, col="green", main="Daily Steps Taken", xlab = "Daily Steps", ylab="Frequency")
```

##Mean & Median Total Number of Steps Taken Per Day
```{r}
mean(dailysteps$totalsteps)
median(dailysteps$totalsteps)
```

##The Average Daily Acitivity Pattern
```{r}
avg_interval <- tapply(data_activity$steps, data_activity$interval, mean, na.rm = TRUE)
```

##Plot
```{r}
plot(row.names(avg_interval), avg_interval, type = "l", xlab = "5-min interval", ylab = "Average across all Days", main = "Average number of steps taken", col = "blue")
```

```{r}
max_steps <- which.max(avg_interval)
names(max_steps)

total_nas <- sum(is.na(data_activity))
total_nas
```

##Replace all NAs with the mean of the 5 minute interval
```{r}
average_steps <- aggregate(steps ~ interval, data = data_activity, FUN = mean)

replace_na <- numeric()
for (i in 1:nrow(data_activity)) {
  obs <- data_activity[i, ]
  if (is.na(obs$steps)) {
    steps <- subset(average_steps, interval == obs$interval)$steps
  } else {
    steps <- obs$steps
  }
  replace_na <- c(replace_na, steps)
}
```

##Create a dataset that has all NAs filled in
```{r}
all_activity <- data_activity
all_activity$steps <- replace_na
```

##Histogram of total steps (NAs have been filled in with the mean of 5 minute interval)
```{r}
total_steps <- aggregate(steps ~ date, data = all_activity, sum, na.rm = TRUE)
hist(total_steps$steps, main = "Total steps by day", xlab = "day", col = "green")
```

##Mean & Median
```{r}
mean(total_steps$steps)
median(total_steps$steps)
```

##Differences in Activity Patterns between weekdays and weekends
```{r}
data_activity$date <- as.Date(data_activity$date, "%Y-%m-%d")

# add a new column indicating day of the week 
data_activity$day <- weekdays(data_activity$date)

# add a new column called day type and initialize to weekday
data_activity$day_type <- c("weekday")

# If day is Saturday or Sunday, make day_type as weekend
for (i in 1:nrow(data_activity)){
    if (data_activity$day[i] == "Saturday" || data_activity$day[i] == "Sunday"){
        data_activity$day_type[i] <- "weekend"
      }
    }

# convert day_time from character to factor
data_activity$day_type <- as.factor(data_activity$day_type)

# aggregate steps as interval to get average number of steps in an interval across all days
interval_steps <- aggregate(steps ~ interval+day_type, data_activity, mean)

qplot(interval, steps, data=interval_steps, geom=c("line"), xlab="Interval", ylab="Number of steps", main="") + facet_wrap(~ day_type, ncol=1)
```

=======
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data



## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
>>>>>>> 80edf39c3bb508fee88e3394542f967dd3fd3270
