---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
global seetings

```{r setoptions, echo = TRUE }
```
Load R package libraries required to answer the questions in the project.
```{r loadRpackages}
library(dplyr)
library(Hmisc)
library(ggplot2)
```
## Loading and preprocessing the data
Load the raw data and process/transform the data into a format suitable to calculate
the total number of steps taken per day - NAs are not imputed from the original data
ad = activity data (steps)
ads = activity data (steps) sum over all days
```{r totalnumberofsteps}

ad <- read.table("activity.csv",sep = ",", header = TRUE)
ad1 <- split(ad, ad$date)
ad2 <- split(ad, ad$interval)
ads <- sapply(ad1, function(x) {colSums(x["steps"])})
ads <- as.data.frame(ads)
names(ads) <- c("steps")
rownames(ads) <- c(1:61)
hist(ads$steps, col = "green", breaks = 10,xlab = "steps each day",main = title("Histogram of the Total number of Steps Taken Each Day"))
```
## What is mean total number of steps taken per day?
md = daily mean
```{ r dailyaveragesteps}
md <- sapply(ad1, function(x) {colMeans(x["steps"])})
md <- as.data.frame(md)
names(md) <- c("steps")
rownames(md) <- c(1:61)
md <- mutate(md, day = c(1:61))
```
print mean and media

```{r showtable, results = 'asis'}
library(xtable)
xt <- xtable(md)
print(xt, type = "html")
```

## What is the average daily activity pattern?
split the original data based on time intervals
compute the 5-minutes interval average steps over all days
mi = mean over 5-minutes interval
plot average the 5-minute interval (x-axis) vs. the dialy mean (y-axis)
```{r computeMean }

mi <- sapply(ad2, function(x) {colMeans(x["steps"], na.rm = TRUE)})
mi <- as.data.frame(mi)
names(mi) <- c("steps")
rownames(mi) <- c(1:288)
mi <- mutate(mi, intervalID = c(1:288))

g <- ggplot(mi, aes(intervalID))
g <- g + geom_line(aes(y = steps, color = "blue"))
#g <- ggplot(mi, aes(x = intervalID, y = steps))
#g <- g + geom_line()
g <- g + labs(x = "5-minutes interval ID", y = "average steps")
g <- g + labs(title = " Time Series Plot Average Daily Activity")
g
```

compute the 5-minutes interval where maximum activity  is
```{r maxInterval}
maxintervalID <- which.max(mi$steps)
```
the maximum 5-minutes interval ID `r maxintervalID`.

## Impute Misssing Values

To find the numbers of NAs use the following simple functions
```{r missingvlaues}
stat <- summary(ad$steps)

```
The number of NAs = `r stat[7]`

To impute missing values use the Hmisc package - imputed with mean value
adim - daialy activity (steps) from imputed data
adims - total steps over all days
```{r meanwithimputeddata}
ad$imputed <- with(ad, impute(steps, mean))
adim <- select(ad, date, imputed, interval)
names(adim) <- c("date", "steps", "interval")

# split the original data based on dates
adim1 <- split(adim, adim$date)
adims <- sapply(adim1, function(x) {colSums(x["steps"])})
adims <- as.data.frame(adims)
names(adims) <- c("steps")
rownames(adims) <- c(1:61)
```
draw histogram of the daily total steps
```{r histograpm, fig.height = 5}
hist(adims$steps, col = "green", breaks = 10,xlab = "steps each day", main = title("Histogram of the Total number of Steps Taken Each Day with Imputed NAs"))
```
compute  and report the mean and median of the total number of 
steps taken per day from imputed data set - 
imd = mean over days from imputed data
```{r imputedmean}
imd <- sapply(adim1, function(x) {colMeans(x["steps"])})
imd <- as.data.frame(imd)
names(imd) <- c("steps")
rownames(imd) <- c(1:61)
imd <- mutate(imd, day = c(1:61))
```
Here is a table of mean values from imputed data

```{r showtable2, results = 'asis'}
library(xtable)
xt <- xtable(imd)
print(xt, type = "html")
```
## Are there differences in activity patterns between weekdays and weekends?
Check differences in activity patterns between weekdays and weekends?
using imputed data set adim
first create separe dat ser froweekdays and weekend
```{r weekdays-ends}
adim$date <- as.Date((adim$date))
notweekend <- c("Monday","Tuesday", "Wednesday", "Thursday", "Friday") 
adim$wday <- factor((weekdays(adim$date) %in% notweekend), levels = c(FALSE, TRUE),
                    labels = c("weekend", "weekday"))
adweekdays <- filter(adim, adim$wday == "weekday")
adweekends <- filter(adim, adim$wday == "weekend")
```
split the original data based on time intervals for both weekdays and weekends
awd = weekdays activity
awdm - weekdays activity mean
awe = weekends activity
awem - weekends activity mean
```{r weekday-endssmean}
awd <- split(adweekdays, adweekdays$interval)
awdm <- sapply(awd, function(x) {colMeans(x["steps"])})
awdm <- as.data.frame(awdm)
names(awdm) <- c("steps")
rownames(awdm) <- c(1:288)
awdm <- mutate(awdm, intervalID = c(1:288))

awe <- split(adweekends, adweekends$interval)
awem <- sapply(awe, function(x) {colMeans(x["steps"])})
awem <- as.data.frame(awem)
names(awem) <- c("steps")
rownames(awem) <- c(1:288)
awem <- mutate(awem, intervalID = c(1:288))
```
merged the weekend and the weekdays data
mrgm - data frame containing all weekends andweekdays mean values
```{r mergedmean}
mrgm <- merge(awdm, awem, by = "intervalID")
names(mrgm) <- c("intervalID", "steps.weekdays", "steps.weekends")
```
plot the weekedn and the weekdays data on the same graph to show differences 
in activities
here is the line graph with contrasting weekdays and weekends daily activities
```{r linegraph, fig.height = 5}
g <- ggplot(mrgm, aes(intervalID))
g <- g + geom_line(aes(y = steps.weekdays, color = "red"))
g <- g + geom_line(aes(y = steps.weekends, color = "blue"))
g <- g + labs(x = "5-minutes interval ID", y = "average daily steps")
g <- g + labs(title = " Time Series Plot Average Daily Activity red: weekdays, blue:weekend")
g
```

The end

             
