---
title: "Peer Assessment 1, Reproducable Research"
author: "redson"
date: "2023-10-09"
output: html_document
---

The variables used:

1. **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA ) 
2. **date**: The date on which the measurement was taken in YYYY-MM-DD format 
3. **interval**: Identifier for the 5-minute interval in which measurement was taken

```{r}
#Loading the data
dat <- read.csv("activity.csv", header = T)
names(dat)
str(dat)
head(dat)
```


###Histogram, Mean & Median

---

```{r}
library(data.table)
table1 = data.table(dat)
table1_summary = table1[, list(total_steps = sum(steps, na.rm = T)), 
                          by = date]
```

**Drawind the Histogram**


```{r}
hist_it = function(x, title){
        hist(x, 
             breaks = 20,
             main = title,
             xlab = 'Total Number of Steps', col = 'grey',
            
             cex.main = .9)
        
        #caluclate mean and median
        mean_value = round(mean(x), 1)
        median_value = round(median(x), 1)
        
        #lines for mean and median on histogram
        abline(v=mean_value, lwd = 3, col = 'blue')
        abline(v=median_value, lwd = 3, col = 'red')
        
        #legend
        legend('topright', lty = 1, lwd = 3, col = c("blue", "red"),
               cex = .8, 
               legend = c(paste('Mean: ', mean_value),
               paste('Median: ', median_value))
               )
}

hist_it(table1_summary$total_steps, 'Number of Steps Taken Per Day')
```




##Daily Activity Pattern

----------------------------


```{r}
#summarize dataset by interval
summary_intv = table1[, list(avg_steps = mean(steps, na.rm = T)), 
                          by = interval]
#plot the time series
with(summary_intv, {
        plot(interval, avg_steps, type = 'l',
             main = 'Average Steps by Time Interval',
             xlab = '5 Minute Time Interval',
             ylab = 'Average Number of Steps')
        })
#Find Interval That Has The Maximum Avg Steps
max_steps = summary_intv[which.max(avg_steps), ]

#Generate Label String
max_lab = paste('Maximum Of ', round(max_steps$avg_steps, 1), ' Steps \n On ', max_steps$interval, 'th Time Interval', sep = '')

#Collect Cooridinates of The Max Interval For Graphing
points(max_steps$interval,  max_steps$avg_steps, col = 'red', lwd = 3, pch = 19)

#Add Label To Annotate Maximum # Steps And Interval
legend("topright",
       legend = max_lab,
       text.col = 'red',
       bty = 'n'
       )

```

---

###Missing Values

1. Calculate & Report The Number of Missing Values
```{r}
sum(is.na(dat$steps))
```

#Filling missing data
```{r}
setkey(dat_tbl, interval)
setkey(summary_intv, interval)


#Create function that will return the second value if the first value is NA
NA_replace = function(x,y){
        if(is.na(x)){
                
                return(y)
        }
        return(x)
}

#create new dataset that replaces NAs with average values
dat_tbl_miss = dat_tbl[dat_tbl_summary_intv]
dat_tbl_miss$new_steps = mapply(NA_replace,dat_tbl_miss$steps, dat_tbl_miss$avg_steps)

#summaryize new dataset by day
dat_tbl_summary_miss = dat_tbl_miss[, list(new_steps = sum(new_steps, na.rm = T)), 
                          by = date]
#preview new dataset
head(dat_tbl_summary_miss)
```

4.  Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

**Note: Mean and Median Are Reported In Legend Of Histogram**

```{r}

gen_hist(dat_tbl_summary$total_steps, 'Missing Values Removed')
gen_hist(dat_tbl_summary_miss$new_steps, 'Missing Values Replaced With \n Mean For Interval')

```

**Answer To Question:**
The mean and the median are now almost the same after replacing missing values with the mean value for the relevant interval. It makes sense that the median value would now move closer to the mean. So the Median value increased after this method of missing value replacement.

###Are there differences in activity patterns between weekdays and weekends?

---
1.  Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
#Make Function To Return Either "Weekday" or "Weekend"
weekpart = function(x){
        if(x %in% c('Saturday', 'Sunday')){
                return('Weekend')
        }
        
        return('Weekday')
}

#Add Name of Week
dat_tbl_miss$dayname = weekdays(as.Date(dat_tbl_miss$date))

#Add Factor Variable To Differentiate Weekday and Weekend
dat_tbl_miss$daytype = as.factor(apply(as.matrix(dat_tbl_miss$dayname), 1, weekpart))

#Summarize Dataset: Mean grouped by interval and daytype
dat_tbl_summary_miss = dat_tbl_miss[, list(avg_steps = mean(new_steps, na.rm = T)), 
                          by = list(interval, daytype)]

#inspect dataset
str(dat_tbl_summary_miss)
```
panel plot:
```{r}
library(lattice)
xyplot(avg_steps~interval | daytype, data = dat_tbl_summary_miss,
      type = 'l',
      xlab = 'Interval',
      ylab = 'Number of Steps',
      layout = c(1,2))
```

