---
title: "Reproducible Research - Project 1"
author: "Dan Schreck"
date: "June 13, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Summary

As part of the Coursera Reproducible Research Project 1 the object of this report
is to read data from an anonymous individual consisting of the number of steps 
taken in 5 minute intervals each day over a two month period and document the 
analysis in R Markdown.


#Data Set Fields
steps - number of recorded steps over the interval, which is every 5 minutes
date - date recorded in YYYY-MM-DD
interval - time period (midnight=0 through 11:55pm/23:55)



##Loading and preprocessing the data

1. Load the data (i.e. read.csv())
2. Process/transform the data (if necessary) into a format suitable for your analysis

Download dataset file <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip> 
and unzip it.


```{r}
setwd("/Users/schre/DataScience/wd/repro_research/Project1")
if(!file.exists("./data")){dir.create("./data")}
if (!file.exists('data/activity.zip')) {
      fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
      download.file(fileurl, 'data/activity.zip')
      unzip('data/activty.zip', exdir='./data')
}
activity <- read.csv("data/activity.csv")
```




##What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day
```{r}
#Total steps per day:
total.steps <- aggregate(steps ~ date, data=activity, na.rm=TRUE, sum)
```

2. Make a histogram of the total number of steps taken each day
```{r}
#Total steps per day:
plot(as.Date(total.steps$date), total.steps$steps, type="h", main="Steps per Day", xlab="", ylab="# Steps per Day", col="#0033FF", lwd=5)
abline(h=mean(total.steps$steps, na.rm=TRUE), col="#FF3333", lwd=3)
text(as.Date("2012-11-29"),mean(total.steps$steps)+700," Mean", col="#FF3333")
```      

3. Calculate and report the mean and median of the total number of steps taken per day

Mean=
```{r}
mean(total.steps$steps)
``` 

Median=
```{r}
median(total.steps$steps)
``` 



##What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and 
      the average number of steps taken, averaged across all days (y-axis)
```{r}
avgStepsDay <- tapply(activity$steps, activity$interval, mean, na.rm=T, simplify=T)
plot(as.integer(names(avgStepsDay)), avgStepsDay, type="l", main="Daily Activity Pattern\nAvg Steps per 5 mins", xlab="Time of Day\n 0=midnight to 2355 (11:55pm)", ylab="# Steps", col="#EE5511", lwd=2)
```     
      
2. Which 5-minute interval, on average across all the days in the dataset, contains 
      the maximum number of steps?
```{r}
mean.steps.interval <- data.frame(interval=as.integer(names(avgStepsDay)), average=avgStepsDay)
max.steps <- max(mean.steps.interval$average)
mean.steps.interval[mean.steps.interval$average == max.steps, ]
``` 

We can concluded from this on average the maximum number of steps (206.2) is taken
at 8:35am.



##Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. 
      the total number of rows with NAs)
```{r}
sum(is.na(activity$steps))
``` 

2. Devise a strategy for filling in all of the missing values in the dataset. 
      The strategy does not need to be sophisticated. For example, you could use 
      the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r}      
#populate NA values with the mean for that interval
activity.na.removed <- activity
# get index of NA values
na.index <- which(is.na(activity$steps)==TRUE)  
interval.avg <- tapply(activity$steps, activity$interval, mean, na.rm=TRUE, simplify=T)
 # replace all na values with the mean for that interval
activity.na.removed$steps[na.index] <- interval.avg[as.character(activity$interval[na.index])]
``` 

3. Create a new dataset that is equal to the original dataset but with the missing 
      data filled in.
      
4. Make a histogram of the total number of steps taken each day and Calculate and 
      report the mean and median total number of steps taken per day. Do these values 
      differ from the estimates from the first part of the assignment? What is the 
      impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
#Total steps per day:
total.steps.na.agg <- aggregate(steps ~ date, data=activity.na.removed, na.rm=TRUE, sum)
plot(as.Date(total.steps.na.agg$date), total.steps.na.agg$steps, type="h", main="Steps per Day\n(NA values replaced with average for that time period)", xlab="", ylab="# Steps per Day", col="#0033FF", lwd=5)
abline(h=mean(total.steps.na.agg$steps, na.rm=TRUE), col="#FF3333", lwd=3)
text(as.Date("2012-11-29"),mean(total.steps.na.agg$steps)+700,"  Mean", col="#FF3333")
```      


Mean=
```{r}
mean(total.steps.na.agg$steps)
``` 

Median=
```{r}
median(total.steps.na.agg$steps)
``` 




##Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset 
with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels - "weekday" and 
      "weekend" indicating whether a given date is a weekday or weekend day.
```{r}    
total.steps.day <- activity.na.removed
weekend <- c("Saturday","Sunday")
total.steps.day["day"] <-  factor(weekdays(as.Date(total.steps.day$date)) %in% weekend, levels=c(TRUE,FALSE), labels=c('weekend', 'weekday'))
``` 

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute 
      interval (x-axis) and the average number of steps taken, averaged across all 
      weekday days or weekend days (y-axis). See the README file in the GitHub repository 
      to see an example of what this plot should look like using simulated data.
```{r}      
library(lattice)
avg.step.day <- aggregate(steps ~ day+interval, data=total.steps.day, mean)
xyplot(steps ~ interval | factor(day),
       layout = c(2, 1),
       ylab="# Steps",
       xlab="Interval",
       type="l",
       lty=1,
       data=total.steps.day)
      
``` 

Based on this analysis we can conclude that weekdays generally see higher activity than weekends.