---
title: "Reproducible Research project 1"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

#Load all packages 
library(knitr)
library(dplyr)
library(ggplot2)

#Load data
fileURL= "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
activityFile="activity.zip"
download.file(url=fileURL,destfile=activityFile)
unzip(activityFile)
activity_row =read.csv("activity.csv")

#Process data
activity = activity_row[with(activity_row,{!(is.na(steps))}),]
```

What is mean total number of steps taken per day?
```{r}
by_day =group_by(activity,date)
steps_by_day = summarise(by_day,total=sum(steps))
head(steps_by_day)
hist(steps_by_day$total,main="Histogram of total number of steps taken per day", xlab = "Total no. of steps in a day")
summary(steps_by_day)
```

What is the average daily activity pattern?
```{r}
steps_by_interval = aggregate(steps~interval,activity,mean)
plot(steps_by_interval$interval,steps_by_interval$steps,type='l',main="Average number of steps over all days", xlab="Interval", ylab="Average number of steps")
max_steps_row <- which.max(steps_by_interval$steps)
steps_by_interval[max_steps_row, ]
```


Inputing missing values
Calculate and report the total number of missing values in the dataset
```{r}
sum(is.na(activity_row))
```

Filling in all of the missing values in dataset
```{r}
activity_imputed=activity_row
for(i in 1:nrow(activity_imputed)){
  if(is.na(activity_imputed$steps[i])){
    interval_value=activity_imputed$interval[i]
    steps_value=steps_by_interval[
      steps_by_interval$interval ==interval_value,]
    activity_imputed$steps[i]=steps_value$steps
  }
}
```
Histogram of the total no. of steps taken each day 
```{r}
df_imputed_steps_by_day = aggregate(steps~date,activity_imputed,sum)
hist(df_imputed_steps_by_day$steps,main="Histogram of total no. of steps per day(Imputed)",
     xlab = "Total no. of steps in a day")
mean(df_imputed_steps_by_day$steps)
median(df_imputed_steps_by_day$steps)
mean(steps_by_day$total)
median(steps_by_day$total)
```
Mean values stay the same but a slight difference in the median value

Are there differences in activity patterns between weekdays and weekends?
```{r}
activity_imputed['type_of_day']= weekdays(as.Date(activity_imputed$date))
activity_imputed$type_of_day[activity_imputed$type_of_day  %in% c('Saturday','Sunday') ] <- "weekend"
activity_imputed$type_of_day[activity_imputed$type_of_day != "weekend"]="weekday"
activity_imputed$type_of_day = as.factor(activity_imputed$type_of_day)
df_imputed_steps_by_interval= aggregate(steps~interval+type_of_day,activity_imputed,mean)
qplot(interval,steps,data=df_imputed_steps_by_interval,type='l',
geom=c('line'),xlab="Interval",ylab="Number of steps",
main="")+
facet_wrap(~type_of_day,ncol=1)
```
