---
title: "Reproducible Research: Peer Project 1"
output:
    html_document:
        keep_md: true
---
Packages needed to run code in this report are knitr, ggplot2, and lattice.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro to Project
This project is for the Reproducible Research course, a Coursera Data Science 
program through John Hopkins University. More information about this program is
available [here](https://www.coursera.org/specializations/jhu-data-science).

The data for our project comes from the data accumulated through a personal 
activity monitoring device collected for a single individual during October and
November 2012. The data includes the number of steps taken in five minute
intervals each day.

## Loading and Preprocessing the Data
* Download the data from the github account.

```{r Getting the Data}
## Set working directory.
## Then download data.
URL <- "https://github.com/VickieBailey/RepData_PeerAssessment1/blob/master/activity.zip?raw=true"
temp <- tempfile()
download.file(URL, temp)
ActMonitorData <- read.csv("activity.csv")
dateDownloaded <- date()
unlink(temp)
```

* A Little Look Around
    + Data Structure
```{r Structure of data}
str(ActMonitorData)
```

    +Summary of the Variables
```{r Summary of Variables}
summary(ActMonitorData)
```

    + Information on NAs. From summary, we see NAs only occur in steps column.
```{r Information on NAs }
## How many NAs
NAcount <- sum(is.na(ActMonitorData$steps))
## The percent of NAs as a decimal
evalNA <- mean(is.na(ActMonitorData$steps))
## Convert the decimal to a percent
percentNA <- format(evalNA*100, digits = 3)
```
There are `r NAcount` NAs in the step variable.
This represents a `r percentNA`%.

* Process the data. Make a copy of the raw dataframe to use for processing.
```{r New Dataframe and Date Class}
## Create a copy of the raw dataframe.
process1 <- ActMonitorData
process1$date <- as.Date(process1$date, "%Y-%m-%d")
```

## What is the mean total of steps taken per day?

* Calculate the number of steps taken per day.
```{r Steps per day, , results = 'asis'}
perday1 <- aggregate(steps ~ date, process1, FUN = sum)
library(knitr)
kable(perday1, caption = "Mean Total Steps Per Day")
```



* Create a histogram of the total number of steps taken each day.
```{r Number of Steps per Day, fig.height = 4}
## Set up and create histogram. Add a rug to show number of actual steps.
par(mar = c(5, 4, 1, 1))
with(perday1, {
 hist(perday1$steps, col = "deepskyblue", 
      main = "Total Steps per Day (excluding NAs)",
      xlab = "Number of Steps", ylab = "Count of Days" , las = 1)
rug(perday1$steps)   
})
```

* Find the mean and median number of steps taken each day.
```{r Mean and Median}
meansteps1 <- mean(perday1$steps)
mediansteps1 <- median(perday1$steps)
```
The mean steps per day without imputing of NAs is `r meansteps1`.
The median steps per day without imputing of NAs is `r mediansteps1`.


## What is the average daily activity pattern?
* Make a time series plot of the 5-minute interval (x-axis) and the average 
number of steps taken, averaged across all days (y-axis).
```{r Time Series Plot 1}
perinterval1 <- aggregate(steps ~ interval, data = process1, FUN = mean)
library(ggplot2)
par(mar = c(5, 4, 1, 1))
perintplot <- ggplot(perinterval1, aes(interval, steps)) +
    geom_line(color = "purple") +
    labs(x = "Interval", y = "Average Number of Steps",
         title = "Average Steps per Interval")
perintplot
```

* Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?
```{r Max Steps Interval}
## Find the maximum average steps.
MaxSteps <- max(perinterval1$steps)
## Subset the interval and max average steps.
submax <- subset(perinterval1, steps == MaxSteps)
## Pull out the interval.
maxinterval <- submax$interval
```
The 5-minute interval, on average across all days in the dataset, that contains 
the maximum average number of steps is `r maxinterval`.


## Imputing Missing Values
* Information on NAs. From original summary, we see NAs only occur in steps column.
```{r Count and Percent of NAs}
NAcount <- mean(is.na(process1$steps))
percentNA <- format(evalNA*100, digits = 3)
```
There are `r NAcount` NAs in the step variable.
This represents a `r percentNA`%.

* Impute missing values based on average of steps for the day of week and interval.
```{r Give values to NAs}
## Copy of process1 datafram.
process2 <- process1
## Create variable for day of week.
process2$day <- weekdays(as.Date(process2$date))
## Impute NAs by using average of steps based on interval and day of week.
process2$steps <- with(process2, ave(steps, day, interval,
                           FUN = function(x) replace(x, is.na(x), 
                                                     mean(x, na.rm = TRUE))))
```

* Create a histogram of the total number of steps taken each day.
```{r Number of Steps per Day, fig.height = 4}
## Compute number of steps per day.
perday2 <- aggregate(steps ~ date, process2, FUN = sum)
## Set up and create histogram. Add a rug to show number of actual steps.
par(mar = c(5, 4, 1, 1))
with(perday2, {
 hist(perday2$steps, col = "green2", 
      main = "Total Steps per Day with Imputing of NAs", 
      xlab = "Number of Steps", ylab = "Count of Days" , las = 1)
rug(perday2$steps)   
})
```

* Find the mean and median number of steps taken each day.
```{r Mean and Median}
meansteps2 <- mean(perday2$steps)
mediansteps2 <- median(perday2$steps)
```
The mean steps per day with imputing of NAs is `r meansteps2`.
The median steps per day with imputing of NAs is `r mediansteps2`.

* These values do differ from the estimates from the first part of the assignment.
``` {r Impact of Imputing}
meandif <- meansteps2 - meansteps1
meandif2 <- abs(meandif)
meanresult <- if(meandif > 0) {
    "increased"
}  else if (meandif < 0) {
    "decreased"
}  else {
    "did not change"
}
mediandif <- mediansteps2 - mediansteps1
mediandif2 <- abs(mediandif)
medianresult <- if(mediandif > 0) {
    "increased"
}  else if (mediandif < 0) {
    "decreased"
}  else {
    "did not change"
}
```
By imputing missing data, the estimates of the mean daily number of 
steps `r meanresult` by `r meandif2`.
The estimates for the median daily number of steps `r medianresult` by `r mediandif2`.


## Are there differences in activity patterns between weekdays and weekends?
* Create a new factor variable in the dataset with two levels - "weekday" and 
"weekend" indicating whether a given date is a weekday day or weekend day.
``` {r Weekday or Weekend}
weekends <- c('Saturday', 'Sunday')
process2$dayorend <- factor((weekdays(process2$date) %in% weekends), 
         levels=c(FALSE, TRUE), labels=c('weekday', 'weekend'))
```

* Make a panel plot containing a time series plot of the 5-minute interval and 
the average number of steps taken, averaged across all weekday days or weekend days.
``` {r Time Series Plot 2}
par(mar = c(5, 4, 1, 1))
panelplot <- xyplot(steps ~ interval | dayorend, data = process2, 
       layout = c(1,2),
       xlab = "Interval", 
       ylab = "Number of Steps",
       ylim = c(-25, 250),
       panel = function(x, y, ...) {
           panel.average(x, y, type = "l", horizontal = FALSE, col = "blue")
       }
)
panelplot
```

