---
title: "RR Project1"
author: "Shengbai Zhang"
date: "7/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load and Preprocessing 

```{r}
activity <- read.csv("~/Downloads/activity.csv")
head(activity)
dim(activity)
summary(activity)
activity$date<-as.Date(activity$date,"%Y-%m-%d")
```

## Mean Total Steps Taken Per Day

```{r, echo=TRUE}
a1<- data.frame(tapply(activity$steps,activity$date,sum,na.rm = T))
a1$date<-rownames(a1)
rownames(a1)<-NULL
library(ggplot2)
plot1<-ggplot(a1,aes(x=a1$date,y=a1$tapply.activity.steps..activity.date..sum..na.rm...T.))+geom_bar(stat="identity")+ylab("Total Steps")+xlab("Date")+ggtitle("Steps per Day")
plot1
mean(a1[,1])
median(a1[,1])
```

## Average Daily Pattern

```{r,echo=TRUE}
a2<-data.frame(tapply(activity$steps,activity$date,mean,na.rm=T))
a2$date<-rownames(a2)
rownames(a2)<-NULL
a2$average<-a2$tapply.activity.steps..activity.date..mean..na.rm...T.
ggplot(a2,aes(x=date,y=average))+ylab("Average Steps")+xlab("Date")+ggtitle("Average Steps per Day")+ geom_bar(stat="identity")
activity[which.max(activity$steps),]
```
2012-11-27 6:15 contain max steps

## Imputing Missing Data
The presence of missing days may introduce bias into some calculations or summaries of the data.

And the way to solve this problem is to impute values. Common imputations used include constant, regression model output, or mean value.

For simplicity, mean imputation will be used.
```{r,echo=TRUE}
a3<-activity
a3$missing<-is.na(a3$steps)
a3<-aggregate(data=a3, missing~date+interval,FUN = "sum")
a31<-data.frame(tapply(a3$missing,a3$date,sum))
a31$date<-rownames(a31)
rownames(a31)=NULL
names(a31)<-c("missing","date")
```
```{r}
a32<-data.frame(tapply(a3$missing,a3$interval,sum))
a32$date<-rownames(a32)
rownames(a32)<-NULL
names(a32)<-c("missing","Interval")
```
plots
```{r,echo=TRUE}
ggplot(a31,aes(x=date,y=missing))+ggtitle("Missing Value Distribution by Date")+
  geom_bar(stat = "identity")
```
From the plot, we could observe that there are 8 days that have no steps value, 
we donot know what happened at those days, but there is a pattern, and the mean 
imputation is desirable


