---
title: "Week2 Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Code for reading in the dataset and/or processing the data 

```{r} 
data <- read.csv(unz("activity.zip", "activity.csv"))
ChartValue <- complete.cases(data) 

# number of records with NA 
MissingValues <- length(ChartValue[ChartValue = FALSE])  

# number of records
AllValues <- length(ChartValue[ChartValue = TRUE])  
```

2. Histogram of the total number of steps taken each day

```{r}
# build a subset of the values
completes <- subset(data, ChartValue = TRUE)              

# split the values by date
StepsByDay <- split(completes, completes$date, drop = TRUE) 

# build a numeric vector all steps in a day  
DaySteps <- sapply(StepsByDay, function(x) sum(x$steps)) 

# plot a histogram
hist(DaySteps, main="Hist Total Steps per Day", xlab="# Steps", col="bisque3") 
```

3. Mean and median number of steps taken each day

```{r}
# print summary (includes mean & median)
summary(DaySteps)
```

4. Time series plot of the average number of steps taken

```{r}
# Split up the data for intervals
intervalsSplit <- split(data$steps, data$interval)

# Find the average per time interval - ignore NA values
StepsPerInterval <- sapply(intervalsSplit, mean, na.rm=TRUE)

uniqueIntervals <- unique(data$interval)

# Plot the time-series graph
plot(uniqueIntervals, StepsPerInterval, type="l",
     main="Average number of steps ", 
     xlab="Interval INDEX", ylab="Average # of steps", 
     lwd=2, col="blue")

# Find the location of where the maximum is
maxIntervalDays <- max(StepsPerInterval, na.rm=TRUE)
maxIndex <- as.numeric(which(StepsPerInterval == maxIntervalDays))

# Plot a vertical line where the max is
maxInterval <- uniqueIntervals[maxIndex]
abline(v=maxInterval, col="red", lwd=3)    

```

5. The 5-minute interval that, on average, contains the maximum number of steps

```{r}
# The 5-minute interval 
names(which.max(StepsPerInterval))

# that contains the number of steps
round(max(StepsPerInterval))

# at the position 
which.max(StepsPerInterval) 
```

6. Code to describe and show a strategy for imputing missing data

```{r}
# Missing values
maxInterval
isna<- is.na(data$steps)
sum(isna)

# Remove NaN values and replace with 0.  
# NaN values are produced when the entire day was filled with NA values
meanDaySteps <- sapply(intervalsSplit, mean, na.rm=TRUE)
meanDaySteps[is.nan(meanDaySteps)] <- 0

# Now create a replicated vector 288 times
# Now, all we have to do is find where in the data set there are missing steps, 
# and simply do a copy from one vector to the other
meanColumn <- rep(meanDaySteps, 288)

# The steps before replacement
rawSteps <- data$steps

# Find any values that are NA in the raw steps data
stepsNA <- is.na(rawSteps)

# Now replace these values with their corresponding mean
rawSteps[stepsNA] <- meanColumn[stepsNA]

# Create new data 
NewData <- data
NewData$steps <- rawSteps
NewData

# First split up the data frame for steps by day
SplitNewSteps <- split(NewData$steps, data$interval)

# Next find the total number of steps over each day
# There should not be an NA values and so we don't need
# to set the flag
totalStepsPerDayNew <- sapply(SplitNewSteps, sum)
```

7. Histogram of the total number of steps taken each day after missing values are imputed

```{r}
par(mfrow=c(1,2))

# plot the original histogram
hist(DaySteps, main="Hist Total Steps per Day", xlab="# Steps", col="bisque3", ylim=c(0,35)) 

# plot the histogram  
abline(v=mean(DaySteps), lty=3, col="blue")                      
abline(v=median(DaySteps), lty=4, col="red")                     
text(mean(DaySteps), 25,labels="mean", pos=4, col="blue")           
text(mean(DaySteps), 23,labels="median", pos=4, col="red")       
rug(DaySteps, col="chocolate")

splitByDayNew <- split(NewData,NewData$date, drop=TRUE)                 
DayStepsNew <- sapply(splitByDayNew, function(x) sum(x$steps)) 

#plot the imputed histogram
hist(DayStepsNew, main="NEW Hist: Total Steps per Day", xlab="# Steps", col="bisque3", ylab="") # plot a histogram  
abline(v=mean(DaySteps), lty=3, col="blue")                      
abline(v=median(DaySteps), lty=4, col="red")                     
text(mean(DaySteps), 35,labels="mean", pos=4, col="blue")         
text(mean(DaySteps), 33,labels="median", pos=4, col="red")         
rug(DayStepsNew,col="chocolate")
```

8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends  

```{r}
# Create weekdays/weekends record
NewData$date <- as.Date(strptime(NewData$date, format="%Y-%m-%d")) 
NewData$day <- weekdays(NewData$date)                              
for (i in 1:nrow(NewData)) {                                       
    if (NewData[i,]$day %in% c("Saturday","Sunday")) {            
        NewData[i,]$day<-"weekend"                                
    }
    else{
        NewData[i,]$day<-"weekday"                                 
    }
}

## aggregate NewData by steps as a function of interval + day  
stepsByDay <- aggregate(NewData$steps ~ NewData$interval + NewData$day, NewData, mean)

## reset the column names to be pretty & clean
names(stepsByDay) <- c("interval", "day", "steps")

## plot weekday over weekend time series
par(mfrow=c(1,1))  
with(stepsByDay, plot(steps ~ interval, type="n", main="Weekday vs. Weekend Avg."))  
with(stepsByDay[stepsByDay$day == "weekday", ], lines(steps ~ interval, type="l", col="chocolate"))  
with(stepsByDay[stepsByDay$day == "weekend", ], lines(steps ~ interval, type="l", col="16" ))  
legend("topright", lty=c(1,1), col = c("chocolate", "16"), legend = c("weekday", "weekend"), seg.len=3)
```






